\subsection{Introduction}
\label{intro}

The ga library is intended to make Geometric Algebra (GA) more accessible by providing
functionality for numerical calculations required for simulation of physical systems. \\

It implements functionality for two- and three-dimensional spaces with Euclidean geometry.
Currently it provides data types and operations to handle the algebras of regular
Euclidean geometry $(ega2d, ega3d)$ and projective Euclidean geometry $(pga2dp, pga3dp)$.
The library handles fully populated multivectors, even and odd grade multivectors as well
as their components like scalars, vectors, bivectors, trivectors, and pseudoscalars of the
corresponding linear spaces efficiently. It also provides operations to work with these
objects. It was designed with extensibiliy in mind and could be easily extended to provide
functionality for handling additional algebras, e.g. spacetime algebra $(STA)$, with
limited effort. \\

The main library is a header only library in the ga sub-folder of the library which can be
used by including either
\begin{verbatim}
    #include ga/ga_ega.hpp // or
    #include ga/ga_pga.hpp
\end{verbatim}
and by making the corresponding namespaces accessible by stating
\begin{verbatim}
    using namespace hd::ga;      // and either
    using namespace hd::ga::ega; // or
    using namespace hd::ga::pga;
\end{verbatim}
in your application code. For a simple usage example please have a look either at
\path{ga_test/src/ga_ega_test.cpp} or at \path{ga_test/src/ga_pga_test.cpp}.
\\

All product expressions for inner, outer, interior and geometric products and their
regressive counterparts are generated via the file \path{ga_prdxpr/ga_prdxpr_main.cpp} and
user input provided in the header files of the corresponding algebra in that folder. The
generated coefficient expressions are used in the source code to implement dot products,
wedge products, contractions, expansions, geometric products or their regressive
counterparts, as well as sandwich products describing rotors and motors of the
corresponding algebras. If certain products are required for your applicaton, but not yet
implemented in the ga library they can easily be generated within
\path{ga_prdxpr/src_prdxpr} as needed and added with minimal effort. \\

As mentioned before, we distinguish inner, interior, outer and geometric products.
\begin{itemize}
    \item Inner and interior products are grade reducing operations.
    \item The inner product combines arguments of the same type and grade, while the
    interior products involve duals of one of their arguments, combining and linking
    arguments of different grades. Inner products are linked to the metric of the modelled
    space. 
    \item The outer product is a grade-increasing operation adding the grades of its
    arguments. Outer products, like the wedge product, connect subspaces represented by
    their operands and create new subspaces defined by the span of the subspaces of the
    operands. They establish a join operation between the involved subspaces.
    \item Geometric products are the combination of both types of products in one
    expression. They can result in multivectors consisting of parts with different grades
    as a result of the operation. Their main advantage is invertibility under certain
    circumstances.
    \item Regressive products use complement operations on their arguments before and
    after applying the operation linking the arguments of the product. Regressive products
    create an intersection of the spaces used by their arguments, thus effectively
    creating a meet operation of between the involved spaces.
    \item Sandwich products are used to implement orthogonal transformations of geometric
    objects. They realize general motions including rotation and translation by
    concatenating two consecutive reflections, which are used as primitives. This can be
    achieved by multiplying the object symmetrically from the left and right hand side
    with the appropriate objects, we either call rotors in $ega$ or motors in $pga$.
\end{itemize}

The complement operation used in this library is defined
in~\cite{Lengyel_pga-illuminated:2024}. It is uniquely determined with respect to the
outer product (not with respect to the geometric product, as commonly used for dualization
by many autors working with GA). This helps to generate consistent signs of expressions,
operators, complements and duals. The left complement $\ubar{u}$ and the right complement
$\bar{u}$ are defined as
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \ubar{u} \wedge u & = & I_n \slabel{eq:lcmpl_def} \\
    u \wedge \bar{u}  & = & I_n \slabel{eq:rcmpl_def}
\end{subeqnarray}
with $I_n  = \bv{1} \wedge \bv{2} \wedge \ldots \wedge \bv{n}$ as the pseudoscalar of the
$n$-dimensional algebra, and the wedge symbol $(\wedge)$ as the operator symbol of the
outer product. The effect of the complement is to turn basis elements not contained in the
object $u$ into the constituting basis elements of the complement of the object $\ubar{u}$
or $\bar{u}$. Connecting the object and its complement with the wedge products creates an
object that fills the available space in the sense of requiring all basis elements of the
space to represent it. The full space itself is represented by the pseudoscalar of that
space, since the pseudoscalar $I_n$ is formed formed by the product of all $n$ basis
vectors of the space. An implication of the definition is that the complement ($\ubar{u}$
or $\bar{u}$) is orthogonal to $u$. \\

It can be shown that the left and right complements fulfill
\begin{equation}
    \bar{\ubar{a}} = a
    \label{eq:double_cmpl}
\end{equation}
thus the left and right complement operations are inverses to each other. This is valid
independent of the sequence of application, independent of the grade of the argument $a$,
and independent of the dimension of the space modelled in the algebra. Based on this it is
possible to define so-called regressive products using rules analog to deMorgan's rules in
boolean algebra as follows (definitions can be found
in~\cite{Lengyel_pga-illuminated:2024} and in \cite{Browne_Grassmann-Algebra_Vol1:2012}):
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    a \vee b & \equiv & \underline{\bar{a} \wedge \bar{b}} \slabel{eq:rwdg_std} \\
    a \vee b & \equiv & \overline{\ubar{a} \wedge \ubar{b}} \slabel{eq:rwdg_alt}
\end{subeqnarray}
The same complement is applied to both arguments first, the resulting expressions are
combined with the wedge product, and the inverse complement operation is used to transform
the outcome into the final result, the regressive product. The inverted wedge symbol
$(\vee)$ is used for the regressive wedge product. Both definitions~(\ref{eq:rwdg_std})
and~(\ref{eq:rwdg_alt}) are equivalent. Equation~(\ref{eq:rwdg_std}) is used to derive
expressions for regressive products in \path{ga_prdxpr}. \\

There are some additional formulas for complements which are useful for working with GA
expressions and thus are provided here for reference. Taking the right and left
complements of~(\ref{eq:rwdg_std}) and~(\ref{eq:rwdg_alt}) we end up with
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \overline{a \vee b}  & = & \overline{\underline{\bar{a} \wedge \bar{b}}}
    \quad  =  \quad \bar{a} \wedge \bar{b}
    \slabel{eq:intermediate_std} \\
    \underline{a \vee b} & = & \underline{\overline{\ubar{a} \wedge \ubar{b}}}
    \quad  =  \quad  \ubar{a} \wedge \ubar{b}
    \slabel{eq:intermediate_alt}
\end{subeqnarray}
Substituting $a = \ubar{u}$ and $b = \ubar{v}$ into~(\ref{eq:intermediate_std}) and $a =
\bar{u}$ and $b = \bar{v}$ into~(\ref{eq:intermediate_alt}) and exchanging the left and
right hand sides of the equations we end up with
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    u \wedge v & = & \overline{\ubar{u} \vee \ubar{v}} \slabel{eq:wdg_std} \\
    u \wedge v & = & \underline{\bar{u} \vee \bar{v}} \slabel{eq:wdg_alt}
\end{subeqnarray}
which shows that the outer and the regressive outer product can be expressed by each
other. \\

The complement operation is used to define a unique dualization operation that works for
cases with non-degenerate as well as for cases with degenerate metric, like in projective
geometric algebra $PGA$. This is the main advantage compared to the usual prodedure of
dualization by multiplying the object with the pseudoscalar (or its inverse) using the
geometric product, since the latter breaks down for spaces with degenerate metrics. \\

The right dual $A^{\star}$ is defined as $A^{\star} = \overline{GA}$, where $A$ is an
arbitrary multivector, $G$ the extended metric and $GA$ a matrix-vector-product between
them. $A^{\star}$ is also called the Hodge dual. There is also a corresponding left
dualization operation defined as $A_{\star} = \underline{GA}$. $\star$ is used as
dualization operator (Hodge star) in both cases. The difference between the hodge dual and
the complement is taking the metric into account before taking the corresponding
complement. However, this does not lead to different results in spaces where the metric is
the identity, but is relevant for spaces with a more involved metric, e.g. in projective
spaces. \\

In a two-dimensional algebra, like $ega2d$ the effect of the right dual on a vector $u$ is
to turn the vector by $+90°$, i.e. into the orientation of $\bv{12}$, while the left dual
turns the vector by $-90°$, i.e. into the direction against the orientientation of
$\bv{12}$. This statement is valid for both right- and left-handed coordinate systems. \\

Relations to the geometric product are given by
\begin{subeqnarray}
    A^{\star} & = &  A^{\dagger} \wedgedot I_n \slabel{eq:hodge_right} \\
    A_{\star} & = & I_n \wedgedot A^{\dagger} \slabel{eq:hodge_left}
\end{subeqnarray}
with $I_n = \bv{1} \wedge \bv{2} \wedge \ldots \wedge \bv{n}$ as the pseudoscalar of the
$n$-dimensional space and where $\dagger$ is the reversion operation and $\wedgedot$ is
the symbol for the geometric product. Following identities hold for operands of same
grade. They link the outer product and the dot product in that case: 
\begin{subeqnarray}
    A \wedge B^{\star} & = & (A \cdot B) I_n, \quad \text{when \path{gr(A)} =
    \path{gr(B)}} \\
    A_{\star} \wedge B & = & (A \cdot B) I_n, \quad \text{when \path{gr(A)} =
    \path{gr(B)}}
\end{subeqnarray}
The influence of the
metric for the dualization operation can be seen in its effect for the different algebras
which also depends on the dimension of the modelled space:
\begin{itemize}
\item $ega2d: A^{\star} = \text{\path{right_dual(A)}}$, $A_{\star} =
\text{\path{left_dual(A)}}$
\item $ega3d: A^{\star} = \text{\path{dual(A)}}$, $A_{\star} = \text{\path{dual(A)}}$
\item $pga2dp: A^{\star} = \text{\path{bulk_dual(A)}}$, $A_{\star} =
\text{\path{bulk_dual(A)}}$
\item $pga3dp: A^{\star} = \text{\path{right_bulk_dual(A)}}$, $A_{\star} =
\text{\path{left_bulk_dual(A)}}$
\end{itemize}
In even dimensional spaces ($ega2d$, $pga3dp$) we have to distinguish left and right
operations, while in odd dimensional spaces ($ega3d$, $pga2dp$) we don't, since they
produce the same result. For projective algebras, like $pga$, we use only the bulk part
for the dualization operation with the Hodge star (a consequence of the extended
degenerate metric, which results in some terms, i.e. the weight terms, being multiplied by
zero). The "left" operations are typically used on the left-hand side of the operation and
the "right" operations on the right-hand side of the product operator. The only deviation
from that pattern occurs for the definition of the regressive products, as defined below.
\\

Following bi-argument operations $\emph{op}(a,b) = a \; op \; b$ and their regressive
variants $\emph{rop}(a,b) = \underline{\bar{a} \; \emph{op} \; \bar{b}}$ are implemented
(with \path{op(a,b)} or \path{rop(a,b)} as the name of the implemented function). Using
$A$ vs. $a$ implies $gr(A) \ge gr(a)$, where $gr(arg)$ returns the grade of the argument
$arg$:
\begin{itemize}
    \item inner product, delivering a scalar: $a \cdot b$, \path{dot(a,b)}
    \item outer product of a $j-$ and $k-$vector, delivering a $j+k$-vector in
    $span(a,b)$: $a \wedge b$, \path{wdg(a,b)}
    \item geometric product:  $a \wedgedot b$, \path{operator*(a,b)}
    \begin{itemize}
        \item For vectors $a$ and $b$:  $a \wedgedot b = a \cdot b + a \wedge b$
        \item For a vector $a$ and a multivector $B$:
        $a \wedgedot B = B \gg a + a \wedge B$
                \item For a multivector $A$ and a vector $b$:
        $A \wedgedot b = b \ll A + A \wedge b$
    \end{itemize}

    \item commutator product (the asymmetric part of the geometric product): $cmt(a,b) =
    \frac{1}{2}(a \wedgedot b - b \wedgedot a)$,
    \path{cmt(a,b)}
    \item regressive inner product, delivering a pseudoscalar: $a \circ b$,
    \path{rdot(a,b)}
    \item regressive outer product: $a \vee b$, \path{rwdg(a,b)}
    \item regressive geometric product: $a \veedot b$, \path{rgpr(a,b)}
    
    \item left (bulk) contraction: $a \lcontr B = a \ll B = a_{\star} \vee B $,
    \path{operator<<(a,B)}, which implements \path{left_bulk_contract(a,B)}  
    \item right (bulk) contraction: $B \rcontr a = B \gg a = B \vee a^{\star} $,
    \path{operator>>(B,a)}, which implements \path{right_bulk_contract(B,a)}

    \item left weight contraction: $a_{\smallstar} \vee B $,
    \path{left_weight_contract(a,B)} 
    \item right weight contraction: $B \vee a^{\smallstar} $,
    \path{right_weight_contract(B,a)} 

    \item left bulk expansion: $A_{\star} \wedge b $,
    \path{left_bulk_expand(A,b)} 
    \item right bulk expansion: $a \wedge B^{\star} $,
    \path{right_bulk_expand(a,B)} 

    \item left weight expansion: $A_{\smallstar} \wedge b $,
    \path{left_weight_expand(A,b)} 
    \item right bulk expansion: $a \wedge B^{\smallstar} $,
    \path{right_weight_expand(a,B)} 
\end{itemize}
where the items from the left contraction downwards are all considered interior products.
They operate on subspaces after applying a dualization operation to one operand. \\

\cite{Lengyel_transwedge_product:2025} has shown that the geometric product can be derived
by using operations from Grassmann algebra exclusively. These are the wedge product, the
complement and the duality operations. This demonstrates that the geometric product is not
a fundamental operation in its own right, but can be derived from even more fundamental
basic operations.

For that purpose Lengyel introduced the transwedge product of order $k$ by following
definition:
\begin{equation}
    a \underset{k}{\doublewedge} b = \sum_{c \, \in \, \mathcal{B}_k}
    (\ubar{c}\vee a) \wedge (b \vee \rightbulkdual{c})
    \label{eq:twdg_orig}
\end{equation}
where $\mathcal{B}_k$ is the set of all basis elements of order $k$. For $ega3d$ it means
for example $\mathcal{B}_0 = \{1\}$, $\mathcal{B}_1 = \{\bv{1}, \bv{2}, \bv{3}\}$,
$\mathcal{B}_2 = \{\bv{23}, \bv{31}, \bv{12}\}$, and $\mathcal{B}_3 = \{\bv{123}\}$. From
these expressions the geometric product can be formed by
\begin{equation}
    a \wedgedot b = \sum_{k \, = \, 0}^{n} (-1)^{k(k-1)/2} a \underset{k}{\doublewedge} b
    \label{eq:grp_twdg}
\end{equation}

For $k=0$ the formula yields the outer product (wedge product). For higher values of $k$
other products are generated which sum up to the full geometric product. The transwedge
products helps to decompose the geometric product into a sum of terms. The most
interesting terms for applications beyond the wedge product resulting for $k=0$ is the
transwedge product resulting for the order $k=1$. \\

The product tables for all products are generated in \path{ga_prdxpr}, and coefficient
expressions can be generated based on those product tables. For that purpose the
regressive wedge product can be replaced by the wedge product using the complement-based
definitions given in equations~(\ref{eq:double_cmpl}) to~(\ref{eq:wdg_alt}). That means
for example that equation~(\ref{eq:twdg_orig}) can be written equivalently as
\begin{equation}
    a \underset{k}{\doublewedge} b = \sum_{c \, \in \, \mathcal{B}_k}
    (\underline{c \wedge \bar{a}}) \wedge
    (\underline{\bar{b} \wedge \overline{\rightbulkdual{c}}})
    \label{eq:twdg_wdg}
\end{equation}
expressing it by only using the wedge product. \\

Since every product can by assigned a corresponding regressive product in analogy to
definition~(\ref{eq:rwdg_std}) or~(\ref{eq:rwdg_alt}), the transwedge
product~(\ref{eq:twdg_orig}) also has a regressive form:
\begin{equation}
    a \underset{k}{\doublevee} b \equiv
    \overline{\ubar{a} \underset{k}{\doublewedge} \ubar{b}} =
    \sum_{c \, \in \, \mathcal{B}_k} \overline{
    (\ubar{c}\vee \ubar{a}) \wedge (\ubar{b} \vee \rightbulkdual{c})}
    \label{eq:rtwdg_orig}
\end{equation}
which in turn can be expressed in terms of the wedge product exclusively as
\begin{equation}
    a \underset{k}{\doublevee} b = \sum_{c \, \in \, \mathcal{B}_k}  \overline{
    (\underline{c\vee a}) \wedge (\underline{b \vee \overline{\rightbulkdual{c}}})}
    \label{eq:rtwdg_wdg}
\end{equation}
for calculation of the product tables in \path{ga_prdxpr}. \\

% Missing products needed for specific applications can be implemented in the ga library
% easily, if required with minor effort. \\



TODO: \emph{fill in further introductory notes here} \\

\newpage
