\subsection{Basics of the mathematical approach}
\label{subsection:basics}

\subsubsection{Basis vectors for each modelled algebra}
\label{subsection:basis_vectors_each_algebra}

The following is provided for Euclidean geometric algebra of two-dimensional space
($EGA2D$) using an orthonormal basis \bv{1}, \bv{2}:
\begin{subequations}
    \begin{align}
        (\bv{1})^2 = \bv{1} \wedgedot \bv1 & = \bv{1} \cdot \bv1 = \bv{1}^2 = +1,
        \;\text{and}\; \bv{2}^2 = +1 \nonumber\\
        s_{2d} & = s \, \bm{1}
        \label{eq:scalar2d} \\
        v_{2d} & = v_1\bv{1} + v_2\bv{2} 
        \label{eq:vec2d} \\
        ps_{2d} & = ps \, \bv{12}  = ps \, \ps
        \label{eq:pscalar2d}
    \end{align}
\end{subequations}
equation (\ref{eq:scalar2d}) is the scalar part $s_{2d}$, equation (\ref{eq:vec2d})
contains the vector part $v_{2d}$, and equation (\ref{eq:pscalar2d}) contains the
pseudoscalar part $ps_{2d}$. The index $2d$ is omitted when clear from context.
The four basis elements are $\{\bm{1}, \bv{1}, \bv{2}, \bv{12}\}$. Using this basis a
multivector $M$ of $2d$-space is defined as
\begin{equation}
     M = s \, \bm{1} + v_1\bv{1} + v_2\bv{2} + ps \, \bv{12}
    \label{eq:mvec2d}   
\end{equation}
where equation (\ref{eq:mvec2d}) contains three parts: the scalar part $s \, \bm{1}$
(basis element is the scalar \textbf{1}; if \textbf{1} is not shown, it is implicitly
assumed for scalar values), the vector part $v$ (basis elements \bv{1} and \bv{2}) and the
pseudoscalar part $ps \, \bv{12} = ps \, \ps$ (basis element is $I_{2d} =
\bv{12}$, which is sometimes written as $\ps$ to show its character as
pseudoscalar of this space. $I_{2d}$ is a bivector in two-dimensional Euclidean space). \\

For Euclidean geometric algebra of three-dimensional space ($EGA3D$) using an
orthonormal basis \bv{1}, \bv{2}, \bv{3} there are:
\begin{subequations}
    \begin{align}
        (\bv{1})^2 = \bv{1} \wedgedot \bv1 & = \bv{1} \cdot \bv1 = \bv{1}^2 = +1,
        \bv{2}^2 = +1, \;\text{and}\; \bv{3}^2 = +1 \nonumber\\ 
        s_{3d} & = s \, \bm{1}
        \label{eq:scalar3d} \\
        v_{3d} & = v_1\bv{1} + v_2\bv{2} + v_3\bv{3} 
        \label{eq:vec3d} \\ 
        B_{3d} & = b_{1}\bv{23} + b_{2}\bv{31} + b_{3}\bv{12} 
        \label{eq:bivec3d} \\ 
        ps_{3d} & = ps \, \bv{123}  = ps \, \ps
        \label{eq:pscalar3d}
    \end{align}
\end{subequations}
equation (\ref{eq:scalar3d}) is the scalar part $s_{3d}$, equation (\ref{eq:vec3d}) is the
vector part $v_{3d}$, equation (\ref{eq:bivec3d}) is the bivector part $B_{3d}$, and
equation (\ref{eq:pscalar3d}) is the pseudoscalar part $ps_{3d}$. The index $3d$ is
omitted when clear from context. Compared to the $2d$-case it is obvious, that all parts
depend on context, specifically on the dimensionality of the modelled space, and thus need
to be defined and treated accordingly.\footnote{\emph{hint}: Since \Cpp is a statically
typed language, those types need to be well-defined and distinguishable from each other.
This is especially important for the different scalar and pseudoscalar types. The ga
library uses type tagging extensively for that purpose.} The basis elements are $\{\bm{1},
\bv{1}, \bv{2}, \bv{3}, \bv{23}, \bv{31}, \bv{12}, \bv{123}\}$. Using this basis a
multivector $M$ of $3d$-space is defined as
\begin{equation}
    M = s \, \bm{1} + v_1\bv{1} + v_2\bv{2} + v_3\bv{3} 
    + b_{1}\bv{23} + b_{2}\bv{31} + b_{3}\bv{12} + ps \, \bv{123}
    \label{eq:mvec3d}  
\end{equation}
where equation (\ref{eq:mvec3d}) contains four parts: the scalar component $s \, \bm{1}$
(basis element is the scalar \textbf{1}), the vector part $v$ (basis elements \bv{1},
\bv{2} and \bv{3}), the bivector part $B$ (basis elements \bv{23}, \bv{31} and \bv{12})
and the pseudoscalar part $ps \, \bv{123} = ps \, \ps$ (basis element is $I_{3d}
= \bv{123}$, which is sometimes written as $\ps$ to show its character as
pseudoscalar of this space. $I_{3d}$ is a trivector in three-dimensional Euclidean space).
\\

Projective geometric algebra of $2D$ Euclidean space ($PGA2DP$) is modelled by embedding
it in a three-dimensional space. Using an orthonormal basis \bv{1}, \bv{2}, \bv{3} there
are following basis vectors:
\begin{subequations}
    \begin{align}
        (\bv{1})^2 = \bv{1} \wedgedot \bv1 & = \bv{1} \cdot \bv1 = \bv{1}^2 = +1,
        \bv{2}^2 = +1, \;\text{and}\; \bv{3}^2 = 0  \nonumber\\ 
        s_{2dp} & = s \, \bm{1}
        \label{eq:scalar2dp} \\
        v_{2dp} & = v_1\bv{1} + v_2\bv{2} + v_3\bv{3} 
        \label{eq:vec2dp} \\ 
        B_{2dp} & = b_{1}\bv{31} + b_{2}\bv{32} + b_{3}\bv{12} 
        \label{eq:bivec2dp} \\ 
        ps_{2dp} & = ps \, \bv{321}  = ps \, \ps
        \label{eq:pscalar2dp}
    \end{align}
\end{subequations}
The most obvious difference is the basis vector~\bv{3} that squares to zero and the
pseudoscalar of that space that is chosen to be \bv{321}. The trivector \bv{321} squares
to $0$, because of it's component \bv{3} contained as a factor. Equation
(\ref{eq:scalar2dp}) is the scalar part $s_{2dp}$, equation (\ref{eq:vec2dp}) is the
vector part $v_{2dp}$, equation (\ref{eq:bivec2dp}) is the bivector part
$B_{2dp}$\footnote{Here we deviate from Lengyel's approach chosen in
\cite{Lengyel_pga-illuminated:2024}. Since we want to model physics, it is important to
generate unified handling for two- and three-dimensional cases.  The chosen sequence of
basis bivectors assures that lines, which are used to model forces and moments later on,
have their directions encoded in a unified way.}, and equation (\ref{eq:pscalar2dp}) is
the pseudoscalar part $ps_{2dp}$. The index $2dp$ is omitted when clear from context. The
eight basis elements are $\{\bm{1}, \bv{1}, \bv{2}, \bv{3}, \bv{31}, \bv{32}, \bv{12},
\bv{321}\}$. Using this basis a multivector $M$ of two-dimensional projective space
($2dp$) is defined as
\begin{equation}
    M = s \, \bm{1} + v_1\bv{1} + v_2\bv{2} + v_3\bv{3} 
    + b_1\bv{31} + b_2\bv{32} + b_3\bv{12} + ps \, \bv{321}
    \label{eq:mvec2dp}  
\end{equation}
where equation (\ref{eq:mvec2dp}) contains four parts: the scalar component $s \, \bm{1}$
(basis element is the scalar \textbf{1}), the vector part $v$ (basis elements \bv{1},
\bv{2} and \bv{3}), the bivector part $B$ (basis elements \bv{31}, \bv{32} and \bv{12})
and the pseudoscalar part $ps \, \bv{321} = ps \, \ps$ (basis element is
$I_{2dp} = \bv{321}$, which is sometimes written as $\ps$ to show its character
as pseudoscalar of this space. It is a trivector in two-dimensional Euclidean projective
space that squares to zero). \\

In $2D$ $PGA$ a projective point $Q$ is modelled as a vector
\begin{equation}
    Q_{pga2dp} = \underbrace{q_x \bv{1} + q_y \bv{1}}_{position} +
                 \underbrace{q_z \bv{3}}_{weight}
\end{equation}
where the position is encoded in the bulk part\footnote{Bulk and weight parts are defined
further below in this section.} of the point, and the weight is the projective component
$q_z$ that must be unitized (=scaled to 1.0) in order to project the point into the
Euclidean space. \\

A line $L$ is modelled as a bivector
\begin{equation}
    L_{pga2dp} = \underbrace{l_x \bv{31} + l_y \bv{32}}_{direction} +
                 \underbrace{l_z \bv{12}}_{position}
\end{equation}
where the weight part encodes the two-dimensional direction vector $(l_x,l_y)$ of the line
and the bulk part encodes the position of the line as distance to the origin $O_{2dp} =
(0,0,1)$. When $l_z = 0$ the line passes through the origin. \\

Projective geometric algebra of $3D$ Euclidean space ($PGA3DP$) is modelled by embedding
it in a four-dimensional space. Using an orthonormal basis \bv{1}, \bv{2}, \bv{3}, \bv{4}
there are following basis vectors:
\begin{subequations}
    \begin{align}
        (\bv{1})^2 = \bv{1} \wedgedot \bv1 & = \bv{1} \cdot \bv1 = \bv{1}^2 = +1,
        \bv{2}^2 = +1, \bv{3}^2 = +1,\;\text{and}\; \bv{4}^2 = 0 \nonumber\\ 
        s_{3dp} & = s \, \bm{1}
        \label{eq:scalar3dp} \\
        v_{3dp} & = v_1\bv{1} + v_2\bv{2} + v_3\bv{3} + v_4\bv{4} 
        \label{eq:vec3dp} \\ 
        B_{3dp} & = b_{1}\bv{41} + b_{2}\bv{42} + b_{3}\bv{43} +
                    b_{4}\bv{23} + b_{5}\bv{31} + b_{6}\bv{12} 
        \label{eq:bivec3dp} \\ 
        t_{3dp} & = t_{1}\bv{423} + t_{2}\bv{431} + t_{3}\bv{412} + t_{4}\bv{321} 
        \label{eq:trivec3dp} \\ 
        ps_{3dp} & = ps \, \bv{1234}  = ps \, \ps
        \label{eq:pscalar3dp}
    \end{align}
\end{subequations}
Now we have a basis vector~\bv{4} that squares to zero. The bivector has six components,
and there is a trivector $t$. The pseudoscalar \bv{1234} is a quadvector that squares to
$0$, because of its component \bv{4} contained as a factor. Equation (\ref{eq:scalar3dp})
is the scalar part $s_{3dp}$, equation (\ref{eq:vec3dp}) is the vector part $v_{3dp}$,
equation (\ref{eq:bivec3dp}) is the bivector part $B_{3dp}$, equation (\ref{eq:trivec3dp})
is the trivector part $t_{3dp}$, and equation (\ref{eq:pscalar3dp}) is the pseudoscalar
part $ps_{3dp}$. The index $3dp$ is omitted when clear from context. The sixteen basis
elements of $PGA3DP$ are: $\{\bm{1}, \bv{1}, \bv{2}, \bv{3}, \bv{4}, \bv{41}, \bv{42},
\bv{43}, \bv{23}, \bv{31}, \bv{12}, \bv{423}, \bv{431}, \bv{412}, \bv{321}, \bv{1234}\}$.
A multivector $M$ of three-dimensional projective space ($3dp$) is defined as
\begin{equation}
    \begin{split}
    M & = s \, \bm{1} \\
      & + v_1\bv{1} + v_2\bv{2} + v_3\bv{3} + v_4\bv{4} \\
      & + b_{1}\bv{41} + b_{2}\bv{42} + b_{3}\bv{43} 
        + b_{4}\bv{23} + b_{5}\bv{31} + b_{6}\bv{12} \\
      & + t_{1}\bv{423} + t_{2}\bv{431} + t_{3}\bv{412} + t_{4}\bv{321} \\
      & + ps \, \bv{1234}
    \label{eq:mvec3dp}
    \end{split}
\end{equation}
where equation (\ref{eq:mvec3dp}) contains five parts: the scalar component $s \, \bm{1}$
(basis element is the scalar \textbf{1}), the vector part $v$ (basis elements \bv{1},
\bv{2}, \bv{3}, \bv{4}), the bivector part $B$ (basis elements \bv{41}, \bv{42}, \bv{43},
\bv{23}, \bv{31}, \bv{12}), the trivector part $t$ (basis elements \bv{423}, \bv{431},
\bv{412}, \bv{321}) and the pseudoscalar part $ps \, \bv{1234} = ps \, \ps$
(basis element is $I_{3dp} = \bv{1234}$, which is sometimes written as $\ps$ to
show its character as pseudoscalar of this space. It is a quadvector in three-dimensional
Euclidean projective space that squares to zero). \\

In $3D$ $PGA$ a projective point $Q$ is modelled as a vector
\begin{equation}
    Q_{pga3dp} = \underbrace{q_x \bv{1} + q_y \bv{1} + q_z \bv{1}}_{position} +
                 \underbrace{q_w \bv{4}}_{weight}
\end{equation}
where the position is encoded in the bulk part of the point, and the weight is the
projective component $q_w$ that must be unitized (=scaled to 1.0) in order to project the
point into the Euclidean space. \\

A line $L$ is modelled as a bivector
\begin{equation}
    L_{pga3dp} = \underbrace{l_{vx} \bv{41} + l_{vy} \bv{42} + l_{vz} \bv{43}}_{direction}
                 +
                 \underbrace{l_{mx} \bv{23} + l_{my} \bv{31} + l_{mz} \bv{12}}_{moment}
\end{equation}
where the weight part encodes the three-dimensional direction vector of the line $l_v =
(l_{vx},l_{vy},l_{vz})$, and the bulk part encodes the moment of the line $l_m =
(l_{mx},l_{my},l_{mz})$ when interpreted as a vector. Both are encoded in the bivector
object that captures all parameters needed to model the line. \\

Additionally, there are planes $T$ that are modelled as trivectors
\begin{equation}
    T_{pga3dp} = \underbrace{t_x \bv{423} + t_y \bv{431} + t_z \bv{412}}_{normal} +
                 \underbrace{t_w \bv{312}}_{position}
                 \label{eq:plane_def_pga3dp}
\end{equation}
where the first three components represent the normal vector of the plane $n = (t_x, t_y,
t_z)$ and the fourth component $t_w$ corresponds to the signed distance to the origin
$O_{3dp} = (0,0,0,1)$. \\

Many more examples of how geometric objects are modelled in projective geometric algebra
and which operations can be used to manipulated them can by found in Lengyel's book
\cite{Lengyel_pga-illuminated:2024}. Browne explains very well how Grassmann algebra and
projective geometry and geometric algebra are related in
\cite{Browne_Grassmann-Algebra_Vol1:2012}, p. 216ff. It is helpful to fully grasp and
understand that relation in oder to see what is possible using the algebra effectively to
model physical phenomena. \\

Geometric algebra seems complex and a little overwhelming at first. Especially so, when
new objects beyond vectors and several specific products are introduced.
Expressions tend to quickly become large for fully occupied multivectors. However,
it is important to note that it is rarely required to use full multivectors in the
respective algebra. For most applications it is sufficient to use a blade or $k$-vector
exclusively, which represents only one grade of the multivector, e.g. a vector or a
bivector, which can be represented with very few coefficients. This reduces the effort in
terms of memory and computational requirements significantly -- typically reduced to a
level that is required anyway to solve the underlying geometrical question at hand.


\newpage

\subsubsection{The complement}
\label{subsection:complement}

The complement operation used in the library is defined
in~\cite{Lengyel_pga-illuminated:2024}. It is uniquely determined with respect to the
outer product (\emph{not} with respect to the geometric product, as commonly used for
dualization by many authors working with GA). This ensures consistent signs of
expressions, operators, complements and duals. The left complement $\lcmpl{u}$ and the
right complement $\rcmpl{u}$ are defined as
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \lcmpl{u} \wedge u & = & I_n \slabel{eq:lcmpl_def} \\
    u \wedge \rcmpl{u}  & = & I_n \slabel{eq:rcmpl_def}
    \label{eq:cmpl_def}
\end{subeqnarray}
with $I_n  = \bv{1} \wedge \bv{2} \wedge \ldots \wedge \bv{n}$ as the pseudoscalar of the
$n$-dimensional algebra, and the wedge symbol $(\wedge)$ as the operator symbol of the
outer product. The effect of the complement is to turn basis elements not contained in the
object $u$ into the constituting basis elements of the complement of the object
$\lcmpl{u}$ or $\rcmpl{u}$. Connecting the object and its complement with the wedge
products creates an object that fills the available space in the sense of requiring all
basis elements of the space to represent it. The full space itself is represented by the
pseudoscalar of that space, since the pseudoscalar $I_n$ is formed by the product of all
$n$ basis vectors of the space. An implication of the definition is that the complement
($\lcmpl{u}$ or $\rcmpl{u}$) is orthogonal to $u$, since we exclusively use orthogonal
basis vectors to model our space. \\

The complement is defined as given above in equations~(\ref{eq:cmpl_def}) for any basis
element. It is extended to full multivectors $A$ and $B$ by linearity as
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \rcmpl{sA}  \!  & = & \! s \rcmpl{A}, \quad
    \rcmpl{A + B} \:\: = \:\: \rcmpl{A} + \rcmpl{B}
    \slabel{eq:rcmpl_linearity_def} \\
    \lcmpl{sA}  \!  & = & \! s \lcmpl{A}, \quad
    \lcmpl{A + B} \:\: = \:\: \lcmpl{A} + \lcmpl{B}
    \slabel{eq:lcmpl_linearity_def}
    \label{eq:cmpl_linearity_def}
\end{subeqnarray}
where $s$ is a scalar. The left and right complements fulfill
\begin{equation}
    \rcmpl{\lcmpl{u}} = u
    \label{eq:double_cmpl}
\end{equation}
thus the left and right complement operations are inverses to each other. This is valid
independent of the sequence of application, independent of the grade of the argument $a$,
and independent of the dimension of the space modelled in the algebra. In spaces of odd
dimension, where we only have one complement, applying the complement twice of course
returns the original object as well. \\

However, applying the same complement operation several times in spaces of even dimension
does not ensure that you get back the original object. The result is dependent on the
grade of the basis vector and the dimension $n$ of the space. When applying the same
complement twice we get
\begin{equation}
    \rcmpl{\rcmpl{u}} = \lcmpl{\lcmpl{u}} = (-1)^{gr(u)(n-gr(u))} u
\end{equation}
i.e. this involves grade-dependent sign changes. So it is favorable to use left and right
complements as inverses to each other. \\

Based on the given complement definition it is possible to define regressive products
using rules analog to deMorgan's rules in Boolean algebra as follows (definitions can be
found in~\cite{Lengyel_pga-illuminated:2024} and in
\cite{Browne_Grassmann-Algebra_Vol1:2012}):
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    a \vee b & \equiv & \lcmpl{\rcmpl{a} \wedge \rcmpl{b}} \slabel{eq:rwdg_std} \\
    a \vee b & \equiv & \rcmpl{\lcmpl{a} \wedge \lcmpl{b}} \slabel{eq:rwdg_alt}
\end{subeqnarray}
The same complement is applied to both arguments first, the resulting expressions are
combined with the wedge product, and the inverse complement operation is used to transform
the outcome into the final result, the regressive product. The inverted wedge symbol
$(\vee)$ is used for the regressive wedge product. Both definitions~(\ref{eq:rwdg_std})
and~(\ref{eq:rwdg_alt}) are equivalent. Equation~(\ref{eq:rwdg_std}) is used to derive
expressions for regressive products in \path{ga/ga_prdxpr}. \\

There are some additional formulas for complements which are useful for working with GA
expressions and thus are provided here for reference. Taking the right and left
complements of~(\ref{eq:rwdg_std}) and~(\ref{eq:rwdg_alt}) we end up with
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \rcmpl{a \vee b}  & = & \rcmpl{\lcmpl{\rcmpl{a} \wedge \rcmpl{b}}}
    \quad  =  \quad \rcmpl{a} \wedge \rcmpl{b}
    \slabel{eq:intermediate_std} \\
    \lcmpl{a \vee b} & = & \lcmpl{\rcmpl{\lcmpl{a} \wedge \lcmpl{b}}}
    \quad  =  \quad  \lcmpl{a} \wedge \lcmpl{b}
    \slabel{eq:intermediate_alt}
\end{subeqnarray}
Substituting $a = \lcmpl{u}$ and $b = \lcmpl{v}$ into~(\ref{eq:intermediate_std}) and $a =
\rcmpl{u}$ and $b = \rcmpl{v}$ into~(\ref{eq:intermediate_alt}) and exchanging the left
and right hand sides of the equations we end up with
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    u \wedge v & = & \rcmpl{\lcmpl{u} \vee \lcmpl{v}} \slabel{eq:wdg_std} \\
    u \wedge v & = & \lcmpl{\rcmpl{u} \vee \rcmpl{v}} \slabel{eq:wdg_alt}
\end{subeqnarray}
which shows that the outer and the regressive outer product can be expressed by each
other. \\

The complement operation is used to define a unique dualization operation that works for
cases with non-degenerate as well as for cases with degenerate metric, like in projective
geometric algebra $PGA$. This is the main advantage compared to the usual prodedure of
dualization by multiplying the object with the pseudoscalar (or its inverse) using the
geometric product, since the latter breaks down for spaces with degenerate metrics. For an
overview on all complements in $PGA$ see tables \ref{tab:pga2dp_complement} and
\ref{tab:pga3dp_complement} at the end of this section.


\newpage

\subsubsection{The metric}
\label{subsection:metric}

The metric is defined by the dot product of the basis vectors as
\begin{equation}
    \mathfrak{g}_{ij} \equiv \bv{i} \cdot \bv{j}
\end{equation}
It is a square matrix that is the identity matrix $I_n$ for $EGA$, since all basis vectors
square to $\bv{i} \cdot \bv{i} = +1$, and $\bv{i} \cdot \bv{j} = 0$ for $i \neq j$. It is
invertible in this case. However, for $PGA$ one basis vector squares to zero, and thus
there is one zero on the main diagonal of the corresponding matrix of the metric, which is
therefore not invertible (its determinant is zero). It is called degenerate for that
reason. \\

The extended metric $\metric{G}$ is a so-called exomorphism matrix that fulfills
\begin{equation}
    \metric{G}(A \wedge B) = \metric{G}(A) \wedge \metric{G}(B)
    \label{eq:exomorphism}
\end{equation} for any multivector $A$ and $B$ (see~\cite{Lengyel_pga-illuminated:2024},
pp. 61ff.). \\
Regarding $\metric{G}(u) = \metric{G} u$ as a unary operation on $u$ we can define the
corresponding regressive operation by first taking the complement of the argument and then
applying the opposed complement on the result. This defines the regressive metric
$\rmetric{G}$ (named anti-metric by Lengyel). It is defined as
\begin{equation}
    \rmetric{G} u \equiv \lcmpl{\metric{G} \rcmpl{u}}
    = \rcmpl{\metric{G} \lcmpl{u}}
\end{equation}
$\metric{G}$ and $\rmetric{G}$ are related by 
\begin{equation}
    \metric{G} \rmetric{G} = det(\mathfrak{g}) I_n
\end{equation}
Thus, if all basis vectors square to $\pm 1$, we have $\metric{G} \rmetric{G} =
\pm I_n$, and both are invertible. However, if one basis vector squares to zero, we get
$\metric{G} \rmetric{G} = 0$, which means that none of them is invertible, i.e. we
have a degenerate metric.\\

For $\rmetric{G}$ there is the relation
\begin{equation}
    \rmetric{G}(A \vee B) = \rmetric{G}(A) \vee \rmetric{G}(B)
\end{equation}
that is complementary to equation (\ref{eq:exomorphism}). \\

In projective geometric algebra the product $\metric{G} u$ defines the \emph{bulk} of of
an object $u$. It is defined as
\begin{equation}
    \bulk{u} = \metric{G} u
\end{equation}
The bulk only contains components that \emph{do not} contain the projective dimension
(\bv{3} for $2D$ $PGA$ or \bv{4} for $3D$ $PGA$) as a factor. The \emph{bulk $\bulk{u}$
contains positional information} on $u$ relative to the origin. Zero bulk ($\bulk{u} = 0$)
means that $u$ contains the origin.\\

The product $\rmetric{G} u$ defines the \emph{weight} of an object $u$. It is defined as
\begin{equation}
    \weight{u} = \rmetric{G} u
\end{equation}
The weight selects components that \emph{do} indeed contain the projective dimension
(\bv{3} for $2D$ $PGA$ or \bv{4} for $3D$ $PGA$) as a factor. The \emph{weight
$\weight{u}$ contains directional information} on $u$, i.e. its attitude and direction
independent of its position. Zero weight ($\weight{u} = 0$) means that $u$ is contained in
the horizon, i.e. it is infinitely far away.\\

This effective split for each object into bulk and weight in projective algebras can be
written as
\begin{equation}
    u = \bulk{u} + \weight{u}
\end{equation}

This split is also relevant for the corresponding norms
\begin{equation}
    \nrm{u} = \bulknrm{u} + \weightnrm{u} = \sqrt{A \cdot A} + \sqrt{A \circ A}
\end{equation}
where the left hand side is called the \emph{geometric norm} and the right-hand side
consists of the sums of the \emph{bulk} and \emph{weight norms}. The \emph{bulk norm}
returns a scalar value of the corresponding algebra, whereas the \emph{weight norm}
provides a pseudoscalar value of the respective algebra. The dot product (symbol
'$\cdot$') and its regressive counterpart (symbol '$\circ$') are defined in
tables~\ref{tab:pga2dp_dot_products} and \ref{tab:pga3dp_dot_products} in subsection
\ref{subsection:tables_of_products}. \\

Consequently, the geometric norm is an object consisting of two distinct parts, a scalar
and a pseudoscalar of the respective algebra, defined here as a \emph{dual number} in the
corresponding algebra. When we apply \emph{unitization} to geometric objects we devide the
object in such a way that the magnitude of the weight norm becomes a value of one. Both,
bulk and weight norms, have a positive argument under the square root due to the
definitions of the inner product and the regressive inner product.


\newpage

\subsubsection{The dual}
\label{subsection:dual}

In $EGA$ the dual operation (marked by a filled star \rightbulkdual{}) is identical with
the complement (reason: the Euclidean metric is the identity matrix; this changes for
$PGA$ as shown below). Therefore, in $2D$ $EGA$ for the left and right dual operations are
defined as
\begin{eqnarray}
    \rightbulkdual{u} &=& \rcmpl{u} \qquad \textit{(right dual)}\\
    \leftbulkdual{u}  &=& \lcmpl{u}  \qquad \textit{(left dual)}
\end{eqnarray}
while in $3D$ $EGA$, an odd-dimensional algebra, where no distinction between left and
right operations exists, we get
\begin{equation}
    \rightbulkdual{u} = \rcmpl{u} \qquad \textit{(dual)}
\end{equation}
where the overbar in $\rcmpl{u}$ is the unique complement operation. \\

Now for $PGA$: Using the metric and the right complement, the \emph{(right) bulk dual}
\rightbulkdual{A} is defined as
\begin{equation}
    \rightbulkdual{A} = \rcmpl{\metric{G}A} = \rcmpl{\bulk{A}}
    \label{eq:rightbulkdual}
\end{equation}
where $A$ is an arbitrary multivector, $\metric{G}$ the extended metric and $\metric{G}A$
a matrix-vector-product between them. \rightbulkdual{A} is also called the Hodge dual. The
multiplication with the metric happens before the application of the complement
operation.\\

There is also a corresponding \emph{(left) bulk dual} $\leftbulkdual{A}$ operation using
the metric and the left complement. It is defined as
\begin{equation}
    \leftbulkdual{A} = \lcmpl{\metric{G}A} = \lcmpl{\bulk{A}}
    \label{eq:leftbulkdual}
\end{equation}
Where the left and right operations only differ for $3D$ $PGA$, while for $2D$ $PGA$, a
three-dimensional and thus an odd-dimensional algebra, there is no difference between the
left and right operations. \\

The Hodge star (symbolized by a filled star $\star$) is used as dualization operator in
both cases. The lower or upper position symbolizes the left and right application. The
difference between the Hodge dual and the complement is in both cases that multiplying the
multivector with the extended metric happens \emph{before} taking the corresponding
complement. However, this does not lead to different results in spaces where the metric is
the identity matrix. It only yields different results in spaces with a more involved
metric, e.g. in projective spaces with a non-invertible (degenerate) metric as in $PGA$.
\\

The effect of applying the metric before taking the complement is equivalent to applying
the complement operation on the bulk part of the object $u$ exclusively, and not on the
full object $u$. Of course, the dualization can be applied to the weight part as well.
This operation (symbolized by an unfilled star~``$\smallstar$'') uses the regressive
extended metric and is defined for the right weight dual \rightweightdual{A} using the
right complement as
\begin{equation}
    \rightweightdual{A} = \rcmpl{\rmetric{G}A} = \rcmpl{\weight{A}}
        \label{eq:rightweightdual}
\end{equation}
for an arbitrary multivector $A$. The left weight dual \leftweightdual{A} is defined as
\begin{equation}
    \leftweightdual{A} = \lcmpl{\rmetric{G}A} = \lcmpl{\weight{A}}
        \label{eq:leftweightdual}
\end{equation}
using the left complement on the weight part of $A$.\\

Relations to the (regressive) geometric product are given by
\begin{subeqnarray}
    \rightbulkdual{A} & = &  \rev{A} \wedgedot I_n = \rev{A} \wedgedot
    \ps \slabel{eq:hodge_right} \\
    \leftbulkdual{A} & = & I_n \wedgedot \rev{A} = \ps \wedgedot
    \rev{A} \slabel{eq:hodge_left} \\
    \rightweightdual{A} & = &  \rrev{A} \veedot \bm{1} \slabel{eq:weight_right} \\
    \leftweightdual{A} & = & \bm{1} \veedot \rrev{A} \slabel{eq:weight_left}
\end{subeqnarray}
with $I_n = \bv{1} \wedge \bv{2} \wedge \ldots \wedge \bv{n} = \ps$ as the pseudoscalar of
the $n$-dimensional space. $\rev{A}$ is the reversion operation, and $\undertilde{A}$ is
the regressive reversion operation applied to $A$ (e.g. with $A = \bv{1} \wedge \bv{2}$
the reverse is $\rev{A} = \bv{2} \wedge \bv{1}$), and ``$\wedgedot$'' and ``$\veedot$''
are the symbols for the geometric and regressive geometric products. As the complement,
the reverse is defined for any basis element. It is extended to full multivectors $A$ and
$B$ by linearity as
\begin{subeqnarray}
    % use \label for full label
    % use \slabel for sublabel
    \rev{sA}  \!  & = & \! s \rev{A}, \quad
    \rev{A + B} \:\: = \:\: \rev{A} + \rev{B}, \quad
    \rev{A \wedgedot B} \:\: = \:\: \rev{B} \wedgedot \rev{A}
    \slabel{eq:rev_linearity_def} \\
    \rrev{sA}  \!  & = & \! s \rrev{A}, \quad
    \rrev{A + B} \:\: = \:\: \rrev{A} + \rrev{B}, \quad
    \rrev{A \veedot B} \:\: = \:\: \rrev{B} \veedot \rrev{A}
    \slabel{eq:rrev_linearity_def} \\
    \rev{\rev{A}} \!  & = & \! A, \quad \text{and} \quad \rrev{\rrev{A}}  =  A
    \label{eq:reverse_linearity_def}
\end{subeqnarray}
where $s$ is a scalar. The reverse and regressive reverse are their own inverses, i.e.
applying the same reverse twice returns the original object. \\

For an overview on all bulk and weight duals in $PGA$ see tables
\ref{tab:pga2dp_bulk_weight_duals} and \ref{tab:pga3dp_bulk_weight_duals} at the end of
this section. The reverses can be found in tables~\ref{tab:pga2dp_rev_rrev} and
\ref{tab:pga3dp_rev_rrev} at the end of the section as well.\\

Following identities link the expansions and contractions and thus the outer and inner
products: 
\begin{subeqnarray}
    \rightbulkdual{(A \wedge B)} & = & \rightbulkdual{A} \vee \rightbulkdual{B} \\
    A \wedge \rightbulkdual{B} & = & (A \cdot B) \wedge \ps,
    \quad \text{when \path{gr(A)} = \path{gr(B)}} \\
    A \wedge \rightweightdual{B} & = & A \circ B,
    \quad \text{when \path{gr(A)} = \path{gr(B)}} \\[10pt]
    \rightweightdual{(A \vee B)} & = & \rightweightdual{A} \wedge
    \rightweightdual{B} \\
    A \vee \rightbulkdual{B} & = & A \cdot B,
    \quad \text{when \path{gr(A)} = \path{gr(B)}} \\
    A \vee \rightweightdual{B} & = & (A \circ B) \vee \bm{1},
    \quad \text{when \path{gr(A)} = \path{gr(B)}}
\end{subeqnarray}

The influence of the metric for the dualization operation can be seen in its effect for
the different algebras which also depends on the dimension of the modelled space:
\begin{itemize}
\item $EGA2D$: $\;\:A_{\star} = \text{\path{left_dual(A)}}$, $A^{\star} =
\text{\path{right_dual(A)}}$
\item $EGA3D$: $\;\:A_{\star} = \text{\path{dual(A)}}$, $A^{\star} =
\text{\path{dual(A)}}$
\item $PGA2DP$: $A_{\star} = \text{\path{bulk_dual(A)}}$, $A^{\star} =
\text{\path{bulk_dual(A)}}$ 
\item $PGA2DP$: $A_{\smallstar} = \text{\path{weight_dual(A)}}$, $A^{\smallstar} =
\text{\path{weight_dual(A)}}$
\item $PGA3DP$: $A_{\star} = \text{\path{left_bulk_dual(A)}}$, $A^{\star} =
\text{\path{right_bulk_dual(A)}}$
\item $PGA3DP$: $A_{\smallstar} = \text{\path{left_weight_dual(A)}}$, $A^{\smallstar} =
\text{\path{right_weight_dual(A)}}$
\end{itemize}
In even-dimensional spaces ($EGA2D$, $PGA3DP$) we have to distinguish left and right
operations, while in odd-dimensional spaces ($EGA3D$, $PGA2DP$) we don't, since they
produce the same result. The ``left'' operations are typically used on the left-hand side
of the binary operation and the ``right'' operations on the right-hand side of the product
operator. The only deviation from that pattern occurs for the definition of the regressive
operations, as defined in~(\ref{eq:rwdg_std}) or (\ref{eq:rwdg_alt}). \\

In a two-dimensional algebra, like $EGA2D$, the effect of the right dual of a vector $u$
is to turn the vector by $+90$°, i.e. into the orientation of $\bv{12}$, while the left
dual turns the vector by $-90$°, i.e. into the direction against the orientation of
$\bv{12}$. This statement is valid for both right- and left-handed coordinate systems. \\

% potentially bring the explanation of the next couple of paragraphs later when all
% other concepts have been introduced

Before we go on to the geometric product, let's return to eqn.~(\ref{eq:plane_def_pga3dp})
for a moment, which provided the definition of a plane as trivector in projective space.
We are going to use this example to show what can be done geometrically using the
definitions provided above: \\

The trivector defining a plane can be created in multiple ways, e.g. either by wedge
products between three unitized projective points $P, Q, R$ that span the plane as $pl = P
\wedge Q \wedge R$ as their common subspace, or alternatively as wedge products between a
unitized projective point $P$ and two direction vectors $u$ and $v$ that span the plane as
a bivector such that $pl = P \wedge u \wedge v$. In the latter case, the plane $pl$ can be
interpreted as bound bivector as suggested by \cite{Browne_Grassmann-Algebra_Vol1:2012}.
For this case it is also easy to shift the plane to contain the unitized point $P+x$
instead of $P$. This can be achieved easily by applying the rules for the wedge product as
$pl_{shifted} = P \wedge u \wedge v + x \wedge u \wedge v = (P+x) \wedge u \wedge v$.\\

The attitude of the plane, i.e. its orientation in space, is given by a bivector that can
be calculated using the attitude function $att(pl)$. It is defined for $3D$ projective
space as $att(u) = u \vee \rcmpl{\bv{4}}$ and as such can be interpreted as the
intersection of the plane with the horizon $H_{3dp} =\rcmpl{\bv{4}}$. It yields the result
$att(pl) = t_x \bv{23} + t_y \bv{31} + t_z \bv{12}$.\\

The definition of a plane as given in eqn.~(\ref{eq:plane_def_pga3dp}) implies that the
normal direction $n$ of the plane $pl$ can be recovered from its trivector definition by
applying the left weight dual to the plane, such that $n = \leftweightdual{pl}$. The left
weight dual operation was introduced in eqn.~(\ref{eq:leftweightdual}). It can also be
found in table \ref{tab:pga3dp_bulk_weight_duals}. \\

The reverse is also true, i.e. if the plane normal $n$ is known, the trivector
representing the plane can be directly found by applying the right bulk dual operation to
the normal $n$, such that $pl_O = \rightbulkdual{n}$. The right bulk dual operation was
defined in eqn.~(\ref{eq:rightbulkdual}), and can be found in table
\ref{tab:pga3dp_bulk_weight_duals}. This operation yields a plane that contains the origin
and has the required orientation with $n$ as normal. The signed distance to the origin
could now be given as coefficient $t_w$ to the bulk component \bv{312} in case the plane
needs to be shifted parallel. \\

As demonstrated, this enables quite a number of relevant geometric operations by applying
rather simple operations to the modelled objects in $3D$ $PGA$.


\newpage

\subsubsection{Geometric products}
\label{subsection:geometric_products}

Now let's move on in more detail to the geometric products. The multiplicative identity of
the geometric product and the wedge product is the scalar basis element $\bm{1}$:
\begin{subeqnarray}
    \bm{1} \wedgedot a = a \wedgedot \bm{1} = a
    \label{eq:wedgedot_identity} \\
    \bm{1} \wedge a = a \wedge \bm{1} = a
    \label{eq:wedge_identity} 
\end{subeqnarray}

The multiplicative identity of the regressive geometric product and the regressive wedge
product is the pseudoscalar basis element $\ps = I_n$ of the respective algebra:
\begin{subeqnarray}
    \ps \veedot a = a \veedot \ps = a
    \label{eq:veedot_identity} \\
    \ps \vee a = a \vee \ps = a
    \label{eq:vee_identity}
\end{subeqnarray}

The geometric product combines the dimensions that are present in an object
representation, e.g. for $PGA2DP$:
\begin{eqnarray}
    \label{eq:gpr_combine_what_is_there}
    \bv{1} \wedgedot \bv{1} = 1 \nonumber\\
    \bv{2} \wedgedot \bv{2} = 1 \\
    \bv{3} \wedgedot \bv{3} = 0 \nonumber
\end{eqnarray}

The regressive geometric product combines the dimensions that are \lcmpl{not} present
in an object
representation, e.g. for $PGA2DP$:
\begin{eqnarray}
    \label{eq:rgpr_combine_what_is_not_there}
    \rcmpl{\bv{1}} \veedot \rcmpl{\bv{1}} = \ps \nonumber\\
    \rcmpl{\bv{2}} \veedot \rcmpl{\bv{2}} = \ps \\
    \rcmpl{\bv{3}} \veedot \rcmpl{\bv{3}} = 0 \nonumber
\end{eqnarray}
where eqs.~(\ref{eq:rgpr_combine_what_is_not_there}) can be shown to be equivalent to
eqs.~(\ref{eq:gpr_combine_what_is_there}) by taking the left complement of both sides and
applying the complement definition~(\ref{eq:rwdg_std}) to it. \\

The inner product between multivectors $A$ and $B$ is defined as
\begin{equation}
    A \cdot B = A^{T} \metric{G} B
    \label{eg:inner_product_metric}
\end{equation}
with $\metric{G}$ as the extended metric and using standard matrix multiplication on the
right-hand side. It satisfies the identity
\begin{equation}
    A \cdot B  = \grprj{A \wedgedot \tilde{B}}{0} = \grprj{ B \wedgedot \tilde{A} }{0}
    \label{eq:inner_product_link_to_gpr}
\end{equation}
where $\wedgedot$ stands for the geometric product within the grade projection operator
\grprj{}{k} for grade $k$. \\
\\

Using the various notations used in literature, the contraction is defined as
\begin{subequations}
    \begin{align}
    A \lcontr B & = A \ll B = A_{\star} \vee B
    \label{eq:lcontr} \\
    B \rcontr A & = B \gg A = B \vee A^{\star}
    \label{eq:rcontr}
    \end{align}
\end{subequations}
with $\star$ denoting the Hodge dual which is formed by the respective complement
operation (in spaces of even dimension the left or right complement respectively, and in
spaces of odd dimension the complement function regardless of the side of the operand).
For blades A and B, they satisfy
\begin{subequations}
    \begin{align}
    A \lcontr B &  = A \ll B = \grprj{B \wedgedot \tilde{A}}{gr(B)-gr(A)}
    \label{eq:lcontr_gpr} \\
    B \rcontr A & = B \gg A = \grprj{\tilde{A} \wedgedot B}{gr(B)-gr(A)}
    \label{eq:rcontr_gpr}
    \end{align}
\end{subequations}
and fulfill the requirement that $A \lcontr B = A \rcontr B = A \cdot B$ whenever $A$
and $B$ have the same grade. In this case the contractions reduce to the inner product. \\


The geometric product between a vector $v$ and a blade $B$ is defined in terms of interior
and exterior products, i.e. the right contraction ``$\rcontr$'', the left contraction
``$\lcontr$'' and wedge product ``$\wedge$'' as follows:
\begin{subequations}
    \begin{align}
    v \wedgedot B & =  B \rcontr v + v \wedge B
    \label{eq:gpr_Brc} \\
    B \wedgedot v & =  v \lcontr B + B \wedge v
    \label{eq:gpr_Blc}
    \end{align}
\end{subequations}
or sometimes expressed using the right shift ``$\gg$'' or left shift ``$\ll$'' operators
for the contractions (as implemented in the ga library)
\begin{subequations}
    \begin{align}
    v \wedgedot B & = B \gg v + v \wedge B
    \label{eq:gpr_Bshrc} \\
    B \wedgedot v & =  v \ll B + B \wedge v
    \label{eq:gpr_Bshlc}
    \end{align}
\end{subequations}
or sometimes by exclusively using the operations of the exterior algebra (wedge product
``$\wedge$'', bulk dual ``$\star$'' and regressive wedge product ``$\vee$'')
\begin{subequations}
    \label{eq:gpr_Bsh_exterior}
    \begin{align}
    v \wedgedot B & = B \vee \rightbulkdual{v} + v \wedge B
    \slabel{eq:gpr_Bshrc_exterior} \\
    B \wedgedot v & = \leftbulkdual{v} \vee B + B \wedge v
    \slabel{eq:gpr_Bshlc_exterior}
    \end{align}
\end{subequations}
as an alternative. Each one is equivalent. \\

Be careful, when using the equivalent formulas like the right-hand side of
equations~(\ref{eq:gpr_Bsh_exterior}) when deriving regressive products. Resulting
formulas may not be valid! The equivalent formulas for the regressive products must be
derived directly from the regressive product table. To give one example, here is the
corresponding formula for the regressive geometric product for the case of a regressive
product between a vector $v$ and a bivector $B$:
\begin{subequations}
    \begin{align}
    v \veedot B & = -\leftweightdual{B} \wedge v + \mathrm{rcmt}(v,B) \\
    B \veedot v & = -v \wedge \rightweightdual{B}  + \mathrm{rcmt}(B,v)
    \end{align}
\end{subequations}
It is obviously not the result one would get from creating the regressive versions
directly from the right-hand side of equations~(\ref{eq:gpr_Bsh_exterior}). The reason is
that the specific pre-conditions for the validity of the equivalence between the geometric
product and the corresponding right-hand sides may change when taking the complements. \\


Every vector $v$ can be decomposed into parts $v_{\parall}$ parallel to a $k$-blade $B$
and $v_{\perp}$ perpendicular to $B$, such that $v = v_{\parall} + v_{\perp}$. For
equation {(\ref{eq:gpr_Bshrc})}: $B \gg v$ is a $(k-1)$-blade. If $v_{\parall} \neq 0$
then $B \gg v$ represents a subspace of $B$. $v \wedge B$ is a $(k+1)$-blade. If
$v_{\perp} \neq 0$ then $v \wedge B$ represents $span(v,B)$.

For arguments of equal grade the contractions reduce to the dot product, so that one can
write specifically for two vectors $a$ and $b$:
\begin{equation}
    a \wedgedot b =  a \cdot b + a \wedge b
\end{equation}


\newpage

\subsubsection{Transwedge products}
\label{subsection:transwedge_products}

\cite{Lengyel_transwedge_product:2025} has shown that the geometric product can be derived
by using operations from Grassmann algebra exclusively. These are the wedge product, the
complement and the duality operations. \emph{This demonstrates that the geometric product
is not a fundamental operation in its own right, but can be derived from more fundamental
operations defined in Grassmann algebra.} \\

For that purpose Lengyel introduced the transwedge product of order $k$ by following
definition:
\begin{equation}
    a \underset{k}{\doublewedge} b = \sum_{c \, \in \, \mathcal{B}_k}
    (\lcmpl{c}\vee a) \wedge (b \vee \rightbulkdual{c})
    \label{eq:twdg_orig}
\end{equation}
where $\mathcal{B}_k$ is the set of all basis elements of order $k$. For $EGA3D$ it means
for example $\mathcal{B}_0 = \{1\}$, $\mathcal{B}_1 = \{\bv{1}, \bv{2}, \bv{3}\}$,
$\mathcal{B}_2 = \{\bv{23}, \bv{31}, \bv{12}\}$, and $\mathcal{B}_3 = \{\bv{123}\}$. From
these expressions the geometric product can be formed by
\begin{equation}
    a \wedgedot b = \sum_{k \, = \, 0}^{n} (-1)^{k(k-1)/2} a \underset{k}{\doublewedge} b
    \label{eq:grp_twdg}
\end{equation}

For $k=0$ this yields the outer product (wedge product). For higher values of $k$ other
products are generated which sum up to the full geometric product. Thus, the transwedge
products helps to decompose the geometric product. The most interesting for applications
are the wedge product resulting for $k=0$ and the transwedge product of order $k=1$. \\

The regressive wedge product in equation~(\ref{eq:twdg_orig}) can be replaced by the wedge
product using the complement-based definitions given in equations~(\ref{eq:double_cmpl})
to (\ref{eq:wdg_alt}). That means for example that equation~(\ref{eq:twdg_orig}) can be
transformed to the equivalent expression
\begin{equation}
    a \underset{k}{\doublewedge} b = \sum_{c \, \in \, \mathcal{B}_k}
    (\lcmpl{c \wedge \rcmpl{a}}) \wedge
    (\lcmpl{\rcmpl{b} \wedge \rcmpl{\rightbulkdual{c}}})
    \label{eq:twdg_wdg}
\end{equation}
expressing it by using the wedge product exclusively. \\

Since every product can by assigned a corresponding regressive product in analogy to
definition~(\ref{eq:rwdg_std}) or~(\ref{eq:rwdg_alt}), the transwedge
product~(\ref{eq:twdg_orig}) also has a regressive form:
\begin{equation}
    a \underset{k}{\doublevee} b \equiv
    \rcmpl{\lcmpl{a} \underset{k}{\doublewedge} \lcmpl{b}} =
    \sum_{c \, \in \, \mathcal{B}_k} \rcmpl{
    (\lcmpl{c}\vee \lcmpl{a}) \wedge (\lcmpl{b} \vee \rightbulkdual{c})}
    \label{eq:rtwdg_orig}
\end{equation}
which in turn can be expressed in terms of the wedge product exclusively as:
\begin{equation}
    a \underset{k}{\doublevee} b = \sum_{c \, \in \, \mathcal{B}_k}  \rcmpl{
    (\lcmpl{c\wedge a}) \wedge (\lcmpl{b \wedge \rcmpl{\rightbulkdual{c}}})}
    \label{eq:rtwdg_wdg}
\end{equation}
Equations~(\ref{eq:twdg_wdg}) and (\ref{eq:rtwdg_wdg}) are used for calculation of the
product tables of the (regressive) transwedge products of order $k$ and the resulting
(regressive) geometric products in \path{ga/ga_prdxpr}. \\

The transwedge product is very useful to show that the geometric product can be derived
from primitives of Grassmann algebra. When (regressive) transwedge product expressions are
calculated, it can be shown that each transwedge product and the resulting coefficient
expressions are equivalent to specific combinations of wedge, commutator or dot products,
and their regressive counterparts. This was done in
\path{ga/src_prdxpr/ga_pga2dp_ops_products.hpp} for the (regressive) transwedge product of
order $k=1$ and as well in \path{ga/src_prdxpr/ga_pga3dp_ops_products.hpp} to show the
equivalence.

\newpage

\subsubsection{Dot products, complements, duals, and reverses in \textit{PGA}}
\label{subsection:tables_of_products}

% Problem: \begin{table}...\end{table} is a float environment. LaTeX floats cannot be
% placed inside boxes — including \rotatebox{90}{...} with a minipage. This causes an
% error because LaTeX tries to defer float placement but has no valid location inside a
% box.
%
% Fix: Removed \begin{table} / \end{table} (commented them out to preserve intent)
% Replaced all four \caption{...} calls with \captionof{table}{...} Why \captionof works:
% It generates a caption with the correct counter and formatting for a table, but without
% requiring a table float environment. The caption package (already loaded in your
% preamble via \usepackage{caption}) provides this command — no new package needed.
%
% The \label commands after each \captionof remain unchanged and will still work correctly
% for \ref cross-references.

\rotatebox{90}{%
\begin{minipage}{0.9\textheight}
\centering % a single '\centering' directive suffices

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% pga2dp from here on
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \begin{sidewaystable}
% \begin{table}
    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$             & $\bm{1}$
                         & $\bv{1}$       & $\bv{2}$      & $\bv{3}$
                         & $\bv{31}$      & $\bv{32}$     & $\bv{12}$
                         & $\ps$
                         \\ \hline
    $\bm{u}\cdot\bm{u}$  & $\bm{1}$
                         & $\bm{1}$       & $\bm{1}$      & $0$
                         & $0$            & $0$           & $\bm{1}$
                         & $0$
                         \\ \hline
    $\bm{u}\circ\bm{u}$  & $0$
                         & $0$            & $0$           & $\ps$
                         & $\ps$          & $\ps$         & $0$
                         & $\ps$
                         \\ \hline
    \end{tabular}
    \captionof{table}{\small Dot and regressive dot products in $2D$ $PGA$ for every
    basis element.}
    \label{tab:pga2dp_dot_products}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$          & $\bm{1}$
                      & $\bv{1}$      & $\bv{2}$      & $\bv{3}$
                      & $\bv{31}$     & $\bv{32}$     & $\bv{12}$
                      & $\ps$
                      \\ \hline
    $\rcmpl{\bm{u}}$  & $\ps$
                      & $\bv{32}$     & $-\bv{31}$    & $-\bv{12}$
                      & $-\bv{2}$     & $ \bv{1} $    & $-\bv{3}$
                      & $\bm{1}$
                      \\ \hline
    \end{tabular}
    \captionof{table}{\small Complements in $2D$ $PGA$ for every basis element.}
    \label{tab:pga2dp_complement}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$                    & $\bm{1}$
                                & $\bv{1}$      & $\bv{2}$     & $\bv{3}$
                                & $\bv{31}$     & $\bv{32}$    & $\bv{12}$
                                & $\ps$
                                \\ \hline
    $\rightbulkdual{\bm{u}}$    & $\ps$
                                & $\bv{32}$     & $-\bv{31}$   & $0$
                                & $0$           & $0$          & $-\bv{3}$
                                & $0$
                                \\ \hline
    $\rightweightdual{\bm{u}}$  & $0$
                                & $0$           & $0$          & $-\bv{12}$
                                & $-\bv{2}$     & $\bv{1}$     & $0$
                                & $\bm{1}$
                                \\ \hline
    \end{tabular}
    \captionof{table}{\small Bulk duals and weight duals for every basis element
    (=complement applied to bulk and weight).}
    \label{tab:pga2dp_bulk_weight_duals}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$        & $\bm{1}$
                    & $\bv{1}$       & $\bv{2}$      & $\bv{3}$
                    & $\bv{31}$      & $\bv{32}$     & $\bv{12}$
                    & $\ps$
                    \\ \hline
    $\rev{\bm{u}}$  & $\bm{1}$
                    & $\bv{1}$       & $\bv{2}$      & $\bv{3}$
                    & $-\bv{31}$     & $-\bv{32}$    & $-\bv{12}$
                    & $-\ps$
                    \\ \hline
    $\rrev{\bm{u}}$ & $-\bm{1}$
                    & $-\bv{1}$      & $-\bv{2}$     & $-\bv{3}$
                    & $\bv{31}$      & $\bv{32}$     & $\bv{12}$
                    & $\ps$
                    \\ \hline
    \end{tabular}
    \captionof{table}{\small Reverse and regressive reverse in $2D$ $PGA$ for every basis
    element.}
    \label{tab:pga2dp_rev_rrev}
    \end{center}
% \end{sidewaystable}
% \end{table}

\end{minipage}
} % end of scope of '\rotatebox' directive

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% pga3dp from here on
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{sidewaystable}
    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$             & $\bm{1}$
                         & $\bv{1}$    & $\bv{2}$      & $\bv{3}$     & $\bv{4}$
                         & $\bv{41}$   & $\bv{42}$     & $\bv{43}$ 
                         & $\bv{23}$   & $\bv{31}$     & $\bv{12}$
                         & $\bv{423}$  & $\bv{431}$    & $\bv{412}$   & $\bv{321}$
                         & $\ps$
                         \\ \hline
    $\bm{u}\cdot\bm{u}$  & $\bm{1}$        
                         & $\bm{1}$    & $\bm{1}$      & $\bm{1}$     & $0$
                         & $0$         & $0$           & $0$
                         & $\bm{1}$    & $\bm{1}$      & $\bm{1}$ 
                         & $0$         & $0$           & $0$          & $\bm{1}$
                         & $0$
                         \\ \hline
    $\bm{u}\circ\bm{u}$  & $0$        
                         & $0$         & $0$           & $0$          & $\ps$
                         & $\ps$       & $\ps$         & $\ps$
                         & $0$         & $0$           & $0$ 
                         & $\ps$       & $\ps$         & $\ps$        & $0$
                         & $\ps$
                         \\ \hline
    \end{tabular}
    \caption{\small Dot and regressive dot products in $3D$ $PGA$ for every basis
    element.}
    \label{tab:pga3dp_dot_products}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$         & $\bm{1}$
                     & $\bv{1}$       & $\bv{2}$      & $\bv{3}$     & $\bv{4}$
                     & $\bv{41}$      & $\bv{42}$     & $\bv{43}$ 
                     & $\bv{23}$      & $\bv{31}$     & $\bv{12}$
                     & $\bv{423}$     & $\bv{431}$    & $\bv{412}$   & $\bv{321}$
                     & $\ps$
                     \\ \hline
    $\rcmpl{\bm{u}}$ & $\ps$        
                     & $\bv{423}$    & $\bv{431}$    & $\bv{412}$    & $\bv{321}$
                     & $-\bv{23}$    & $-\bv{31}$    & $-\bv{12}$
                     & $-\bv{41}$    & $-\bv{42}$    & $-\bv{43}$ 
                     & $-\bv{1}$     & $-\bv{2}$     & $-\bv{3}$     & $-\bv{4}$
                     & $\bm{1}$
                     \\ \hline
    $\lcmpl{\bm{u}}$ & $\ps$        
                     & $-\bv{423}$   & $-\bv{431}$   & $-\bv{412}$   & $-\bv{321}$
                     & $-\bv{23}$    & $-\bv{31}$    & $-\bv{12}$
                     & $-\bv{41}$    & $-\bv{42}$    & $-\bv{43}$ 
                     & $\bv{1}$      & $\bv{2}$      & $\bv{3}$      & $\bv{4}$
                     & $\bm{1}$
                     \\ \hline
    \end{tabular}
    \caption{\small Right and left complements in $3D$ $PGA$ for every basis element.}
    \label{tab:pga3dp_complement}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$                   & $\bm{1}$
                               & $\bv{1}$      & $\bv{2}$      & $\bv{3}$    & $\bv{4}$
                               & $\bv{41}$     & $\bv{42}$     & $\bv{43}$ 
                               & $\bv{23}$     & $\bv{31}$     & $\bv{12}$
                               & $\bv{423}$    & $\bv{431}$    & $\bv{412}$  & $\bv{321}$
                               & $\ps$
                               \\ \hline
    $\rightbulkdual{\bm{u}}$   & $\ps$        
                               & $\bv{423}$    & $\bv{431}$    & $\bv{412}$  & $0$
                               & $0$           & $0$           & $0$
                               & $-\bv{41}$    & $-\bv{42}$    & $-\bv{43}$ 
                               & $0$           & $0$           & $0$         & $-\bv{4}$
                               & $0$
                               \\ \hline
    $\leftbulkdual{\bm{u}}$    & $\ps$        
                               & $-\bv{423}$   & $-\bv{431}$   & $-\bv{412}$ & $0$
                               & $0$           & $0$           & $0$
                               & $-\bv{41}$    & $-\bv{42}$    & $-\bv{43}$ 
                               & $0$           & $0$           & $0$         & $\bv{4}$
                               & $0$
                               \\ \hline
    $\rightweightdual{\bm{u}}$ & $0$        
                               & $0$          & $0$    & $0$  & $\bv{321}$
                               & $-\bv{23}$   & $-\bv{31}$    & $-\bv{12}$
                               & $0$          & $0$           & $0$ 
                               & $-\bv{1}$    & $-\bv{2}$     & $-\bv{3}$    & $0$
                               & $\bm{1}$
                               \\ \hline
    $\leftweightdual{\bm{u}}$  & $0$        
                               & $0$          & $0$           & $0$          & $-\bv{321}$
                               & $-\bv{23}$   & $-\bv{31}$    & $-\bv{12}$
                               & $0$          & $0$           & $0$
                               & $\bv{1}$     & $\bv{2}$      & $\bv{3}$     & $0$
                               & $\bm{1}$
                               \\ \hline
    \end{tabular}
    \caption{\small Right/left bulk duals and right/left weight duals in $3D$ $PGA$
    for every basis element (=complement applied to bulk and weight).}
    \label{tab:pga3dp_bulk_weight_duals}
    \end{center}

    \begin{center}
    %\scriptsize
    \begin{tabular}{*{17}{|c}|}
    \hline
    \rowcolor{blue!20}
    $\bm{u}$        & $\bm{1}$
                    & $\bv{1}$       & $\bv{2}$      & $\bv{3}$      & $\bv{4}$
                    & $\bv{41}$      & $\bv{42}$     & $\bv{43}$ 
                    & $\bv{23}$      & $\bv{31}$     & $\bv{12}$
                    & $\bv{423}$     & $\bv{431}$    & $\bv{412}$    & $\bv{321}$
                    & $\ps$
                    \\ \hline
    $\rev{\bm{u}}$  & $\bm{1}$     
                    & $\bv{1}$       & $\bv{2}$      & $\bv{3}$      & $\bv{4}$
                    & $-\bv{41}$     & $-\bv{42}$    & $-\bv{43}$ 
                    & $-\bv{23}$     & $-\bv{31}$    & $-\bv{12}$
                    & $-\bv{423}$    & $-\bv{431}$   & $-\bv{412}$   & $-\bv{321}$
                    & $\ps$
                         \\ \hline
    $\rrev{\bm{u}}$ & $\bm{1}$    
                    & $-\bv{1}$      & $-\bv{2}$     & $-\bv{3}$     & $-\bv{4}$
                    & $-\bv{41}$     & $-\bv{42}$    & $-\bv{43}$ 
                    & $-\bv{23}$     & $-\bv{31}$    & $-\bv{12}$
                    & $\bv{423}$     & $\bv{431}$    & $\bv{412}$    & $\bv{321}$
                    & $\ps$
                        \\ \hline
    \end{tabular}
    \caption{\small Reverse and regressive reverse in $3D$ $PGA$ for every basis element.}
    \label{tab:pga3dp_rev_rrev}
    \end{center}
\end{sidewaystable}

\newpage