\subsection{Physics modelling}
\label{physics_modelling}

The key idea is to represent physical entities using objects that are represented directly
in the underlying algebra. They have a geometric interpretation and can be directly
manipulated by applying algebraic operations. \\

$EGA$ is limited to objects that contain the origin. Thus, it can be used to directly
model vectors (i.e. directions), lines, and planes through the orgin. In addition, it can
model complex numbers, quaternions and thus rotations in $2d$ and $3d$ well. To realize
them, it uses sandwich products based on the geometric product for calculating the
transformation resulting from two consequtive reflections either across lines ($2d$-case)
or across planes ($3d$-case). The resulting transformations are rotations. The advantage
of using geometric algebra to model those rotations is the intuitive modelling, due to
geometric meaning of the objects and the fact that the sandwich products can be used to
transform different objects, like vectors, bivectors or rotation operators using the same
sandwich product formulas in a unified fashion regardless of the type of object to be
transformed.\\

In addition, $PGA$ is capable to model objects that do not necessarily contain the origin.
Thus it can be used to directly model points, directions, lines, and planes that may or
may not contain the origin. Translations can be implemented in $PGA$ as an additional
orthogonal transformation using a sandwich product in the algebra. The operation is
realized by two consequtive reflections across parallel lines or planes. It is based on
the regressive form of the geometric product. The reason for this deviation from $EGA$,
which uses the geometric product for that purpose, is that the Euclidean isometries
(rotation, translation, and screw motion) can only be performed in a space that is dual to
the space we use for modelling our projective objects (for explanation
see~\cite{Lengyel_pga-illuminated:2024}). The use of the regressive geometric product on
these objects is equivalent to first transforming the object into dual space by applying
the complement operation, then using the geometric product in that space for the intended
transformation and finally transforming the resulting transformed object back into the
original space, by applying the inverse complement operation. Thus the sandwich product
transformation formulas in $PGA$ differ from the formulas used in $EGA$ by using the
regressive geometric product instead of the geometric product, but are otherwise the same.
\\

The additional degree of freedom in $PGA$ is possible due to embedding the Euclidean space
of dimension $n$ into a projective space of dimension $n+1$, e.g. two-dimensional
Euclidean space is modelled by embedding it in a three-dimensional projective space, and
three-dimensional Euclidean space is modelled in a four-dimensional projective space. In
the $n+1$-dimensional projective space all algebraic objects contain the origin (exactly
as in $EGA$), but in the modelled $n$-dimensional Euclidean space the objects do not need
to contain the origin. They can be located anywhere in the modelled space including
postions at infinity. In projective geometry the objects must be projected into the
modelled space. Thus the objects of the embedding space have one dimension more than the
corresponding objects in modelled space. A projective point (a scalar or $0d$-object) is a
vector representing a ray or line in the embedding space (a $1d$-object), and a bivector
in projective space (a $2d$-object) is used to model a line (a $1d$-object) in modelled
space, and so on for higher grade objects. The projection reduces the object dimension by
one. Objects modelled in projective space can be multiplied by a scalar value without
changing the geometrical meaninig of the object in the modelled Euclidean space. \\

Due to its extended expressiveness $PGA$ is suitable to model physics of mass points and
rigid bodies. For example it can model position, velocity, acceleration, force, torque,
momentum, angular momentum, etc., and thus it can be used to describe kinematics and
dynamics of mass points and rigid bodies. The approach used here is inspired by
\cite{Plane-based_PGA_Dorst-DeKennik:2022} and
\cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023}, but it does not use the
plane-based ansatz with planes as vectors, which are based on dual representations of the
objects to be modelled. It rather uses the approach as described by
\cite{Lengyel_pga-illuminated:2024} using projective geometric algebra based on direct
representations of points, lines and planes, who also gives reasons for choosing the
latter approach and explains the background in more detail compared to the short
indications given above. \\

\cite{Lengyel_poor-foundations_GA:2024} also points out potential inconsistencies in
applications of geometric algebra as often used up-to-now. Further hints on
inconsistencies in some GA approaches can be found as well in
\cite{Kritchevsky_case_against_GA:2024}. The goal of the this work is to be as consistent
and mathematically rigoros as possible implementing GA functionality, and at the same time
providing an intuitively applicable framework for modelling of physics. \\

\cite{Browne_Grassmann-Algebra_Vol1:2012} (see sections 4.1-4.5) is a very good source
showing how to use Grassmann algebra and projective geometry in order to model forces and
moments. Classical approaches based-on vector algebra only model forces by providing
vectors of adequate magnitude and direction to represent the force. However, the
physically very meaningful ``line of action'' is not modelled mathematically at all. It
rather needs to be derived from the description of the context in order to know where the
force is actually acting. \\

This changes in $PGA$ since any force can be directly modelled by its line of action which
can be fully encoded within the mathematical approach: A force is directly modelled by a
line $F$ in $PGA$. If $f$ is the force vector of corresponding magnitude and direction,
and $P$ is any point on its line of action then the force line of action is modelled as
\begin{equation}
F = P \wedge f
\end{equation}
The resulting bivector $F$ represents a line as geometric object which contains all
physically relevant information of the force $F$ and its effects, like moments generated
by the force at any point in space. \\

The force vector $f$ itself can be retrieved from $F$ by using the attitude function
$att()$:
\begin{equation}
    f = att(F) = F \vee \overline{\bv{n}}
    \label{force_attitude}
\end{equation}
with $\bv{n}$ denoting the basis vector of the projective dimension that squares to $0$ in
the respective algebra. The moment generated relative to the origin $O$ is encoded in the
bulk of the bivector:
\begin{equation}
    M_{origin} = bulk(F)
    \label{force_moment_origin}
\end{equation}
while the resulting moment from the application of $f$ at any given
point $R$ in space can be retrieved from $F$, either with any known point $P$ on $F$
(which is in principle always available via the $support(F)$ function which provides the
point on $F$ with the shortest distance to the origin $O$), or alternatively by projecting
$R$ onto $F$ and using the perpendicular distance vector $d_\perp$ from the line $F$
towards $R$ in order to calculate the resulting moment $M_F(R)$ as
\begin{subeqnarray}
    M_F(R) & = &  (R-P) \wedge f = (R-P) \wedge att(F)
    \slabel{eq:action_line_moment_simple} \\
    M_F(R) & = & d_\perp \wedge f  = (R - F \vee (R \wedge \rightweightdual{F})) \wedge
    att(F)
    \slabel{eq:action_line_moment_projection}
\end{subeqnarray}
where equation~(\ref{eq:action_line_moment_simple}) is the formula with the least
computational effort using a known point $P$, while
equation~(\ref{eq:action_line_moment_projection}) uses the othogonal projection of $R$
onto $F$ on the left-hand side to calculate the perpendicular distance vector $d_\perp$.
\\

% The geometrical situation is depicted in figure XXX for a two-dimensional case.

\newpage