\subsection{Modelling motion}
\label{subsection_modelling_motion}

For translational motion in physical contexts we typically use the following naming
convention for scalars or vectors:
\begin{align*}
s(t)        & = s \\
\dot{s}(t)  & = v(t) = \dot{s} = v \\
\ddot{s}(t) & = \dot{v}(t) = a(t) =\ddot{s} = \dot{v} = a
\end{align*}
while for rotational motion we typically use:
\begin{align*}
\phi(t)        & = \phi \\
\dot{\phi}(t)  & = \omega(t) = \dot{\phi} = \omega \\
\ddot{\phi}(t) & = \dot{\omega}(t) =\ddot{\phi} = \dot{\omega}
\end{align*}
Especially in $PGA$, we will be able to combine the translational and rotational aspects
in one physical variable. Thus, we will use a convention of using capital letters for
physical quantities to distinguish them from the classical convention, especially so for
bivectors and even-grade multivectors that do not occur in classical descriptions. We will
use in $PGA$:
\begin{align*}
B(t)        & = B \qquad \qquad \qquad  \quad \! \text{(bivector)} \\
\dot{B}(t)  & = \Omega(t) = \dot{B} = \Omega \qquad
                \text{(rate of change of bivector, ``bivector speed'')} \\
\ddot{B}(t) & = \dot{\Omega}(t) =\ddot{B} = \dot{\Omega} \qquad
                \text{(rate of change of bivector speed)}
\end{align*}

In order to model motion in $PGA$ we need to work with the exponential function applying
it in the context of the geometric product (symbol ``$\wedgedot$'') and the regressive
geometric product (symbol ``$\veedot$'').
% Exponential function
\begin{align}
\exp(x) &= \sum_{n=0}^{\infty} \frac{x^n}{n!} 
&= 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \frac{x^4}{4!} +
\frac{x^5}{5!} + \frac{x^6}{6!} + \frac{x^7}{7!} + \cdots
\label{eq:exp_taylor}
\end{align}
% Cosine function
\begin{align}
\cos(x) &= \sum_{n=0}^{\infty} \frac{(-1)^n x^{2n}}{(2n)!} 
&= 1 - \frac{x^2}{2!} + \frac{x^4}{4!} - \frac{x^6}{6!} + \frac{x^8}{8!}
- \frac{x^{10}}{10!} + \frac{x^{12}}{12!} - \frac{x^{14}}{14!} + \cdots
\label{eq:cos_taylor}
\end{align}
% Sine function
\begin{align}
\sin(x) &= \sum_{n=0}^{\infty} \frac{(-1)^n x^{2n+1}}{(2n+1)!} 
&= x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} + \frac{x^9}{9!}
- \frac{x^{11}}{11!} + \frac{x^{13}}{13!} - \frac{x^{15}}{15!} + \cdots
\label{eq:sin_taylor}
\end{align}

If we substitute $x = i \phi$ with $i$ as the imaginary unit ($i^2 = -1$), we get Euler's
formula:
% Exponential function with x = i*phi
\begin{align}
\exp(i\phi) &= \sum_{n=0}^{\infty} \frac{(i\phi)^n}{n!} \nonumber \\ 
&= 1 + i\phi + \frac{(i\phi)^2}{2!} + \frac{(i\phi)^3}{3!} + \frac{(i\phi)^4}{4!} +
   \frac{(i\phi)^5}{5!} + \frac{(i\phi)^6}{6!} + \frac{(i\phi)^7}{7!} +
   \cdots \\
&= 1 + i\phi - \frac{\phi^2}{2!} - i\frac{\phi^3}{3!} + \frac{\phi^4}{4!} +
   i\frac{\phi^5}{5!} - \frac{\phi^6}{6!} - i\frac{\phi^7}{7!} + \cdots
   \nonumber \\
&= 1 - \frac{\phi^2}{2!} + \frac{\phi^4}{4!} - \frac{\phi^6}{6!} + \cdots
   + i (\phi - \frac{\phi^3}{3!} + \frac{\phi^5}{5!} - \frac{\phi^7}{7!} +
   \cdots) \\[10pt]
\exp(i\phi) &= \cos(\phi) + i\sin(\phi)
\end{align}

Applying this for geometric algebra we have to distinguish between calculating the powers
in the Taylor series using the geometric product (in $EGA$) and using the regressive
geometric product (in $PGA$). For the rotation case we assume a unitized bivector $B$ and
a scalar $\phi$. Based on the geometric product we know that $B \wedgedot B = -\bm{1}$,
and for its regressive counterpart we know that $B \veedot B = -\ps$. Thus, $B$ plays a
role equivalent to $i$ in both cases. In addition, taking any value to the power of $0$
results in the neutral element of the product, which is $\bm{1}$ for the geometric
product, and $\ps$ for the regressive geometric product. Taking the powers then works like
this:
\begin{subeqnarray}
\cdots \nonumber \\[5pt]
\frac{(B\phi)^2_\wedgedot}{2!} & = & \frac{1}{2!} (B \phi \wedgedot B \phi) =
\frac{\phi^2}{2!}  (B \wedgedot B) = \frac{\phi^2}{2!}  (-\bm{1})= - \frac{\phi^2}{2!} \\
\frac{(B\phi)^3_\wedgedot}{3!} & = & \frac{\phi^3}{3!} (B \wedgedot B \wedgedot B) =
\frac{\phi^3}{3!} (-\bm{1} \wedgedot B) = - \frac{\phi^3}{3!}  B \\
\frac{(B\phi)^4_\wedgedot}{4!} & = & \frac{\phi^4}{4!} (B \wedgedot B \wedgedot B
\wedgedot B) =
\frac{\phi^4}{4!} (-\bm{1} \wedgedot -\bm{1}) = \frac{\phi^4}{4!} \\[5pt]
\cdots \nonumber \\[5pt]
\frac{(B\phi)^2_\veedot}{2!} & = & \frac{1}{2!} (B \phi \veedot B \phi) =
\frac{\phi^2}{2!} (B \veedot B) = \frac{\phi^2}{2!}  (-\ps)= - \frac{\phi^2}{2!} \ps \\
\frac{(B\phi)^3_\veedot}{3!} & = & \frac{\phi^3}{3!} B \veedot B \veedot B  =
\frac{\phi^3}{3!} (-\ps \veedot B) = - \frac{\phi^3}{3!} B \\
\frac{(B\phi)^4_\veedot}{4!} & = & \frac{\phi^4}{4!} B \veedot B \veedot B \veedot B =
\frac{\phi^4}{4!} (-\ps \veedot -\ps) = \frac{\phi^4}{4!} \ps \\[5pt]
\cdots \nonumber 
\end{subeqnarray}

Thus, the result of exponentiating a bivector $B$ with respect to the geometric product is
(for $EGA$):
% Geometric Product Version
\begin{align}
\exp_\wedgedot(B\phi) &= \sum_{n=0}^{\infty} \frac{(B\phi)^n_\wedgedot}{n!} \nonumber \\ 
&= 1 + B\phi + \frac{(B\phi)^2_\wedgedot}{2!} + \frac{(B\phi)^3_\wedgedot}{3!} +
\frac{(B\phi)^4_\wedgedot}{4!} + \frac{(B\phi)^5_\wedgedot}{5!} +
\frac{(B\phi)^6_\wedgedot}{6!} + 
\cdots \\
&= 1 + B\phi - \frac{\phi^2}{2!} - B\frac{\phi^3}{3!} + \frac{\phi^4}{4!} +
   B\frac{\phi^5}{5!} - \frac{\phi^6}{6!} - B\frac{\phi^7}{7!} + \cdots
   \nonumber \\
&= 1 - \frac{\phi^2}{2!} + \frac{\phi^4}{4!} - \frac{\phi^6}{6!} + \cdots
   + B (\phi - \frac{\phi^3}{3!} + \frac{\phi^5}{5!} - \frac{\phi^7}{7!} +
   \cdots)  \nonumber \\[10pt]
\exp_\wedgedot(B\phi) &= \cos(\phi) + \sin(\phi) B
\label{eq:exp_B}
\end{align}
The last step of moving $B$ after the brackets is allowed, since scalar factors commute
with any object using the regular product. The result structurally is an even-grade
multivector with a scalar and a bivector component. We get the correct sign and angle of
rotation if we replace $\phi$ by $-\frac{1}{2}\phi$, which was not considered above to
keep the formulas as simple as possible. The substitution yields the final form of a rotor
$R$ in $EGA$ with $B$ as unitized bivector and $\phi$ the rotation angle:
\begin{align}
    R = \exp_\wedgedot(-\frac{\phi}{2}B)
      &= \cos(-\frac{\phi}{2}) + \sin(-\frac{\phi}{2}) B \nonumber \\
    R &= \cos(\frac{\phi}{2}) - \sin(\frac{\phi}{2}) B 
    \label{eq:rotor_ega}
\end{align}

For the case using the regressive geometric product it becomes (for $PGA$):
% Regressive Geometric Product Version
\begin{align}
\exp_\veedot(B\phi) &= \sum_{n=0}^{\infty} \frac{(B\phi)^n_\veedot}{n!} \nonumber \\ 
&= \ps + B\phi + \frac{(B\phi)^2_\veedot}{2!} + \frac{(B\phi)^3_\veedot}{3!} +
   \frac{(B\phi)^4_\veedot}{4!} + \frac{(B\phi)^5_\veedot}{5!} +
   \frac{(B\phi)^6_\veedot}{6!} + 
   \cdots \\
&= \ps + B\phi - \ps\frac{\phi^2}{2!} - B\frac{\phi^3}{3!} + \ps\frac{\phi^4}{4!} +
   B\frac{\phi^5}{5!} - \ps\frac{\phi^6}{6!} - B\frac{\phi^7}{7!} + \cdots
   \nonumber \\
&= \ps (1 - \frac{\phi^2}{2!} + \frac{\phi^4}{4!} - \frac{\phi^6}{6!} + \cdots) 
   + B (\phi - \frac{\phi^3}{3!} + \frac{\phi^5}{5!} - \frac{\phi^7}{7!} +
   \cdots)  \nonumber \\[10pt]
\exp_\veedot(B\phi) &= \cos(\phi)\ps + \sin(\phi) B
\label{eq:rexp_B}
\end{align}
The last step of moving $\ps$ and $B$ after the brackets is allowed, since scalar factors,
like the results of the trigonometric functions, commute with any object using the regular
product. The resulting motor structurally is an even-grade multivector with a bivector and
a pseudoscalar of grade 4 for $3D$ $PGA$. \\

To create a rotor that generates a rotation of angle $\phi$ we use
\begin{equation}
   M = \exp_\veedot{\left( \frac{\phi}{2} B  \right)}
\end{equation}
where the factor $\frac{1}{2}$ is needed to create a resulting angle $\phi$ through the
double-sided sandwich product. Typically, in applications the angle $\phi$ is directly
contained as a scalar factor in the bivector $B$ while the factor $\frac{1}{2}$ is kept
explicit. \\

In $2D$ $PGA$ we have the same expression, but the argument of the exponential to create a
motor for rotation is not a bivector $B$, but a unitized point $P$. Here $P \veedot P =
-\ps$ is valid, and equation~(\ref{eq:rexp_B}) becomes $\exp_\veedot(P\phi) =
\cos(\phi)\ps + \sin(\phi) P$ instead, which leads to a motor $M$ that structurally is an
odd-grade multivector with a vector and a pseudoscalar of grade 3. The resulting rotation
direction of the motor depends on the chosen pseudoscalar. The pseudoscalar for $PGA2DP$
was chosen as \bv{321} in order to make signs of rotation angles (and thus turning
directions) consistent between $2D$ and $3D$ $PGA$. For the actual motor we get
\begin{equation}
      M = \exp_\veedot{\left( \frac{\phi}{2} P  \right)}
\end{equation}
in this case. \\

To model a translation $\delta = (\delta_x, \delta_y, \delta_z, 0)$ in $3D$ $PGA$ we use a
special bivector $B=B_{tra}$ that does not square to $-\ps$, but instead squares to zero.
That makes sure that the expansion of the Taylor series of the exponential function
$\exp_{\veedot}(B_{tra})$ only contains the first two terms. The bivector is resulting
from two consecutive reflections across parallel planes perpendicular to $\delta$ with the
distance $\frac{1}{2}\bulknrm{\delta}$ relative to each other. They intersect in a line at
infinity that is used as ``rotation axis''. This however yields a translation, since the
axis is infinitely far away. The derivation of the formula for the bivector will be done
later in this chapter (see eqn.~(\ref{eq:bivector_trans_rate_v_vec})). We use $B_{tra} =
\rightbulkdual{\delta} \vee \rcmpl{\bv{4}} = \rightbulkdual{\delta} \vee H_{3dp} =
att(\rightbulkdual{\delta})$. This yields $B_{tra} = (\delta_x \bv{23} + \delta_y \bv{31}
+ \delta_z \bv{12})$. The resulting motor is $M = \exp_\veedot(\frac{1}{2} B_{tra}) = \ps
+ \frac{1}{2}B_{tra} = \ps + \frac{1}{2}(0, 0, 0, \delta_x, \delta_y, \delta_z)$. The
motor $M$ is an even-grade multivector containing a pseudoscalar component, which is a
quadvector in $3D$ $PGA$, and a bivector component.\\

To model a translation $\delta = (\delta_x, \delta_y, 0)$ in $2D$ $PGA$ we need a special
vector as argument of the exponential. It represents a direction to a point at infinity.
The ``rotation'' around that infinitely far away point generates a translation (actually
it is resulting from two consecutive reflections across parallel lines perpendicular to
$\delta$ with the distance $\frac{1}{2}\bulknrm{\delta}$ relative to each other). The
derivation of the special vector will be done later in this section (see
eqn.~(\ref{eq:vector_trans_rate_v_vec})). The argument of the exponential $\exp_\veedot$
for a translation is the direction vector $P = \rightbulkdual{\delta} \vee \rcmpl{\bv{3}}
= \rightbulkdual{\delta} \vee H_{2dp} = att(\rightbulkdual{\delta}) = (-\delta_y \bv{1} +
\delta_x \bv{2})$. $P$ points in a direction perpendicular to the translation $\delta$
towards infinity. The main difference of this special point $P$ compared to the one used
for the rotation above is that $P_z = 0$. This results in $P \veedot P = 0$ which causes
the expansion of the exponential to only contain the first two terms: $M =
\exp_\veedot(\frac{1}{2} P) = \ps + \frac{1}{2}P = \ps +
\frac{1}{2}(-\delta_y,\delta_x,0)$. The motor $M$ is an odd-grade multivector containing a
pseudoscalar, which is a trivector in $2D$ $PGA$, and a vector component, i.e. it is
structurally the same as for the rotation case in $2D$ $PGA$ above.\\

Eqn.~(\ref{eq:exp_B}) allows us to model rotations in $EGA$, where the bivector $B$ is the
complement of the axis of rotation and $\theta = -2\phi$ is the angle of the rotation
around that axis. If we directly want to specify the angle of rotation $\theta$ we need to
substitute $\phi$ by $-\frac{\phi}{2}$ in eqn. (\ref{eq:exp_B}). The rotated object $X$
can then be obtained from the original object $X_0$ by:
\begin{equation}
    X = R \wedgedot X_0 \wedgedot \rev{R} = \exp_\wedgedot(-\frac{\phi}{2} B )
    \wedgedot X_0 \wedgedot \exp_\wedgedot(\frac{\phi}{2} B) \quad (\text{for $EGA$})
\end{equation}

If we want to model combined translations and rotations in $PGA$, we need to look into
another extension of the exponential function. We follow the approach defined by
\cite{Lengyel_pga-illuminated:2024} in chapter 3.2 and 3.6.2, where it is shown that a
general motor in $PGA$ can be expressed as a screw motion, i.e. a rotation around a
unitized line $\bm{l}$ by an angle $\phi$ and a translation along $\bm{l}$ by a
displacement $\delta$ as
\begin{eqnarray}
    M = \exp_\veedot\left[(\frac{\delta}{2} \bm{1} + \frac{\phi}{2}\ps) \veedot \bm{l}
    \right]=
    \cos_\veedot(\frac{\delta}{2} \bm{1} + \frac{\phi}{2} \ps) +
    \sin_\veedot(\frac{\delta}{2} \bm{1} + \frac{\phi}{2} \ps) \veedot \bm{l} \nonumber \\
    = -\frac{\delta}{2} \bm{1} \sin(\frac{\phi}{2}) + \ps\cos(\frac{\phi}{2}) + \left(
     \frac{\delta}{2} \bm{1} \cos(\frac{\phi}{2}) + \ps\sin(\frac{\phi}{2}) \right)
     \veedot \bm{l} \nonumber \\
  M = -\frac{\delta}{2}\bm{1}\sin(\frac{\phi}{2}) - \frac{\delta}{2}
      \rightweightdual{\bm{l}}\cos(\frac{\phi}{2}) +
      \bm{l}\sin(\frac{\phi}{2}) + \ps\cos(\frac{\phi}{2}) 
\label{eq:rexp_pga_motor}
\end{eqnarray}
where $M$ structurally is an even grade multivector with scalar, bivector and pseudoscalar
parts in $3D$ $PGA$. \\

For pure rotational motion with $\delta = 0$ and $\phi \ne 0$ we get
\begin{eqnarray}
  M_{rot} = \bm{l}\sin(\frac{\phi}{2}) + \ps\cos(\frac{\phi}{2}) 
\label{eq:rexp_pga_motor_rotation}
\end{eqnarray}
while for pure translational motion with $\delta \ne 0$ and $\phi = 0$ we get
\begin{eqnarray}
  M_{tra} = -\frac{\delta}{2}\rightweightdual{\bm{l}} + \ps 
\label{eq:rexp_pga_motor_translation}
\end{eqnarray}


The argument in the exponential function in eqn.~(\ref{eq:rexp_pga_motor}) for all cases
results in a bivector. This can be derived from inspecting the product table for the
regressive geometric product. However, this time we only have a common factor of
$\frac{1}{2}$, and not $-\frac{1}{2}$ as in $EGA$ as required argument of the exponential
function. This will impact a sign for the transformation in the regressive case for $PGA$
compared to the corresponding transformations in $EGA$.  The resulting transformation is
then obtained from
\begin{equation}
    X = M \veedot X_0 \veedot \rrev{M} \quad (\text{for $PGA$})
    \label{eq:pga_sandwich_tranformation}
\end{equation} 

We can think of the screw motion used to model motion in $3D$ $PGA$ as a combination of a
rotation by an angle $\phi$ around the unitized line $\bm{l}$ and by a translation from a
displacement $\delta$ along that same line $\bm{l}$. The latter part results from a
rotation around a line at infinity given by $\rightweightdual{\bm{l}}$. Thus, the bivector
representing the line $\bm{l}$ can always be split into the parts creating the rotation
(movement around that line) and the parts creating the translation (movement along that
line).\\


\newpage

\subsubsection{Motion in $EGA$}
\label{subsection:motion_EGA}

In order to dive deeper into modelling of motions we first start with the $EGA$ case. It
is identical to the derivation shown in
\cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023} for their plane-based approach.
This will help to better recognize the differences to the $PGA$ approach used here. \\

Before going into more detail, we need to prepare some formulas we later need for our
derivations: \\

For $EGA$ the rotor will now be written as $M$ instead of $R$ to make it easy to compare
to our $PGA$-derivations using motors $M$ later on. $M$ as a rotor in $EGA$ is an
orthonormal transformation and thus satisfies
\begin{equation}
M \wedgedot \rev{M} = \rev{M} \wedgedot M = 1
\label{eq:motor_relation_gpr_neutral_element}
\end{equation}
(remember: $M^{-1} = \rev{M}$), and therefore
\begin{equation}
   \frac{d}{dt}(M\wedgedot\rev{M}) = \dot{M}\wedgedot\rev{M} + M\wedgedot\dot{\rev{M}} = 0
   \label{eq:dot_rev_m}
\end{equation}
It follows that $M\wedgedot\dot{\rev{M}} = -\dot{M}\wedgedot\rev{M}$, which results in
\begin{equation}
   \dot{\rev{M}} = -\rev{M}\wedgedot\dot{M}\wedgedot\rev{M}
   \label{eq:dot_rev_m_as_dot_m}
\end{equation}
Starting from $\rev{M} \wedgedot M = 1$ instead yields the same result for
$\dot{\rev{M}}$. \\

In order to model rigid body mechanics we consider two coordinate frames:
\begin{itemize}
   \item A \emph{world frame} (index $w$) that is normally chosen to be an inertial frame.
   \item A \emph{body frame} (index $b$) that is attached to a rigid body. Typically, at
   its center of gravity, but this is not strictly required. Rigid body's internal forces
   are in equilibrium and thus need not be considered explicitly. It is sufficient to
   describe the translational motion of the body's center of gravity and its rotational
   motion with respect to the center of gravity. The resulting motion of the body frame
   relative to the world frame then fully covers rigid body motion (kinematics and
   dynamics), which depends on distribution of the mass of the body as well as on external
   forces and moments acting on the body.
   \item To describe the motion we use a time dependent motor $M=M(t)$. At time $t=0$ we
   call it $M_0 = M(t=0)$. $M_0$ brings an element $X$ of the body to its position $X_0$
   at time $t=0$.
\end{itemize}

\textbf{\emph{World frame}}: From eqn.~(\ref{eq:rotor_ega}) we know the form of a general
rotor in $EGA$ and use the ansatz for a time dependent rotor $M(t)$ after the object
$X$ was brought into its initial position by $M_0$ as $X_0 = M_0 \wedgedot X \wedgedot
\rev{M}_0$. If there is no difference between world and body frames at $t=0$ then $M_0=1$:
\begin{equation}
   M(t) = \exp_\wedgedot(-\frac{1}{2}B_w(t)) \wedgedot M_0
   \label{eq:rotor_ega_world_frame}
\end{equation}
with $B_w(t)$ as a time-dependent bivector that models the rotational plane and the
turning angle. The derivative of $M$ with respect to time $t$ is
\begin{equation}
   \dot{M}(t) = -\frac{1}{2} \dot{B_w}(t) \wedgedot
                \exp_\wedgedot (-\frac{1}{2}B_w(t)) \wedgedot M_0
              = -\frac{1}{2} \dot{B_w}(t) \wedgedot M(t)   
\end{equation}
where we use that the exponential of the bivector and the bivector rate commute
\emph{(tbd: show why this is the case)}. Using $\Omega_w = \dot{B_w}(t)$ as the bivector
rate of change in the \emph{world frame}, we get
\begin{equation}
   \dot{M} = -\frac{1}{2} \Omega_w \wedgedot M 
   \label{eq:motor_rate_world_frame}   
\end{equation}
By multiplying this from the right with ``$\wedgedot\;\rev{M}$'' this can be transformed
to
\begin{equation}
   \Omega_w = -2 \dot{M} \wedgedot \rev{M}
   \label{eq:bivector_rate_world_frame}
\end{equation}

\textbf{\emph{Body frame}}: Here we first move the object in the \emph{body frame} and
then transform it into the in the world frame by the transformation $M_0$. In this case
eqn.~(\ref{eq:rotor_ega_world_frame}) becomes 
\begin{equation}
   M(t) = M_0 \wedgedot \exp_\wedgedot(-\frac{1}{2}B_b(t))
   \label{eq:rotor_ega_body_frame}
\end{equation}
which in turn results in
\begin{equation}
   \dot{M}(t) = -\frac{1}{2} M_0 \wedgedot
                \exp_\wedgedot (-\frac{1}{2}B_b(t)) \wedgedot \dot{B_b}(t) 
              = -\frac{1}{2} M(t)  \wedgedot \dot{B_b}(t) 
\end{equation}
while eqn.~(\ref{eq:motor_rate_world_frame}) for the bivector rate of change in the
\emph{body frame} becomes
\begin{equation}
   \dot{M} = -\frac{1}{2} M \wedgedot \Omega_b  
   \label{eq:motor_rate_body_frame}   
\end{equation}
using $\Omega_b = \dot{B_b}(t)$ as the bivector rate of change in the \emph{body frame}.
This can be transformed by multiplying it from the left by ``$\rev{M}\;\wedgedot$'' to
obtain
\begin{equation}
   \Omega_b = -2 \rev{M} \wedgedot \dot{M}
   \label{eq:bivector_rate_body_frame}
\end{equation}

From equations (\ref{eq:motor_rate_world_frame}) and (\ref{eq:motor_rate_body_frame}) we
can see how world and body frame rates are connected to each other: $\Omega_w \wedgedot M
= M \wedgedot \Omega_b$, which can be transformed into
\begin{subeqnarray}
   \Omega_w & = M \wedgedot \Omega_b \wedgedot \rev{M} \\
   \Omega_b & = \rev{M} \wedgedot \Omega_w \wedgedot M
\end{subeqnarray}

Now we derive the rate of change of a point $X(t)$, i.e. its speed $\dot{X}(t)$, in the
\emph{world frame}, first assuming that $X(t=0) = X_0 = const.$, i.e. as a constant
element that is transformed to become $X$ by a time dependent motor $M(t)$. During the
derivation we will use the associativity of the geometric product and the neutral element
1 of the geometric product. We can write
\begin{eqnarray*}
   \dot{X}(t) &=& \frac{d}{dt}X(t) =\frac{d}{dt}(M \wedgedot X_0 \wedgedot \rev{M}) \\
              &=& \dot{M} \wedgedot X_0 \wedgedot \rev{M} + M \wedgedot X_0 \wedgedot
                  \dot{\rev{M}} \\
              &\overset{\mathrm{(\ref{eq:dot_rev_m_as_dot_m})}}{=}& \dot{M} \wedgedot 1
                  \wedgedot X_0 \wedgedot \rev{M} -M \wedgedot X_0 \wedgedot
                  (\rev{M}\wedgedot\dot{M}\wedgedot\rev{M}) \\
              &\overset{\mathrm{(\ref{eq:motor_relation_gpr_neutral_element})}}{=}&
                  \dot{M} \wedgedot (\rev{M} \wedgedot M) \wedgedot X_0 \wedgedot \rev{M}
                  -(M \wedgedot X_0 \wedgedot \rev{M}) \wedgedot \dot{M} \wedgedot \rev{M}
                  \\
              &=& (\dot{M} \wedgedot \rev{M}) \wedgedot (M\wedgedot X_0 \wedgedot \rev{M})
                  -(M\wedgedot X_0 \wedgedot\rev{M}) \wedgedot (\dot{M} \wedgedot \rev{M})
\end{eqnarray*}
From eqn. (\ref{eq:bivector_rate_world_frame}) we can conclude that $-\frac{1}{2}\Omega_w
(t) = \dot{M} \wedgedot \rev{M}$, resulting in
\begin{eqnarray}
   \dot{X}(t) &=&(-\frac{1}{2}\Omega_w(t)) \wedgedot X(t) - X(t) \wedgedot
              (-\frac{1}{2}\Omega_w(t)) \nonumber \\
              &=& -\frac{1}{2}\left( \Omega_w \wedgedot X - X \wedgedot \Omega_w\right)
              \nonumber \\
              &=& -\mathrm{cmt}(\Omega_w,X) \nonumber \\[10pt]
   \dot{X}_w    &=& \mathrm{cmt}(X_w,\Omega_w)
   \slabel{eq:cmt_ega_world} 
\end{eqnarray}
with $\mathrm{cmt}(X_w,\Omega_w)$ as the commutator product between $X_w$ and $\Omega_w$.
\\

So, if the bivector rate of change $\Omega_w$ is known, the rate of change of $X_w$, which
is $\dot{X}_w$, can be calculated from eqn.~(\ref{eq:cmt_ega_world}) using the commutator
product. \\

It can be shown that $\dot{M} \wedgedot \rev{M}$ is always a bivector: From equations
(\ref{eq:bivector_rate_world_frame}) and (\ref{eq:dot_rev_m}), and the linearity of the
involved operations, we get
\begin{equation}
   \Omega_w = -2 \dot{M} \wedgedot \rev{M} = 2 M \wedgedot
   \dot{\rev{M}} = 2 (\rev{\dot{M}\wedgedot\rev{M}})= -\rev{\Omega}_w
   \label{eq:bivector_poof_gpr}
\end{equation}
Considering the grade-dependent sign-reversal of the reverse in $3D$ $EGA$, $\Omega_w$
must either be a bivector or a trivector. Since $\dot{M}$ and $\rev{M}$ are even-grade
multivectors, their geometric product results in an even-grade multivector as well, which
does not have a trivector part. Thus, $\Omega_w$ and therefore $\dot{M} \wedgedot \rev{M}$
must indeed be a bivector. \\

This also assures, that the grade of $X$ and $\dot{X}$ are identical. If $X$ is a vector
then $\dot{X}$ is a vector as well. This statement is true for any grade of $X$ as long as
$\Omega_w$ is a bivector. \\

If $X_0$ is not a constant element, but a function of time, which means that $X_0$ is a
time dependent element in the \emph{body frame} $X_0 = X_b(t)$, we have to add another
term to the derivative in eqn.~(\ref{eq:cmt_ega_world}) as follows:
\begin{eqnarray}
   \dot{X}(t) &=& \frac{d}{dt}(M \wedgedot X_b \wedgedot \rev{M}) \nonumber \\
              &=& \dot{M} \wedgedot X_b \wedgedot \rev{M} + M \wedgedot X_b \wedgedot
                  \dot{\rev{M}} + M \wedgedot \dot{X_b} \wedgedot \rev{M} \nonumber \\
              &=& \mathrm{cmt}(M\wedgedot X_b \wedgedot\rev{M},\Omega_w)
                   + M \wedgedot \dot{X_b} \wedgedot \rev{M} \nonumber \\[10 pt]
   \dot{X}_w  &=& \mathrm{cmt}(X_w,\Omega_w) + (\dot{X_b})_w
   \slabel{eq:cmt_ega_world_incl_X0_of_t}
\end{eqnarray}
Eqn.~(\ref{eq:cmt_ega_world_incl_X0_of_t}) is to be considered in the world frame.
However, we can also do our derivations in a completely equivalent way using the
corresponding bivector rates and rotors of the \emph{body frame}. \\

Now we derive the rate of change of a point $X(t)$ in the \emph{body frame}, again
assuming that $X(t=0) = X_0 = const.$ We can write
\begin{eqnarray*}
   \dot{X}(t) &=& \dot{M} \wedgedot X_0 \wedgedot \rev{M} + M \wedgedot X_0 \wedgedot
                  \dot{\rev{M}} \\
              &\overset{\mathrm{(\ref{eq:dot_rev_m_as_dot_m})}}{=}& 1 \wedgedot \dot{M}
                  \wedgedot 1 \wedgedot X_0 \wedgedot \rev{M} -M \wedgedot X_0 \wedgedot
                  (\rev{M}\wedgedot\dot{M}\wedgedot\rev{M}) \\
              &\overset{\mathrm{(\ref{eq:motor_relation_gpr_neutral_element})}}{=}& M
                  \wedgedot (\rev{M} \wedgedot \dot{M}) \wedgedot \rev{M} \wedgedot (M
                  \wedgedot X_0 \wedgedot \rev{M}) -(M \wedgedot X_0 \wedgedot \rev{M})
                  \wedgedot 1 \wedgedot \dot{M} \wedgedot \rev{M} \\
              &=& \left( M \wedgedot (\rev{M} \wedgedot \dot{M}) \wedgedot \rev{M} \right)
                  \wedgedot X(t) - X(t) \wedgedot \left( M \wedgedot (\rev{M} \wedgedot
                  \dot{M}) \wedgedot \rev{M} \right) \\
              &=& \left( M \wedgedot (\rev{M} \wedgedot \dot{M}) \wedgedot \rev{M} \right)
                  \wedgedot X(t) - X(t) \wedgedot \left( M \wedgedot (\rev{M} \wedgedot
                  \dot{M}) \wedgedot \rev{M} \right)
\end{eqnarray*}
From eqn. (\ref{eq:bivector_rate_body_frame}) we can conclude that $-\frac{1}{2}\Omega_b
(t) = \rev{M} \wedgedot \dot{M}$, resulting in
\begin{eqnarray}
   \dot{X}(t) &=&\left( M \wedgedot (-\frac{1}{2}\Omega_b(t)) \wedgedot \rev{M} \right)
                 \wedgedot X(t) - X(t) \wedgedot \left( M \wedgedot
                 (-\frac{1}{2}\Omega_b(t)) \wedgedot \rev{M} \right) \nonumber \\
              &=& -\frac{1}{2}\left( (\Omega_b)_w \wedgedot X - X \wedgedot
              (\Omega_b)_w\right) \nonumber \\
              &=& -\frac{1}{2}\left( \Omega_w \wedgedot X - X \wedgedot \Omega_w\right)
              \nonumber \\
              &=& -\mathrm{cmt}(\Omega_w,X) \nonumber \\[10pt]
   \dot{X}_w  &=& \mathrm{cmt}((X_b)_w,\Omega_w) = \mathrm{cmt}(X_w,\Omega_w)
   \slabel{eq:cmt_ega_body} 
\end{eqnarray}
with $\mathrm{cmt}(X_w,\Omega_w)$ as the commutator product between $X_w$ and $\Omega_w$.
This is exactly the same result as before, just based on bivector rate measured in the
\emph{body frame}. \\

It can be shown that $\rev{M} \wedgedot \dot{M}$ is always a bivector with the same
argument as before. So we don't repeat that here. \\

If $X_0$ is not a constant element, but itself a function of time $X_b(t)$, we have to add
another term to the derivative in eqn.~(\ref{eq:cmt_ega_body}) again as follows:
\begin{eqnarray}
   \dot{X}(t) &=& \frac{d}{dt}(M \wedgedot X_b \wedgedot \rev{M}) \nonumber \\
              &=& \dot{M} \wedgedot X_b \wedgedot \rev{M} + M \wedgedot X_b \wedgedot
                  \dot{\rev{M}} + M \wedgedot \dot{X_b} \wedgedot \rev{M} \nonumber \\
              &=& \mathrm{cmt}(M\wedgedot X_b \wedgedot\rev{M},\Omega_w)
                  + M \wedgedot \dot{X_b} \wedgedot \rev{M} \nonumber \\
              &=& \mathrm{cmt}((X_b)_w,\Omega_w) + (\dot{X_b})_w \nonumber \\[10 pt]
   \dot{X}_w  &=& \mathrm{cmt}(X_w,\Omega_w) + (\dot{X_b})_w
   \slabel{eq:cmt_ega_body_incl_X0_of_t}
\end{eqnarray}
Eqn.~(\ref{eq:cmt_ega_body_incl_X0_of_t}) is again to be considered in the \emph{world
frame}. The result derived in the \emph{body frame} is of course identical to the result
derived in the \emph{world frame}. That means most of the equations can be used simply by 
replacing $Omega_w$ by $Omega_b$. \\

In $EGA$ rotation as a Euclidean isometry can be described with sandwich products based on
the geometric product. The corresponding sandwich product is available in the library
either by using the geometric product and the grade operator to extract the grade of the
results directly, or by using the function \path{rotate(X, R)} which takes the object $X$
and the rotor $R$ as arguments and returns the rotated object.\\


\newpage

\subsubsection{Motion in $PGA$}
\label{subsection:motion_PGA}

Now we are looking into $PGA$. Lengyel has shown in \cite{Lengyel_pga-illuminated:2024}
that Euclidean isometries (covering translations as well as rotations) can be described
with sandwich products based on the regressive geometric product in $PGA$. If we do that,
we can stay in the space $\mathbb{G}(3,0,1)$ and do not need to switch to its dual space
in order to implement these orthonormal transformations as already described above. \\

Here we repeat the derivation of the time-derivative $\dot{X}$ of a projective point
$X(t)$ for the approach of $PGA$ used here: We will use the regressive geometric product,
and $\ps$ as its neutral element. The derivation is then completely analogous to the $EGA$
case with the given substitutions. \\

$M$ as a motor in $PGA$ is an orthonormal transformation and satisfies
\begin{equation}
M \veedot \rrev{M} = \rrev{M} \veedot M = \ps
\label{eq:motor_relation_rgpr_neutral_element}
\end{equation}
as the defining equation for a motor in $PGA$ (remember: $M^{-1} = \rrev{M}$), leads to 
\begin{equation}
   \frac{d}{dt}(M\veedot\rrev{M}) = \dot{M}\veedot\rrev{M} + M\veedot\dot{\rrev{M}} = 0
   \label{eq:dot_rrev_m}
\end{equation}
It follows that $M\veedot\dot{\rrev{M}} = -\dot{M}\veedot\rrev{M}$, which results in
\begin{equation}
   \dot{\rrev{M}} = -\rrev{M}\veedot\dot{M}\veedot\rrev{M}
   \label{eq:dot_rrev_m_as_dot_m}
\end{equation}
Starting from $\rrev{M} \veedot M = \ps$ instead yields the same result for
$\dot{\rrev{M}}$. \\

However, we still have to take additional aspects into account: \\

First, Lengyel has shown that a motor in $3D$ $PGA$ can be created from the exponential in
eqn.~(\ref{eq:rexp_pga_motor}) as $M = \exp_\veedot\left[(\frac{\delta}{2} \bm{1} +
\frac{\phi}{2}\ps) \veedot \bm{l} \right]$ where it is easy to show that the regressive
geometric product of the line $\bm{l}$, a bivector, with the scalar part yields a bivector
and the same is true for the regressive geometric product between the pseudoscalar and the
line. Thus, it is sufficient to consider the exponential function of a bivector argument
for the derivation for $3D$ $PGA$. The resulting motor $M$ will be an even-grade
multivector in this case. \\

Second, for $2D$ $PGA$ the motor is derived from $M = \exp_\veedot\left[\frac{\phi}{2} P
\right]$ with the unitized point $P$, a vector, as an argument. The resulting motor $M$
will be an odd-grade multivector. We need to distinguish these two cases later on.
However, since the equations are structurally similar (an exponential function with either
a vector or a bivector as argument), we only do the derivation for the $3D$ case once and
then will interpret the results for the $2D$ case separately afterwards. \\

\textbf{\emph{World frame}}: From eqn.~(\ref{eq:rexp_pga_motor}) we know the form of a
general motor in $PGA$ and use the ansatz for a time dependent motor $M(t)$ after the
object $X$ was brought into its initial position by $M_0$ as $X_0 = M \veedot X \veedot
\rrev{M}$. If there is no difference between world and body frames at $t=0$ then
$M_0=\ps$:
\begin{equation}
   M(t) = \exp_\veedot(\frac{1}{2}B_w(t)) \veedot M_0
   \label{eq:motor_pga_world_frame}
\end{equation}
with $B_w(t)$ as a time-dependent bivector that models the rotational plane and the
turning angle. The derivative of $M$ with respect to time $t$ is
\begin{equation}
   \dot{M}(t) = \frac{1}{2} \dot{B_w}(t) \veedot
                \exp_\veedot (\frac{1}{2}B_w(t)) \veedot M_0
              = \frac{1}{2} \dot{B_w}(t) \veedot M(t)   
\end{equation}
where we use that the exponential of the bivector and the bivector rate commute
\emph{(tbd: show why this is the case)}. Using $\Omega_w = \Omega_w(t) = \dot{B_w}(t)$ as
the bivector rate of change in the \emph{world frame}, we get
\begin{equation}
   \dot{M} = \frac{1}{2} \Omega_w \veedot M 
   \label{eq:motor_rate_world_frame_pga}   
\end{equation}
By multiplying this from the right with ``$\veedot\;\rrev{M}$'' this can be transformed to
\begin{equation}
   \Omega_w = 2 \dot{M} \veedot \rrev{M}
   \label{eq:bivector_rate_world_frame_pga}
\end{equation}

\textbf{\emph{Body frame}}: Here we first move the object in the \emph{body frame} and
then transform it into the in the world frame by the transformation $M_0$. In this case
eqn.~(\ref{eq:motor_pga_world_frame}) becomes 
\begin{equation}
   M(t) = M_0 \veedot  \exp_\veedot(\frac{1}{2}B_b(t))
   \label{eq:motor_pga_body_frame}
\end{equation}
which in turn results in
\begin{equation}
   \dot{M}_b(t) = \frac{1}{2} M_0 \veedot
                  \exp_\veedot (\frac{1}{2}B_b(t)) \veedot \dot{B_b}(t) 
                = \frac{1}{2} M(t) \veedot \dot{B_b}(t)
   \label{eq:motor_rate_pga_body_frame}
\end{equation}
while eqn.~(\ref{eq:motor_rate_world_frame_pga}) for the bivector rate of change in the
\emph{body frame} becomes
\begin{equation}
   \dot{M} = \frac{1}{2} M \veedot \Omega_b  
   \label{eq:motor_rate_body_frame_pga}   
\end{equation}
using $\Omega_b = \Omega_b(t) = \dot{B_b}(t)$ as the bivector rate of change in the
\emph{body frame}. This can be transformed by multiplying it from the left by
``$\rrev{M}\;\veedot$'' to obtain
\begin{equation}
   \Omega_b = 2 \rrev{M} \veedot \dot{M}
   \label{eq:bivector_rate_body_frame_pga}
\end{equation}

From equations (\ref{eq:motor_rate_world_frame_pga}) and
(\ref{eq:motor_rate_body_frame_pga}) we can see how world and body frame rates are
connected to each other: $\Omega_w \veedot M = M \veedot \Omega_b$, which can be
transformed into
\begin{subeqnarray}
   \Omega_w & = M \veedot \Omega_b \veedot \rrev{M}
   \slabel{eq:body_to_world_pga} \\
   \Omega_b & = \rrev{M} \veedot \Omega_w \veedot M
   \slabel{eq:world_to_body_pga}
\end{subeqnarray}

Now we derive the rate of change of a point $X(t)$, i.e. its speed $\dot{X}(t)$, in the
\emph{world frame}, first assuming that $X(t=0) = X_0 = const.$, i.e. as a constant
element that is transformed by a time dependent motor $M(t)$. During the derivation we
will use the associativity of the regressive geometric product and the neutral element
$\ps$ of the regressive geometric product. We can write
\begin{eqnarray*}
   \dot{X}(t) &=& \frac{d}{dt}X(t) =\frac{d}{dt}(M \veedot X_0 \veedot \rrev{M}) \\
              &=& \dot{M} \veedot X_0 \veedot \rrev{M} + M \veedot X_0 \veedot
                  \dot{\rrev{M}} \\
              &\overset{\mathrm{(\ref{eq:dot_rrev_m_as_dot_m})}}{=}& \dot{M} \veedot \ps
                  \veedot X_0 \veedot \rrev{M} -M \veedot X_0 \veedot
                  (\rrev{M}\veedot\dot{M}\veedot\rrev{M}) \\
              &\overset{\mathrm{(\ref{eq:motor_relation_rgpr_neutral_element})}}{=}&
                  \dot{M} \veedot (\rrev{M} \veedot M) \veedot X_0 \veedot \rrev{M} -(M
                  \veedot X_0 \veedot \rrev{M}) \veedot \dot{M} \veedot \rrev{M} \\
              &=& (\dot{M} \veedot \rrev{M}) \veedot (M\veedot X_0 \veedot \rrev{M})
                  -(M\veedot X_0 \veedot\rrev{M}) \veedot (\dot{M} \veedot \rrev{M})
\end{eqnarray*}
From eqn. (\ref{eq:bivector_rate_world_frame_pga}) we can conclude that
$\frac{1}{2}\Omega_w (t) = \dot{M} \veedot \rrev{M}$, resulting in
\begin{eqnarray}
   \dot{X}(t) &=&(\frac{1}{2}\Omega_w(t)) \veedot X_w(t) - X_w(t) \veedot
              (\frac{1}{2}\Omega_w(t)) \nonumber \\
              &=& \frac{1}{2}\left( \Omega_w \veedot X_w - X_w \veedot \Omega_w\right)
              \nonumber \\[10pt]
   \dot{X}_w  &=&  \mathrm{rcmt}(\Omega_w,X_w) = -\mathrm{rcmt}(X_w,\Omega_w)
   \slabel{eq:rcmt_pga_world} 
\end{eqnarray}
with $\mathrm{rcmt}(\Omega_w, X_w)$ as the regressive commutator product between
$\Omega_w$ and $X_w$. The resulting minus sign in eqn.~(\ref{eq:rcmt_pga_world}) compared
to eqn.~(\ref{eq:cmt_ega_world}) is the only difference between $PGA$ and $EGA$. This is
caused by the different ansatz for the motors comparing equations (\ref{eq:rotor_ega}) and
(\ref{eq:rexp_pga_motor}).\\

It can be shown that $\dot{M} \veedot \rrev{M}$ is either always vector or a bivector:
From equations (\ref{eq:bivector_rate_world_frame_pga}) and (\ref{eq:dot_rrev_m}), and the
linearity of the involved operations, we get
\begin{equation}
   \Omega_w = 2 \dot{M} \veedot \rrev{M} = - 2 M \veedot
   \dot{\rrev{M}} = - 2 (\rrev{\dot{M} \veedot \rrev{M} })= -\rrev{\Omega}_w
   \label{eq:bivector_poof_rgpr}
\end{equation}
Considering the grade-dependent sign-reversal of the regressive reverse in $3D$ $PGA$,
$\Omega_w$ must either be a vector or a bivector (see table~\ref{tab:pga3dp_rev_rrev} at
the end of section~\ref{subsection:basics}). Since $\dot{M}$ and $\rrev{M}$ are even-grade
multivectors, their regressive geometric product results in an even-grade multivector as
well, which does not have a vector part. Thus, $\Omega_w$ and $\dot{M} \veedot \rrev{M}$
must indeed be a bivector in $3D$ $PGA$. \\

However, in $2D$ $PGA$ due to the grade-dependent sign-reversal of the regressive reverse,
$\Omega_w$ must either be a scalar or a vector (see table~\ref{tab:pga2dp_rev_rrev} at the
end of section~\ref{subsection:basics}). Since $\dot{M}$ and $\rrev{M}$ are odd-grade
multivectors, their regressive geometric product results in an odd-grade multivector as
well, which does not have a scalar part. Thus, $\Omega_w$ and $\dot{M} \veedot \rrev{M}$
must indeed be a vector in $2D$ $PGA$. \\

If $X_0$ is not a constant element, but a function of time, which means that $X_0$ is a
time dependent element in the \emph{body frame} $X_0 = X_b(t)$, we have to add another
term to the derivative in eqn.~(\ref{eq:rcmt_pga_world}) as follows:
\begin{eqnarray}
   \dot{X}(t) &=& \frac{d}{dt}(M \veedot X_b \veedot \rrev{M}) \nonumber \\
              &=& \dot{M} \veedot X_b \veedot \rrev{M} + M \veedot X_b \veedot
                  \dot{\rrev{M}} + M \veedot \dot{X_b} \veedot \rrev{M} \nonumber \\
              &=& -\mathrm{rcmt}(M\veedot X_b \veedot\rrev{M},\Omega_w)
                  + M \veedot \dot{X_b} \veedot \rrev{M} \nonumber \\[10 pt]
   \dot{X}_w  &=& -\mathrm{rcmt}(X_w,\Omega_w) + (\dot{X_b})_w
               = \mathrm{rcmt}(\Omega_w,X_w)+ (\dot{X_b})_w
   \slabel{eq:rcmt_pga_world_incl_X0_of_t}
\end{eqnarray}
Eqn.~(\ref{eq:rcmt_pga_world_incl_X0_of_t}) is to be considered in the world frame.
However, we can also do our derivations in a completely equivalent way using the
corresponding bivector rates and rotors of the \emph{body frame}. \\

The derivation in the \emph{body frame} does not bring new insights, so we don't reproduce
that here for $PGA$ again. \\

\newpage

\subsubsection{Motor generators in $PGA$}
\label{subsection:motor_generators_PGA}

The final step before we move on to modelling physics is to show how points are moved in
$PGA$ using the approach described above. \\

In $3D$ $PGA$ every bivector $B = B(s, \phi)$ or every bivector rate $\dot{B} = \Omega =
\Omega(v, \omega)$ has its basis components
\begin{equation}
   \Omega_{3D} =
   \underbrace{\Omega_{vx} \bv{41} + \Omega_{vy} \bv{42} +
               \Omega_{vz} \bv{43}}_{\text{weight($\Omega$)}
   =\weight{\Omega}}
   + 
    \underbrace{\Omega_{mx} \bv{23} + \Omega_{my}\bv{31} +
    \Omega_{mz}\bv{12}}_{\text{bulk($\Omega$)}
   =\bulk{\Omega}}
   \label{eq:rate_bivector_component_definition}
\end{equation}
where the \emph{weight $\weight{\Omega}$ carries rotational information} and the
\emph{bulk $\bulk{\Omega}$ carries translational information}. \\

In general, the weight only contains components that \emph{do} contain a factor of
$\bv{4}$. It has \emph{directional} information on $\Omega$. If there is zero weight, the
object is contained in the horizon ($H_{3dp}$). The bulk only has components that \emph{do
not} contain a factor of $\bv{4}$. It has \emph{positional} information on $\Omega$. If
there is zero bulk, the object contains the origin ($O_{3dp}$). \\

We can either model a pure translation, which will only be visible in the bulk part of
$\Omega$, with the weight part being zero in that case, or we can model a pure rotation
around a fixed (non-ideal) line $\bm{l}$. In that case the positional information of the
fixed line of rotation becomes encoded in the bulk part of $\Omega$ and the directional
information on the rotation speed vector $\omega$ is encoded in the weight part of
$\Omega$.\\
 
A mixture of a translation and a rotation would either be equivalent to a rotation around
another fixed line shifted by the translational part or, in case of translation along the
axis of rotation, a screw motion around and along that axis. It is sufficient to look at
pure translation and pure rotation around a fixed line as separate cases. This separation
is also important for defining the exponential function with respect to the regressive
geometric product that converts the bivector $B$ to the respective motor $M$: $M(t) =
\exp_{\wedgedot}(\frac{1}{2} B(t)) = \exp_{\wedgedot}(\frac{1}{2} \Omega t)$. \\

The total bivector rate $\Omega$ can be split into
\begin{equation}
   \Omega = \weight{\Omega} + \bulk{\Omega} = \Omega_{rot} + \Omega_{tra}
   \label{eq:rate_bivector_euclidean_split}
\end{equation}
where eqn.~(\ref{eq:rate_bivector_euclidean_split}) is called the ``Euclidean split'' of
the bivector rate $\Omega$ because it separates the rotational and translational aspects
of a moving object. \\

A pure rotation around an axis $\vec{\omega} = \omega$ can be modelled by a fixed
(unitized) line that is scaled by the angular speed of rotation, i.e. by the bulk norm
$\bulk{\nrm{\omega}}$. If we have an arbitrary rotation axis we can either wedge a
unitized point $Q = O_{3dp} + q = \bv{4} + q$ that is contained in the rotational axis
with the rotational speed $\omega$, or create a unitized line $\hat{\omega}$ in the
direction of the rotation vector $\omega$ that is later scaled with the rotational speed
such that the bivector $\Omega_{rot}$ becomes
\begin{eqnarray}
   \Omega_{rot} &=& Q \wedge \omega = (O_{3dp} + q) \wedge \omega \nonumber \\
                &=& \bv{4} \wedge \omega + q \wedge \omega \nonumber \\
                &=& \bulknrm{\omega}(\bv{4} \wedge \hat{\omega} + q \wedge \hat{\omega})
                \nonumber \\ 
   \Omega_{rot} &=&
   \bulknrm{\omega}
   \begin{pmatrix}
      \hat{\omega}_x \\
      \hat{\omega}_y \\
      \hat{\omega}_z \\
      q_y \hat{\omega}_z - q_z \hat{\omega}_y \\
      q_z \hat{\omega}_x - q_x \hat{\omega}_z \\
      q_x \hat{\omega}_y - q_y \hat{\omega}_x
   \end{pmatrix}
   \label{eq:bivector_rot_rate_omega_vec}
\end{eqnarray}
where $\hat{\omega}$ is the bulk normalized rotation vector
$\hat{\omega}=\omega/\bulknrm{\omega}$. Thus, $\Omega_{rot}$ directly represents the
normalized (non-ideal) axis of rotation. It is a line of fixed points not changing due to
the rotation, scaled with the angular speed of rotation $\bulknrm{\omega}$. \\

Of course, $\Omega_{rot}$ could also be expressed directly by the plane of rotation $pl$
(a trivector) when the magnitude of the plane encodes the angular speed
$\bulknrm{\omega}$. Considering that the left weight dual of the plane
$\leftweightdual{pl}$ provides the normal vector of the plane including its magnitude,
eqn.~(\ref{eq:bivector_rot_rate_omega_vec}) can be written in the equivalent form directly
in terms of the plane of rotation $pl$
\begin{equation}
   \Omega_{rot} = Q \wedge \leftweightdual{pl}
   \label{eq:bivector_rot_rate_omega_pl_rot}
\end{equation}

For modelling translational speed $\vec{v} = v$ in $3D$ we need a bivector rate
$\Omega_{tra}$ that does not contain a factor of $\bv{4}$. This can be achieved by
encoding the speed of translation in the normal vector $n_v$ of a plane and then
intersecting this plane with the horizon $\rcmpl{\bv{4}} = \rcmpl{O_{3dp}} = H_{3dp}$. The
plane with the corresponding attitude can be encoded by using the right bulk dual
$\rightbulkdual{v}$ to model the plane (a trivector) with the bulk norm $\bulknrm{v}$ as
magnitude. It contains the origin $O_{3dp}$ and results in a bivector rate $\Omega_{tra}$
of the translation as
\begin{equation}
   \Omega_{tra} = \rightbulkdual{v} \vee \rcmpl{\bv{4}} = \rightbulkdual{v} \vee H_{3dp}
                = att(\rightbulkdual{v})
                = (0, 0, 0, v_x, v_y, v_z)
   \label{eq:bivector_trans_rate_v_vec}
\end{equation}

For the case of screw motion we have the special case that the translational motion is
directed in the same direction as the rotation axis, i.e. we have a special $\Omega_{tra}$
\begin{equation}
   \Omega_{tra, screw}
   = \frac{\delta}{\bulknrm{\omega}} att(\rightbulkdual{\omega})
   = \delta \, (0, 0, 0, \hat{\omega}_x, \hat{\omega}_y, \hat{\omega}_z) 
\end{equation}
where $\hat{\omega}$ is the bulk normalized vector $\omega$, which has magnitude one. This
vector is scaled with the translation distance $\delta$ resulting in $\Omega_{screw}$ of
\begin{equation}
   \Omega_{screw} = \Omega_{rot} + \Omega_{tra, screw} =
   \bulknrm{\omega}
   \begin{pmatrix}
      \hat{\omega}_x \\
      \hat{\omega}_y \\
      \hat{\omega}_z \\
      q_y \hat{\omega}_z - q_z \hat{\omega}_y\\
      q_z \hat{\omega}_x - q_x \hat{\omega}_z\\
      q_x \hat{\omega}_y - q_y \hat{\omega}_x
   \end{pmatrix}
   + \delta
   \begin{pmatrix}
      0 \\
      0 \\
      0 \\
      \hat{\omega}_x\\
      \hat{\omega}_y\\
      \hat{\omega}_z
   \end{pmatrix}
\end{equation}
The components of $\Omega_{rot}$, $\Omega_{tra}$ and $\Omega_{screw}$ are relevant for the
exponential function that transforms our bivector, representing the intended motion, into
a motor that transforms points accordingly. \\

Using for example eqn.~(\ref{eq:rcmt_pga_world}) and the Euclidean split as defined in
eqn.~(\ref{eq:rate_bivector_euclidean_split}) we can write
\begin{subeqnarray}
   \dot{X}_w &=& \mathrm{rcmt}(\Omega_w, X_w) \;\; = \;\; \dot{x}_w \nonumber \\
             &=& \mathrm{rcmt}(\Omega_{rot} + \Omega_{tra}, X_w) \nonumber \\
   \dot{x}_w &=& \mathrm{rcmt}(\Omega_{rot}, X_w) + \mathrm{rcmt}(\Omega_{tra}, X_w) \\
   \dot{x}_w &=& v_{rot} + v_{tra}  
   \label{eq:speed_from_rate}
\end{subeqnarray}
to calculate the combined speed $\dot{x}_w$ from rotational and translational effects at a
position $X_w$. \\

\newpage

In $2D$ $PGA$ the equations are very similar, but the rate of change is not a rate of
change bivector, but is it a rate of change vector, since the rotation is not around a
fixed axis, as in $3D$, but it is around a fixed point. The translation is characterized
by a $2D$ vector, whereas the rotation rate can be characterized by a single scalar value
and a fixed point $Q$ around with the rotation occurs. To show that we still get
comparable equations to the $3D$ case, we will also use $\Omega = \Omega(v, \omega)$ for
the rate of change. In $2D$ $PGA$ has its basis components
\begin{equation}
   \Omega_{2D} = \underbrace{\Omega_x \bv{1} \quad
               + \quad \Omega_y \bv{2}}_{\text{bulk($\Omega$)}
               =  \bulk{\Omega}} \quad
               + \underbrace{\Omega_z \bv{3}}_{\text{weight($\Omega$)}
               = \weight{\Omega}}
   \label{eq:rate_vector_component_definition}
\end{equation}
where the \emph{weight $\weight{\Omega}$ carries rotational information} and the
\emph{bulk $\bulk{\Omega}$ carries translational information}. In general, the weight only
contains components that \emph{do} contain a factor of $\bv{3}$ (which here takes a role
equivalent to $\bv{4}$ in the $3D$ case). The bulk only has components that \emph{do not}
contain a factor of $\bv{3}$.\\

The total vector rate of change $\Omega$ can be split into
\begin{equation}
   \Omega = \bulk{\Omega} + \weight{\Omega} = \Omega_{tra} + \Omega_{rot}
   \label{eq:rate_vector_euclidean_split}
\end{equation}
where eqn.~(\ref{eq:rate_vector_euclidean_split}) is called the ``Euclidean split'' of the
vector rate of change $\Omega$ separating the translational and rotational aspects of a
moving object. \\

A simple rotation in $2D$ with an angular speed $\omega$ (a simple scalar in this case)
around the unitized point $Q = O_{2dp} + q = \bv{3} + q$ can be modelled by multiplying
the point $Q$ with $\omega$ using the wedge product to generate the vector $\Omega_{rot}$.
\begin{equation}
   \Omega_{rot} = Q \wedge \omega = \bv{3} \wedge \omega + q \wedge \omega
                = (0, 0, 1)\,\omega  + (q_x, q_y, 0)\,\omega 
                = (q_x, q_y, 1)\,\omega
   \label{eq:vector_rot_rate_omega}
\end{equation}

For modelling translational speed $\vec{v} = v$ in $2D$ we need a vector rate of change
$\Omega_{tra}$ that does not contain a factor of $\bv{3}$. To show this we use exactly the
same equations as in the $3D$ case by encoding the speed of translation in the normal
vector $n_v$ of a line and then intersecting this line with the horizon $\rcmpl{\bv{3}} =
\rcmpl{O_{2dp}} = H_{2dp}$. The double reflection on parallel lines that effectively
creates the translation, necessitates the need for encoding the translation speed in a
vector pointing normal to the translation direction. The line with the corresponding
attitude can be encoded by using the right bulk dual $\rightbulkdual{v}$ to model the line
(a bivector) with the bulk norm $\bulknrm{v}$ as magnitude. The vector rate of change
$\Omega_{tra}$ of a pure translation is then given as
\begin{equation}
   \Omega_{tra} = \rightbulkdual{v} \vee \rcmpl{\bv{3}} = \rightbulkdual{v} \vee H_{3dp}
                = att(\rightbulkdual{v})
                = (-v_y, v_x, 0)
   \label{eq:vector_trans_rate_v_vec}
\end{equation}
where the direction of the movement is encoded as the normal pointing towards the point at
infinity which is the``rotation center at infinity'' of the modelled translation. \\

We can either model a pure translation, which will only be visible in the bulk part of
$\Omega$, with the weight part being zero in that case, or we can model a pure rotation
around a fixed (non-ideal) point $Q$. In that case the positional information of the fixed
point becomes encoded in the bulk part of $\Omega$ and the rotational speed $\omega$ is
encoded in the weight part of $\Omega$. A mixture of a translation and a rotation would be
equivalent to a rotation around another fixed point shifted by the translational part. So
it is sufficient to look at pure translation and pure rotation around a fixed point as
separate cases.
\begin{equation}
   \Omega_{2D} = \Omega_{rot} + \Omega_{tra} =
   \omega
   \begin{pmatrix}
      q_x \\
      q_y \\
      1
   \end{pmatrix}
   +
   \begin{pmatrix}
      -v_y \\
      v_x \\
      0
   \end{pmatrix}
\end{equation}

This separation is also important for defining the exponential function with respect to
the regressive geometric product that converts the bivector $B$ to the respective motor
$M$: $M(t) = \exp_{\veedot}(\frac{1}{2} B(t)) = \exp_{\veedot}(\frac{1}{2} \Omega t)$ that
enables us to actually transform points according to the vector rates of change as defined
above.


\newpage

\subsubsection{Independent generators, coupled finite displacements}
\label{subsubsection_lie_algebra_vs_lie_group}

The motion generator $\Omega_b$ lives in the \emph{Lie algebra} $\mathfrak{se}(2)$ ($2D$)
or $\mathfrak{se}(3)$ ($3D$). The motor $M$ itself lives in the corresponding \emph{Lie
group} $SE(2)$ or $SE(3)$. These two levels have fundamentally different structure. The
Lie algebra $\mathfrak{se}(2)$ is three-dimensional with basis generators:

\begin{table}[!ht]
\centering\small
\begin{tabular}{l|l|l}
\hline
Generator & \texttt{vec2dp} encoding & Effect \\
\hline
Translation in $x$ & $\{0,\;1,\;0\}$ & body drifts in $+x$ \\
Translation in $y$ & $\{-1,\;0,\;0\}$ & body drifts in $+y$ \\
Rotation about origin & $\{0,\;0,\;1\}$ & body spins about world origin \\
\hline
\end{tabular}
\caption{Basis generators of $\mathfrak{se}(2)$ in $PGA2DP$ encoding.}
\label{tab:lie_generators_2dp}
\end{table}

A general instantaneous velocity $\Omega_b = \{-v_y,\; v_x,\; \omega\}$ means
\emph{simultaneously}: translate at $(v_x, v_y)$ \emph{and} rotate at $\omega$ -- with
\emph{no coupling} between the three components. Integrating the velocity over an
infinitesimal step $dt$ gives
\begin{equation}
    B_b \leftarrow B_b + \Omega_b \cdot dt,
    \label{eq:lie_algebra_accumulation}
\end{equation}
so all three components of $B_b$ accumulate independently. \\

When computing a \emph{finite} displacement via $M = \exp_{\veedot}(\tfrac{1}{2} B_b)$,
the structure of the $2D$ plane imposes a hard constraint:

\begin{quote}
\textbf{Chasles' theorem ($2D$):} Any rigid body displacement in $2D$ is either a pure
rotation about some center, or a pure translation. There is no screw motion.
\end{quote}

Consequently, for any $B_b$ with $B_b.z \neq 0$, the motor $\exp_{\veedot}(\tfrac{1}{2}
B_b)$ is \emph{always a pure rotation} -- the translational components merely determine
the rotation center:
\begin{equation}
    \text{pivot} = \left(\frac{B_b.x}{B_b.z},\;\frac{B_b.y}{B_b.z}\right).
    \label{eq:pivot_from_generator}
\end{equation}
The ``translation'' encoded in $B_b.x$ and $B_b.y$ does not produce an independent
translational displacement; it shifts the rotation center. \\

\emph{At the velocity level (Lie algebra)  the components are independent. At the
displacement level (Lie group) they are not.} \\

In $3D$ the Lie algebra $\mathfrak{se}(3)$ is six-dimensional: three components for the
rotation axis direction, three for the translation velocity. Chasles' theorem for $3D$
states:

\begin{quote}
\textbf{Chasles' theorem ($3D$):} Any rigid body displacement in $3D$ is a \emph{screw
motion}: a rotation about an axis combined with a translation \emph{along} that same axis
(helical motion).
\end{quote}

Exponentiating a combined $3D$ generator therefore yields a genuine screw -- rotation and
translation are simultaneously present in the finite displacement. The rotation axis
direction and the translational pitch are independently adjustable because there exists a
direction (along the axis) that is geometrically orthogonal to the plane of rotation. In
$2D$ no such direction exists: the axis is perpendicular to the entire plane, so
simultaneous rotation and translation always collapse to rotation about a different
center. \\

\begin{table}[!h]
\centering\small
\begin{tabular}{l|c|c}
\hline
\emph{Lie algebra level} & $PGA2DP$ & $3D$ $PGA$ \\
\hline
Dimension of $\mathfrak{se}$ & $3$ & $6$ \\
Velocity generator $\Omega_b$ & $(-v_y,\; v_x,\; \omega)$ & $(v_x,\; v_y,\; v_z,\;
\omega_x,\; \omega_y,\; \omega_z)$ \\
Components independent? & yes & yes \\
\hline
\end{tabular}
\caption{Velocity generators at the Lie algebra level.}
\label{tab:lie_algebra_level}
\end{table}

\begin{table}[!h]
\centering\small
\begin{tabular}{p{3.6cm}|p{4.3cm}|p{4.3cm}}
\hline
\emph{Lie group level} & $PGA2DP$ & $3D$ $PGA$ \\
\hline
Result of $\exp_{\veedot}(\tfrac{1}{2}B)$ & pure rotation & screw motion \\
Finite pivot / screw axis & point $\bigl(\tfrac{B.x}{B.z},\;\tfrac{B.y}{B.z}\bigr)$ & line
(Pl\"{u}cker coordinates) \\
Role of ``translation'' in $B$ & shifts rotation center & adds pitch along axis \\
\hline
\end{tabular}
\caption{Finite displacements at the Lie group level.}
\label{tab:exp_map_2d_vs_3d}
\end{table}

The accumulated generator $B_b$ (and $B_w$) is a Lie algebra element -- its components are
additive and independent \emph{as generators}. The pivot decoded from $B_b$ as
$(B_b.x/B_b.z,\; B_b.y/B_b.z)$ is meaningful as the instantaneous rotation center of the
\emph{finite} displacement $\exp_{\veedot}(\tfrac{1}{2} B_b)$, not of the instantaneous
velocity $\Omega_b$.

When $B_b = B_{b,0} + \Omega_b \cdot t$ (uniform motion from initial condition $B_{b,0}$):

\begin{itemize}
    \item If $B_{b,0} = \mathbf{0}$: the decoded pivot equals the body-frame pivot of
    $\Omega_b$ for all $t$, i.e.\
    $(B_b.x/B_b.z,\; B_b.y/B_b.z) = (\Omega_b.x/\Omega_b.z,\; \Omega_b.y/\Omega_b.z)
    = Q_b$ -- constant and geometrically meaningful.

    \item If $B_{b,0} \neq \mathbf{0}$ with $B_{b,0}.z = 0$ (a translational initial
    offset): the decoded pivot starts at infinity (pure translation at $t=0$) and
    converges asymptotically to $Q_b$ as $t \to \infty$. This makes the decoded pivot hard
    to interpret and should be avoided in visualizations intended to explain the encoding.
\end{itemize}

\emph{Design rule:} use $B_{b,0} = \mathbf{0}$ (body starts at the $M_0$ reference pose)
so that the accumulated generator retains a clean geometric meaning. \\

\newpage