\subsection{Physics modelling}
\label{subsection_physics_modelling}

The key idea is to represent physical entities using objects represented in the underlying
algebra. They have a geometric interpretation and can be manipulated by applying algebraic
operations. This is possible since the algebraic objects we use in geometric algebra,
especially the blades resulting from outer products between vectors, represent subspaces
of the overall space and thus allow for direct algebraic manipulation. At the same time,
they represent or can be attributed to objects of geometric or physical meaning. This
enables implementing geometrically and physically meaningful manipulations by applying
algebraic operations on those subspaces or between them. \\

Euclidean geometric algebra ($EGA$) is limited to objects that contain the origin. Thus,
it can be used to directly model vectors (i.e. directions), lines, and planes through the
orgin. In addition, it can model complex numbers, quaternions and thus rotations in $2D$
and $3D$ well. To realize them, it uses sandwich products based on the geometric product
for calculating the transformation resulting from two consequtive reflections either
across lines ($2D$ case) or across planes ($3D$ case). The resulting transformations are
rotations. The advantage of using geometric algebra to model those rotations is the
intuitive modelling, due to geometric meaning of the objects and the fact that the
sandwich products can be used to transform different objects, like vectors, bivectors or
rotation operators using the same sandwich product formulas in a unified fashion
regardless of the type of object to be transformed. Despite those limitations $EGA$ is
already a very powerful tool to express physical ideas as shown by David Hestenes in
\cite{David_Hestenes_NFCM:1986}. In this work he reformulates classical mechanics using
geometric algebra mainly with $EGA$ as a tool. But he is also showing that even more is
possible, e.g. for formulating the motion of rigid bodies, clearly pointing to $PGA$ in
his chapter 7 on rigid body mechanics, even if he does not call that $PGA$ yet. This is
already looking extremely similar to what is shown later in this section. His book is a
rich source of applications of geometric algebra for many application areas of
mechanics.\\

Projective geometric algebra ($PGA$) is indeed capable to model objects that do not
necessarily contain the origin. Thus it can be used to directly model points, directions,
lines, and planes that may or may not contain the origin. In contrast to $EGA$,
translations can additionally be implemented in $PGA$ directly as orthogonal
transformation using a sandwich product. The operation is realized by two consequtive
reflections across parallel lines ($2D$) or planes ($3D$). It is based on the regressive
form of the geometric product. The reason for this deviation from $EGA$, which uses the
geometric product for that purpose, is that the Euclidean isometries (rotation,
translation, and screw motion) can only be performed in a space that is dual to the space
we use for modelling our projective objects (for a more in depth explanation
see~\cite{Lengyel_pga-illuminated:2024}). The use of the regressive geometric product on
these objects is equivalent to first transforming the object into dual space by applying
the complement operation, then using the geometric product in that space for the intended
transformation and finally transforming the resulting transformed object back into the
original space, by applying the inverse complement operation. Thus the sandwich product
transformation formulas in $PGA$ only differ from the formulas used in $EGA$ by using the
regressive geometric product instead of the geometric product, but are otherwise the same.
\\

The additional degree of freedom in $PGA$ is possible due to embedding the Euclidean space
of dimension $n$ into a projective space of dimension $n+1$, e.g. two-dimensional
Euclidean space is modelled by embedding it in a three-dimensional projective space, and
three-dimensional Euclidean space is modelled in a four-dimensional projective space. In
the $n+1$-dimensional projective space all algebraic objects contain the origin (exactly
as in $EGA$), but in the modelled $n$-dimensional Euclidean space the objects do not need
to contain the origin any longer. They can be located anywhere in the modelled space
including positions at infinity. This can be realized in projective geometry, since
objects of dimension $n+1$ are projected into the modelled space resulting in objects of
dimension $n$ after the projection. A projective point (which models a scalar or
$0D$-object) is a vector representing a ray through the origin in the embedding space (a
$1D$-object), and a $1D$-object, like a line in modelled space, is created by a bivector
in the embedding projective space (a $2D$-object), and so on for higher-grade objects. \\

Due to its larger expressiveness $PGA$ is suitable to model physics of mass points and
rigid bodies including all isometries. For example, it can model position, velocity,
acceleration, force, torque, momentum, angular momentum, etc., and thus it can be used to
describe kinematics as well as dynamics of mass points and rigid bodies. The approach used
here is inspired by \cite{Plane-based_PGA_Dorst-DeKennik:2022} and
\cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023}, but it does not use the
plane-based ansatz with planes as vectors, which are based on dual representations of the
objects to be modelled. It rather uses the approach as described by
\cite{Lengyel_pga-illuminated:2024} using projective geometric algebra based on direct
representations of points, lines and planes. Lengyel also gives reasons for choosing the
latter approach and explains the background in more detail compared to the short
indications given here. \\

Objects modelled in a projective space can be multiplied by a scalar value without
changing the \emph{geometric} meaninig of the object in the modelled Euclidean space, i.e.
its interpretation after the projection. Equality of modelled geometric objects is thus
often only defined up to a scalar factor. This kind of equality up to a scalar factor is
defined as congruence (see e.g. \cite{Browne_Grassmann-Algebra_Vol1:2012}). Therefore, it
is important for modelling of physical facts that unitization is used consciously to avoid
false interpretations. For example, a heavy point can be modelled by multiplying a
\emph{unitized} point with the scalar value of its mass, but of course the multiplication
with $m$ must happen after unitization in order to keep the scalar factor $m$ in the
calculation. Other geometric objects need to be unitized in order to focus on the intended
geometric operation and avoid additional scaling, i.e. the rotation operators. They must
be unitized before use as well. Thus it is advisable to think through the geometric
operations and their physical interpretation well to avoid suprises and to be consistent.
\\

\cite{Lengyel_poor-foundations_GA:2024} points out additional potential mathematical
inconsistencies in applications of geometric algebra as used up-to-now quite often.
Further hints on mathematical inconsistencies in some GA approaches can be found as well
in \cite{Kritchevsky_case_against_GA:2024}. The goal of the this work is to be as
consistent and mathematically rigorous as possible implementing GA functionality, and at
the same time providing an intuitively applicable framework for modelling physics and
applications. \\


\subsubsection{Modelling force and torque}
\label{subsubsection_modelling_force_and_torque}

\cite{Browne_Grassmann-Algebra_Vol1:2012} (see sections 4.1-4.5) is a very good source
showing how to use Grassmann algebra and projective geometry in order to model forces and
moments. Classical approaches based on vector algebra only model forces by providing
vectors of adequate magnitude and direction to represent the force. However, the
physically very meaningful ``line of action'', which determines which torque is created by
the force, is not modelled mathematically in the classical approach at all. It rather
needs to be derived from the description of the context in order to know where the force
is actually acting. Thus force and torqure are not directly interlinked in the classical
description, but must be modelled as separate physical effects.\\

This changes in $PGA$. Any force can be directly modelled by its line of action, which
encodes the force \emph{and} the torque generated by the force at the origin within one
algebraic object. A force $f$ is modelled by a line $F$. If $f$ is the force vector of
corresponding magnitude and direction, and $Q$ is any unitized projective point on its
line of action then $F$ is an algebraic object representing the force $f$ as well as the
resulting line of action directly:
\begin{equation}
F = Q \wedge f
\label{eq:line_of_action}
\end{equation}
The resulting bivector $F$ represents a line, a geometric object which contains all
physically relevant information of the force $f$ and the torque generated by $f$. For that
reason~\cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023} call this object ``forque''.
It can directly be used to recover the force vector and to calculate the resulting free
torque caused by the force either at the origin or at any point in space. \\

But first, let's have a closer look how this is accomplished. We can separate $F$ into its
components. Each unitized projective point $Q$ is the sum of the origin $O$ and a vector
$q$ parallel to the projective plane (in $2D$) or projective space (in $3D$) pointing from
$O$ to $Q$, such that $Q = O + q$ (``projective point split''). Thus $F$ can be split into
two bivector components
\begin{equation}
    F = Q \wedge f = (O + q) \wedge f =
    \underbrace{O \wedge f}_{\text{=force}} + \underbrace{q \wedge f}_{=torque}
    = \weight{F} + \bulk{F}
\label{eq:line_of_action_components}
\end{equation}
where the first term encodes the force within the weight components of the bivector, while
the second term encodes the resulting free torque relative to the origin caused by the
action of $f$ at $Q$ -- or alternatively anywhere along the line of action -- within the
bulk components of the bivector. \\

To see this even more clearly, let's look at the component equations of both terms written
as bivectors for the case of $2D$ $PGA$ (the corresponding basis bivectors are shown on
the right-hand side in square brackets for reference):
\begin{subeqnarray}
    O \wedge f & = &
    \begin{pmatrix}
        0 \\ 0 \\ 1 
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         f_x \\ f_y \\ 0 
    \end{pmatrix}
    =
    \begin{pmatrix}
         f_x \\ f_y \\ 0 
    \end{pmatrix}
    \qquad \qquad \quad \;\;\;\;
    \begin{bmatrix}
         \bv{31} \\ \bv{32} \\ \bv{12} 
    \end{bmatrix}
    \slabel{eq:pga2dp_force_line_force_component} \\[10 pt]
        q \wedge f & = &
    \begin{pmatrix}
        q_x \\ q_y \\ 0 
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         f_x \\ f_y \\ 0 
    \end{pmatrix}
    =
    \begin{pmatrix}
         0 \\ 0 \\ q_x f_y - q_y f_x
    \end{pmatrix}
    \qquad
    \begin{bmatrix}
         \bv{31} \\ \bv{32} \\ \bv{12} 
    \end{bmatrix}
    \slabel{eq:pga2dp_force_line_torque_component}
\end{subeqnarray}
where equation (\ref{eq:pga2dp_force_line_force_component}) shows how the force is encoded
in the first two components of the bivector and  equation
(\ref{eq:pga2dp_force_line_torque_component}) shows how the torque is encoded in the third
component of the bivector $F$. \\

For $3D$ $PGA$ the encoding is similar, but now requires six bivector components instead
of three (the corresponding basis bivectors are shown on the right-hand side in square
brackets for reference again):
\begin{subeqnarray}
    O \wedge f & = &
    \begin{pmatrix}
        0 \\ 0 \\ 0 \\ 1 
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         f_x \\ f_y \\ f_z \\ 0 
    \end{pmatrix}
    =
    \begin{pmatrix}
         f_x \\ f_y \\ f_z \\ 0 \\ 0 \\ 0
    \end{pmatrix}
    \qquad \qquad \quad \quad \, %\;\;
    \begin{bmatrix}
         \bv{41} \\ \bv{42} \\ \bv{43} \\ \bv{23} \\ \bv{31} \\ \bv{12} 
    \end{bmatrix}
    \slabel{eq:pga3dp_force_line_force_component} \\[10 pt]
        q \wedge f & = &
    \begin{pmatrix}
        q_x \\ q_y \\ q_z \\ 0 
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         f_x \\ f_y \\ f_z \\ 0 
    \end{pmatrix}
    =
    \begin{pmatrix}
         0 \\ 0 \\ 0 \\ q_y f_z - q_z f_y \\ q_z f_x - q_x f_z \\ q_x f_y - q_y f_x
    \end{pmatrix}
    \qquad
    \begin{bmatrix}
         \bv{41} \\ \bv{42} \\ \bv{43} \\ \bv{23} \\ \bv{31} \\ \bv{12} 
    \end{bmatrix}
    \slabel{eq:pga3dp_force_line_torque_component}
\end{subeqnarray}

The first three bivector components modell the force while the last three components model
the torque. The force vector $f$ itself can be retrieved from $F$ by using the attitude
function $att()$:
\begin{equation}
    f = att(F) = F \vee \overline{\bv{n}}
    \label{eq:force_attitude}
\end{equation}
with $\bv{n}$ denoting the basis vector of the projective dimension that squares to zero
in the respective algebra (\bv{3} for $2D$ and \bv{4} for $3D$). The torque generated
relative to the origin $O$ is encoded in the bulk of the bivector:
\begin{equation}
    M_{O} = bulk(F) = \bulk{F}
    \label{eq:force_torque_origin}
\end{equation}
while the resulting torque from the application of $f$ at any given point $R$ in space can
be retrieved from $F$, either using any known point $Q$ on $F$\footnote{Which is always
available from the function $sup(F)$. It provides the support point on $F$ with the
shortest distance to the origin $O$ by orthogonally projecting the origin onto $F$.}, or
alternatively by projecting $R$ onto $F$ and using the perpendicular distance vector
$d_\perp$ from $R$ towards the line $F$ in order to calculate the resulting free torque
$M_F(R)$ at the position $R$ as
\begin{subeqnarray}
    M_F(R) & = &  \bulk{\left[ (Q-R) \wedge f \right]} = \bulk{\left[ (Q-R) \wedge att(F)
             \right]} \slabel{eq:action_line_moment_simple} \\
    M_F(R) & = & \bulk{\left[ d_\perp \wedge f \right]} = \bulk{\left[ (F \vee (R \wedge
             \rightweightdual{F}) - R) \wedge att(F) \right]}
    \slabel{eq:action_line_moment_projection}
\end{subeqnarray}
where equation~(\ref{eq:action_line_moment_simple}) is the formula with the least
computational effort using a known point $Q$ on $F$, while
equation~(\ref{eq:action_line_moment_projection}) uses the othogonal projection of $R$
onto $F$ to calculate the perpendicular distance vector $d_\perp$. \\

The resulting total force and torque acting on an object are given by:
\begin{equation}
    F_{total} = \sum_i{F_i}
    \label{eq:total_sum_forque}
\end{equation}
This resulting sum will be used later for connecting kinematics and dynamics. \\

Things to remember:
\begin{itemize}
    \item Neither a vector nor a bivector are located anywhere. They can be moved around
    freely. A line however represents a bound vector that is bound by its line. Thus the
    vector can only be moved along its line of action without changing the geometric and
    physical meaning (see~\cite{Browne_Grassmann-Algebra_Vol1:2012}, p. 169).
    \item A free torque is modelled by a bivector with force components being zero, i.e.
    the resulting force vector (e.g. from summing several forces acting on an object) has
    no direction and zero length, but the bivector still might contain a resulting torque
    value.
    \item The torque calculated from either equation (\ref{eq:action_line_moment_simple})
    or (\ref{eq:action_line_moment_projection}) is necessarily a free torque as can be
    seen from the formulas. The origin does not occur in the result anymore due to the
    difference of the two projective points involved.
    \item The sum of forces and their induced torque values can be directly created by the
    sum of their respective lines of action, i.e. by summing the bivectors representing
    those lines. The resulting lines of action contain all relevant information incl. the
    resulting vector sum of the forces and the resulting torque values from the summation.
    There is no need to calculate these separately. The result will always be a resulting
    bivector that potentially contains a bound vector and a resulting torque.
    (see~\cite{Browne_Grassmann-Algebra_Vol1:2012}, p. 176).
\end{itemize}

% The geometrical situation is depicted in figure XXX for a two-dimensional case.

\newpage

\subsubsection{Modelling linear and angular momentum}
\label{subsubsection_modelling_momentum}

Momentum in the classical approach is split into linear momentum $\vec{p} = m\vec{v}$ and
angular momentum $\vec{L} = \vec{x} \times \vec{p} = \vec{x} \times m \vec{v}$. Both are
represented as vectors, but they represent entities with different physical meaning. So
they cannot be combined into one object in the classical approach, e.g. by adding them. \\

The $PGA$ approach to momentum is to define both, the linear and the angular aspects,
within one algebraic entity. This is in complete analogy to what we have seen in the last
subsection for force and torque resulting in its line of action modelled by a bivector. \\

The momentum $P$ of a point mass $m$ at a projective (unitized) position $X$ that moves
with velocity $\dot{X}$ is defined as a bivector $P$ as
\begin{equation}
    P = X \wedge m \dot{X}
    \label{momentum_definition}
\end{equation}
It defines the overall momentum at position $X$ including linear and angular momentum in
the bivector $P(X)$, which can be shown by splitting the projective point $X$ into its
components $X = O + x$ (``projective point split''). In the $3D$-case with the origin $O =
O_{3dp} = (0,0,0,1)$ and the direction vector $x$, where $x = (X_x, X_y, X_z, 0)$, i.e.
specifically with $X_w = 0$, which represents a direction towards a point at infinity.
This results in
\begin{subeqnarray}
    P(X) &=& X \wedge m \dot{X} = (O + x) \wedge m \dot{X} \nonumber \\
         &=& O \wedge m \dot{X} + x \wedge m \dot{X} \nonumber \\
         &=& O \wedge m \dot{(O + x)} + x \wedge m \dot{(O + x)} \nonumber \\
         &=& O \wedge m \dot{x} + x \wedge m \dot{x} \\
         &=& \underbrace{\quad\weight{P}\quad}_{\text{linear}}
             \;+\underbrace{\quad\bulk{P}\quad}_{\text{angular}}
\end{subeqnarray}
$P$ defines an instantaneous momentum line where the weight part $\weight{P}$ of the
bivector $P$ encodes the linear momentum and the bulk part $\bulk{P}$ encodes the angular
momentum with respect to the origin.\footnote{The ``projective point split'' works in this
way, since the origin is a destinguishable point. This is in contrast to $EGA$ or the
classical approach where the orgin is a null-vector, which has no lasting effect if it is
added to another point. This is one reason why $PGA$ can combine linear and angular
aspects as separate physical effects in a single algebraic object. The other is that the
required 6 degrees of freedom can be encoded in a bivector.} \\

The physical aspects encoded in the momentum line $P$ can be retrieved in complete analogy
to the case of force and torque in subsection
\ref{subsubsection_modelling_force_and_torque}: \\

The linear momentum vector $p$ itself can be retrieved from $P$ by using the attitude
function $att()$:
\begin{equation}
    p = m v = m \dot{x} = att(P) = P \vee \overline{\bv{n}}
    \label{eq:linear_momentum_from_attitude}
\end{equation}
with $\bv{n}$ denoting the basis vector of the projective dimension that squares to zero
in the respective algebra (\bv{3} for $2D$ and \bv{4} for $3D$). \\

The angular momentum $L$ generated relative to the origin $O$ is encoded in the bulk of
the bivector:
\begin{equation}
    L_{O} = x \wedge m \dot{x} = bulk(P) = \bulk{P}
    \label{eq:angular_momentum_wrt_origin}
\end{equation}
while the resulting angular momentum relative to any given point $R$ in space can be
retrieved from the momentum line $P$ by
\begin{equation}
    L(R) = \bulk{\left[ (X-R) \wedge m\dot{X} \right]}
         = \bulk{\left[ (X-R) \wedge att(P) \right]}
           \slabel{eq:action_momentum_line_simple}
\end{equation}
using $X$ as a known point on $P$. This provides the resulting angular momentum $L(R)$
with respect to $R$. The momentum line encodes the resulting momentum including its linear
and angular aspects relative to any arbitrarily chosen point. \\

The equations (\ref{eq:linear_momentum_from_attitude}) and
(\ref{eq:angular_momentum_wrt_origin}) given above permit to separate the linear and
angular effects as in the classical approach. However, these aspects need not be separated
at all and can be calculated together in $PGA$.\\

Using eqn.~(\ref{momentum_definition}) and the definition $X=O + x$ with $X$ as a unitized
projective point, $O$ as the chosen origin and $x$ as the direction vector pointing from
$O$ to $X$ we can write
\begin{equation}
    P_O(X) = X \wedge m \dot{X} = (O + x) \wedge m \dot{(O + x)}
    \label{momentum_written_with_O}
\end{equation}
Compared to eqn.~(\ref{momentum_definition}) an index $O$ was added to $P$, since the
formula (\ref{momentum_written_with_O}) seems to confirm that the momentum depends on the
choice of the origin. Now it is important to demonstrate  that this is \emph{not} the
case. We are going to demonstrate that the described physics does not depend on the chosen
origin and not on the resulting coordinates used for its description. \\

\emph{First approach}: Use a transformation of the standpoint from which we describe the
physical situation without actually moving the coordinate system (we still want use the
chosen coordinate system to describe the vectors and bivectors used): We go to a virtual
origin $O' = Y$, but describe the situation in the same coordinate system. This results in
describing the origin as $O = Y - y$ as seen from $Y$. Inserting that definition into
eqn.~(\ref{momentum_written_with_O}) and using the geometric vector identity
\begin{equation}
X = Y + (X-Y) = Y + (O+x-(O+y)) = Y + (x-y)
\label{eq:identity_transformation_X_Y}
\end{equation}
we express $P(X)$ as viewed from $Y$ by reformulating the equation for $P$ relative to the
origin $O$
\begin{eqnarray}
    P_O(X) &=& X \wedge m \dot{X} = (O + x) \wedge m \dot{(O + x)} \nonumber \\
           &=& (Y - y + x) \wedge m \dot{(Y - y + x)} \nonumber \\
           &=& (Y + (x-y)) \wedge m \dot{(Y + (x - y))} = P_Y(X)
\end{eqnarray}
where the last equation obviously is the same as eqn.~(\ref{momentum_written_with_O}), but
here Y takes the role of $O$ and $(x-y)$ takes the role of the vector $x$ in the old
description. The dot used for the time-derivative refers to the whole expression in
parentheses. So the same physical situation was described from another viewpoint. However,
the algebraic object $P$ is unchanged, i.e. $P_O(X) = P_Y(X)$. This clearly shows that $P$
does not depend on the chosen origin and that eqn.~(\ref{momentum_written_with_O}) is
indeed a coordinate free definition. \\

\emph{Second approach}: This one is not so explicit on being independent of the origin. We
just express $X$ by $Y$ using eqn.~(\ref{momentum_written_with_O}) and
(\ref{eq:identity_transformation_X_Y}) directly to write
\begin{eqnarray}
    P_O(X) &=& X \wedge m \dot{X} \nonumber \\
           &=& (Y + (X-Y)) \wedge m \dot{(Y + (X-Y))} \nonumber \\
           &=& (Y + (x-y)) \wedge m \dot{(Y + (x - y))} = P_Y(X)
\end{eqnarray}
We can again see that this is structurally identical to
eqn.~(\ref{momentum_written_with_O}) just seen from $Y$ instead from $O$. \\

The total momentum of a set of mass points $m_i X_i$ is defined as the sum of all momentum
blades.
\begin{equation}
    P_{total} = \sum_i{P_i}
    \label{eq:total_sum_momentum}
\end{equation}
$P_{total}$ will always be a momentum blade in $2D$ $PGA$, while in $3D$ $PGA$ (a
four-dimensional algebra) it could also result in a general bivector that might not be
factorizable into a blade (see \cite{Browne_Grassmann-Algebra_Vol1:2012}). That would be
the case, if $P \wedge P \ne 0 \; \ps$, i.e. when the result is a pseudoscalar that is not
equal to a zero multiple of the pseudoscalar \bv{1234}. This test can always be used to
determine whether a bivector is a blade in $3D$ $PGA$. \\

Using discrete points $X_{w,i}$ of the world frame indexed by $i$ that can be moved by a
motor $M(t) = exp_\veedot[\frac{1}{2}B_w(t)] \veedot M_0$ in the world frame as by
eqn.~(\ref{eq:pga_sandwich_tranformation}) $X_{w,i} = M \veedot X_{0,i} \veedot \rrev{M}$
results in a local speed of $\dot{X}_{w,i} = rcmt(\Omega_w,X_{w,i})$ as shown in
eqn.~(\ref{eq:rcmt_pga_world}). Therefore, we can write the total momentum $P_{w}$ in the
world frame as
\begin{eqnarray}
    P_{w} &=& \sum_i{P_i} \nonumber \\
          &=& \sum_i X_i \wedge m_i \dot{X}_i \nonumber \\
          &=& \sum_i X_i \wedge m_i \; rcmt(\Omega_w,X_i) \nonumber \\
    P_{w} &\equiv& I_w[\Omega_w]
                \label{eq:intertia_map_world_frame}
\end{eqnarray}
where eqn.~(\ref{eq:intertia_map_world_frame}) is a linear function of $\Omega_w$ that
maps the bivector rate in the world frame $\Omega_w$ to the total momentum $P_w$. It is a
function that takes a bivector as an argument and returns a bivector (the ``intertia map''
$I_w$). It describes the influence of the mass distribution of the points on the movement
via the combined Newton/Euler law
\begin{equation}
    \dot{P}_{w}=F_{w}
    \label{eq:Newton_Euler_law}
\end{equation}
(for the total force in the \emph{world frame} $F_{total} = F_{w}$ see
eqn.~(\ref{eq:total_sum_forque})). If we go from discrete mass points to a continuous mass
distribution, the sum has to be replaced by an integral over $dm = \rho dV$. \\

The equation $\dot{P}_{w}= X_w \wedge m \ddot{X}_w = F_{w}$ cannot be integrated directly
since the second derivative $\ddot{X}_w$ cannot be isolated on the left-hand side of the
equation as required for direct integration. We introduce another variable that can be
isolated, enabling us to solve the ODE. For this purpose we use the rate of change
$\Omega$ (a vector in $2D$ and a bivector in $3D$). The integration of that variable will
then allow for computation of position, speed, etc. as a separate step. \\

Calculating the derivative $\dot{P}_w$ in the world frame in general yields two terms
\begin{equation}
    \dot{P}_w = \dot{I}_w[\Omega_w] + I_w[\dot{\Omega}_w]
    \label{eq:momentum_derivative_world_frame}
\end{equation}
since in the world frame the inertia $I_w[\Omega_w]$ map will change over time due to
moving points as well as due to the rate of change $\Omega_w$. Therefore, it is reasonable
to calculate the inertia map in the body frame, since it is constant in that frame, if the
body frame is chosen to be fixed to the modelled rigid object. This avoids the effort to
calculate the time derivative of the inertia map (the first term on the right-hand side
above) that would be required, if the calculation would be performed in the world frame.
\\

If we want to express the physical problem in the body frame, we have to transform the
components of eqn.~(\ref{eq:intertia_map_world_frame}) accordingly. Using
eqn.~(\ref{eq:world_to_body_pga}), i.e. $I_b = \rrev{M} \veedot I_w \veedot M$ and
eqn.~(\ref{eq:body_to_world_pga}) $\Omega_w  = M \veedot \Omega_b \veedot \rrev{M}$ then
results in
\begin{subeqnarray}
    P_{w} &=& I_w[\Omega_w] \nonumber \\
    \rrev{M} \veedot P_{w} \veedot M &=&
    \rrev{M} \veedot I_w[\rrev{M} \veedot \Omega_w \veedot M] \veedot M \nonumber \\
    P_{b} &=& I_b[\Omega_b] \nonumber
\end{subeqnarray}
which finally results in an expression for the inertia map
\begin{equation}
    I_b[\Omega_b] \equiv \rrev{M} \veedot I_w[ M \veedot \Omega_b \veedot \rrev{M} ]
    \veedot M
    \label{eq:intertia_map_body_frame}
\end{equation}

The same transformation must be applied to $F_w$ for the right-hand side of the
differential equation $\dot{P}_w = F_w$. Therefore, we get in the body frame $F_b =
\rrev{M} \veedot F_w \veedot M$, resulting in $\dot{P}_b = F_b$. \\

The time derivative in the body frame is simpler, as long as we restrict ourselves to
rigid bodies, resulting in only one term, since the rate of change $\Omega_b$ is the only
time dependent function in the body frame
\begin{equation}
    \dot{P}_b = I_b[\dot{\Omega}_b]
    \label{eq:momentum_derivative_body_frame}
\end{equation}
otherwise the term $\dot{I}_b[\Omega_b]$
would have to be covered as well, as in eqn.~(\ref{eq:momentum_derivative_world_frame})
above. \\

The mapping to the classical intertia can be done when splitting $\Omega_b$ relative to
the centroid of the body with a projective point split on $\Omega_b$ and on
$I_b[\Omega_b]$ as shown by \cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023} on page
25f and adjusted to our approach using regressive geometric products. \\

Now we introduce some typically used reference points to describe kinemetics via the
bivector rate of change $\Omega$ and dynamics via the total momentum $P$:
\begin{itemize}
    \item \emph{Fixed body refence point} $O_b$ used as origin of the body frame. When the
    body is moved by the motor $M$ it will move in the world frame to $O_{b,w} = M \veedot
    O_b \veedot \rrev{M}$.
    \item \emph{Relative locations $r_i$ versus $O_b$} are positions in the body frame.
    Their projective counterpart in the body frame is $R_i = O_b + r_i$. When transformed
    to the world frame we get $X_i$ in the world frame by the usual transformation $X_i =
    M \veedot R_i \veedot \rrev{M} = M \veedot (O_b + r_i) \veedot \rrev{M}$.
    \item \emph{Center of mass} in the body frame is chosen to be located at $O_b$, such
    that $\sum_i{m_i r_i} = \vec{0}$. Thus the origin of the body frame is located at the
    center of mass. The center of mass in the world frame $C_w$ then is $C_w \equiv
    \sum_i{m_i X_i} / \sum_i{m_i} = \sum_i{m_i (M \veedot (O_b + r_i) \veedot \rrev{M})} /
    \sum_i{m_i} = M \veedot O_b \veedot \rrev{M}$.
\end{itemize}

The last missing piece for dynamics is the time derivative of $P_{w}$ expressed as time
rate of change $\Omega_b$ of the inertia in the body system. Applying
eqn.~(\ref{eq:rcmt_pga_world}) in the body frame, which is not only valid for any $X$, but
for any object moved by a motor $M$, we get
\begin{equation}
    \frac{d}{dt} P_b = \frac{d}{dt} I_b[\Omega_b] =
    rcmt(\Omega_b,I_b[\Omega_b]) + I_b[\dot{\Omega}_b]
\end{equation}
where the last term arises since $\Omega_b$ is also time-dependent. Solving
eqn.~(\ref{eq:Newton_Euler_law}), but in the body frame, we get
\begin{equation}
    F_b = \dot{P}_b = rcmt(\Omega_b,I_b[\Omega_b]) + I_b[\dot{\Omega}_b]
\end{equation}
resolving this for $\dot{\Omega}_b$ we get
\begin{eqnarray}
    I_b[\dot{\Omega}_b] &=& F_{b} - rcmt(\Omega_b,I_b[\Omega_b]) \\[5pt]
    \dot{\Omega}_b &=& I_b^{-1}\left[ F_{b} - rcmt(\Omega_b,I_b[\Omega_b]) \right]
\end{eqnarray}

Together with the ansatz in the body system eqn.~(\ref{eq:motor_pga_body_frame}) that can
be used to calculate M(t) when we know $\Omega_b$ at a given time $t$ from
\begin{equation*}
   M(t) = M_0 \veedot exp_\veedot(\frac{1}{2}B_b(t))
        = M_0 \veedot exp_\veedot(\frac{1}{2}\Omega_b t) 
\end{equation*}
and the rate of change in the body frame according to
eqn.~(\ref{eq:motor_rate_body_frame_pga}) we get
\begin{equation}
    \dot{\Omega}_b = I_b^{-1}\left[ F_{b} - rcmt(\Omega_b,I_b[\Omega_b]) \right] 
    \label{eq:ode_system_body_frame}
\end{equation}
as an ODE that can be solved as initial value problem. Due to $\dot{\Omega}_b =
\ddot{B}_b$ it actually is a $2^{nd}$ order ODE (like the classical Newton/Euler equations
that can be integrated directly from the instantaneous acceleration and the angular
acceleration, respectively, in order to calculate speed and position). Here we rather
integrate from $\ddot{B}_b$ in order to calculate $\dot{B_b} = \Omega_b$ and $B_b$. Thus
we require initial values for $\Omega_b$ and $B_b$. The actual positions of the points of
the rigid body can then be computed from eqn.~(\ref{eq:pga_sandwich_tranformation})
$X_w(t) = M(t) \veedot X_0 \veedot \rrev{M}(t)$. The speed of the points can be calculated
either from eqn.~(\ref{eq:rcmt_pga_world}) $\dot{X}_w = rcmt(\Omega_w,X_w)$ or
eqn.~(\ref{eq:rcmt_pga_world_incl_X0_of_t}) $\dot{X}_w = rcmt(\Omega_w,X_w)+
(\dot{X_b})_w$ -- which one is applicable depends on whether the rigid body is fixed in
the body frame or moves in the body frame. The expressions in the world frame can of
course be expressed in the body frame as well using eqn.~(\ref{eq:body_to_world_pga})
$\Omega_w = M \veedot \Omega_b \veedot \rrev{M}$ and $X_w = M \veedot X_b \veedot
\rrev{M}$.\\

For that purpose we need the inertia map $I_b[\Omega_b]$ and its inverse to set up the
right-hand side of the ODE system given in equations~(\ref{eq:ode_system_body_frame}). We
start with the $2D$ case where the rate of change $\Omega_b$ is a vector, as shown at the
end of subsection \ref{subsection_modelling_motion}. We assume discrete points $X_i$ and
use the wedge product $X \wedge Y$ and the regressive commutator product term $m \;
rcmt(\Omega_b,X)$ to express the total momentum $P$ written in component form:
\begin{subeqnarray}
    X \wedge Y & = &
    \begin{pmatrix}
        X_x \\ X_y \\ X_z  
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         Y_x \\ Y_y \\ Y_z
    \end{pmatrix}
    =
    \begin{pmatrix}
         X_z Y_x - X_x Y_z \\ X_z Y_y - X_y Y_z \\ X_x Y_y - X_y Y_x
    \end{pmatrix}
    \slabel{eq:wedge_X_Y_2D} \\[10 pt]
    m \; rcmt(\Omega_b,X) & = & m
    \begin{pmatrix}
         \Omega_y X_z - \Omega_z X_y \\ \Omega_z X_x - \Omega_x X_z \\ 0
    \end{pmatrix} 
\end{subeqnarray}

Now the inertia map for a point $X$ with mass $m$ and rate of change $\Omega$ can be
expressed in the $2D$ case as
\begin{eqnarray}
    I_{2D}[\Omega] &=& X \wedge m \; rcmt(\Omega,X) \nonumber \\[10pt]
    &=& m 
    \begin{pmatrix}
        X_z (\Omega_y X_z - \Omega_z X_y) \\
        X_z (\Omega_z X_x - \Omega_x X_z) \\
        X_x (\Omega_z X_x - \Omega_x X_z) - X_y (\Omega_y X_z - \Omega_z X_y)
    \end{pmatrix} \nonumber \\[10pt]
    I_{2D}[\Omega] &=& m
    \begin{bmatrix}
        0        &  X_z^2   & -X_y X_z      \\
        -X_z^2   &  0       &  X_x X_z      \\
        -X_x X_z & -X_y X_z & X_x^2 + X_y^2
    \end{bmatrix}
    \begin{bmatrix}
        \Omega_x \\ \Omega_y \\ \Omega_z
    \end{bmatrix}
    \label{eq:intertia_map_components_2D}      
\end{eqnarray}
where eqn.~(\ref{eq:intertia_map_components_2D}) is linear in $\Omega$ as expected. Of
course, the total inertia map needs to be summed for the contributions from each point,
i.e. $I_{total} = \sum_i{I_i}$. Since all points of the rigid body see the same rate of
change the total inertia map can be summed for all mass point contributions independent of
the rate of change and the resulting total interia map can be inverted either symbolically
or numerically to get its inverse $I_{2D}^{-1}$ as needed to set up the right-hand side of
the system in eqn.~(\ref{eq:ode_system_body_frame}). The inertia map in $2D$ maps the rate
of change vector to the momentum bivector. The inverse inertia map than maps from a
momentum bivector to the corresponding rate of change vector. \\

Now let's do the same for the $3D$ case, where the rate of change is a bivector. The
component expressions are generated from the corresponding product tables in
\path{ga_prdxpr} and integrated into the products in the ga library. 
\begin{subeqnarray}
    X \wedge Y & = &
    \begin{pmatrix}
        X_x \\ X_y \\ X_z  \\ X_w
    \end{pmatrix}
    \wedge
    \begin{pmatrix}
         Y_x \\ Y_y \\ Y_z \\ Y_w
    \end{pmatrix}
    =
    \begin{pmatrix}
         X_w Y_x - X_x Y_w \\
         X_w Y_y - X_y Y_w \\
         X_w Y_z - X_z Y_w \\
         X_y Y_z - X_z Y_y \\
         X_z Y_x - X_x Y_z \\
         X_x Y_y - X_y Y_x
    \end{pmatrix}
    \slabel{eq:wedge_X_Y_3D} \\[10 pt]
    m \; rcmt(\Omega_b,X) & = & m
    \begin{pmatrix}
         \Omega_{vy} X_z - \Omega_{vz} X_y + \Omega_{mx} X_w \\
         \Omega_{vz} X_x - \Omega_{vx} X_z + \Omega_{my} X_w \\
         \Omega_{vx} X_y - \Omega_{vy} X_x + \Omega_{mz} X_w \\
         0
    \end{pmatrix} 
\end{subeqnarray}
 
Now the intertia map for a point $X$ with mass $m$ and rate of change $\Omega$ can be
expressed in the $3D$ case as
\begin{align}
    &I_{3D}[\Omega] = X \wedge m \; rcmt(\Omega,X) = \nonumber \\[10pt]
    &m 
    \begin{pmatrix}
        X_w (\Omega_{vy} X_z - \Omega_{vz} X_y + \Omega_{mx} X_w) \\
        X_w (\Omega_{vz} X_x - \Omega_{vx} X_z + \Omega_{my} X_w) \\
        X_w (\Omega_{vx} X_y - \Omega_{vy} X_x + \Omega_{mz} X_w) \\
        X_y (\Omega_{vx} X_y - \Omega_{vy} X_x + \Omega_{mz} X_w) 
      - X_z (\Omega_{vz} X_x - \Omega_{vx} X_z + \Omega_{my} X_w) \\
        X_z (\Omega_{vy} X_z - \Omega_{vz} X_y + \Omega_{mx} X_w)
      - X_x (\Omega_{vx} X_y - \Omega_{vy} X_x + \Omega_{mz} X_w) \\
        X_x (\Omega_{vz} X_x - \Omega_{vx} X_z + \Omega_{my} X_w) - X_y (\Omega_{vy} X_z -
      \Omega_{vz} X_y + \Omega_{mx} X_w) \end{pmatrix} = \nonumber \\[10pt]
    &m
    \begin{bmatrix}
      0              &  X_z X_w       & -X_y X_w       & X_w^2    &  0       &  0       \\
      -X_z X_w       &  0             &  X_x X_w       & 0        &  X_w^2   &  0       \\
       X_y X_w       & -X_x X_w       &  0             & 0        &  0       &  X_w^2   \\
       X_y^2 + X_z^2 & -X_x X_y       & -X_x X_z       & 0        & -X_z X_w &  X_y X_w \\
      -X_x X_y       &  X_x^2 + X_z^2 & -X_y X_z       &  X_z X_w &  0       & -X_x X_w \\
      -X_x X_z       & -X_y X_z       &  X_x^2 + X_y^2 & -X_y X_w &  X_x X_w &  0
    \end{bmatrix}
    \begin{bmatrix}
        \Omega_{vx} \\ \Omega_{vy} \\ \Omega_{vz} \\
        \Omega_{mx} \\ \Omega_{my} \\ \Omega_{mz}
    \end{bmatrix} 
    \label{eq:intertia_map_components_3D}   
\end{align}
where eqn.~(\ref{eq:intertia_map_components_3D}) is linear in $\Omega$ as expected. Again,
the total inertia map needs to be summed for the contributions from each point, i.e.
$I_{total} = \sum_i{I_i}$. The resulting total interia map can be inverted numerically to
get its inverse $I_{3D}^{-1}$ as needed to set up the right-hand side of the system in
eqn.~(\ref{eq:ode_system_body_frame}). The inertia map in $3D$ maps the rate of change
bivector to the momentum bivector. The inverse inertia map than maps from a momentum
bivector to the corresponding rate of change bivector.\\

As shown in \cite{Dynamics_in_PGA_plane-based_Dorst-DeKennik:2023} the inertia map is
additive. That means that inertia maps of parts of a body consisting of several parts that
move with the same bivector rate can be added to get the total inertia. \\

\newpage


\subsubsection{Solving the Newton/Euler equations numerically}
\label{subsubsection_solving_newton-euler_numerically}

In classical mechanics the Newton equation states that change of linear momentum $\vec{p}
= m \vec{v}$ of a mass point of mass $m$ moving with speed $\vec{v}$ is driven by an
external force $\vec{f}$ acting on the mass. This results in $\dot{\vec{p}} = m
\dot{\vec{v}} = m \ddot{\vec{x}}_{cm} = \vec{f}$ for translational motion in a system with
constant mass $m$. The center of mass at position $\vec{x}_{cm}$ moves with the velocity
$\vec{v}$ as a result. \\

The analogous equation for rotational motion, attributed to Euler, states that the
temporal change of angular momentum $\vec{L} = \vec{x} \times \vec{p} = \vec{x} \times m
\vec{v}$ is driven by an external angular momentum $\vec{M}$. $\vec{M}$ is given as
$\vec{M} = \vec{r} \times \vec{f}$ with $\vec{r}$ as the vector pointing from the
rotational axis to the point of action of the force $\vec{f}$. The equation for rotational
motion $\dot{\vec{L}} = J \dot{\vec{\omega}} = \vec{M}$ for a rigid body with moment of
inertia $J = \int_{m}{r^2_\perp dm} = \int_{V}{r^2_\perp \rho dV}$ with $r_\perp$ as the
perpendicular distance of the mass point within the rigid body to the momentary rotational
axis. The equation determines the temporal development of the orientation of the rigid
body which spins with the angular speed $\nrm{\vec{\omega}}$ around the momentary rotation
axis defined by $\vec{\omega}$  as a result. \\

Due to the same structure the equations can be written as one system of equations that can
be solved in parallel (as uncoupled equations). In the $2D$ case the translational part
requires two degrees of freedom (DoF) and the rotational part needs one DoF resulting in a
total of three DoF. It can be written compactly as a system introducing the inertia matrix
$I$ as a linear function that maps the acceleratory degrees of freedom to the
corresponding external force and torque components:
\begin{equation}
    \underbrace{
    \begin{bmatrix}
        m & 0 & 0 \\
        0 & m & 0 \\
        0 & 0 & J
    \end{bmatrix}
    }_{inertia \; I}
    \begin{pmatrix}
        \ddot{x}_x \\
        \ddot{x}_y \\
        \ddot{\phi} 
    \end{pmatrix}
    =
    \begin{pmatrix}
        {f}_x \\
        {f}_y \\
        {M} 
    \end{pmatrix} 
\end{equation}
Solving this system requires to isolate the accelerations by multiplying the system from
the left with the inverse of the inertia map $I^{-1}$ resulting in
\begin{equation}
    \begin{pmatrix}
        \ddot{x}_x \\
        \ddot{x}_y \\
        \ddot{\phi} 
    \end{pmatrix}
    =
    \underbrace{
    \begin{bmatrix}
        \frac{1}{m} & 0           & 0 \\
        0           & \frac{1}{m} & 0 \\
        0           & 0           & \frac{1}{J}
    \end{bmatrix}
    }_{inverse \; inertia \; I^{-1}}
    \begin{pmatrix}
        {f}_x \\
        {f}_y \\
        {M} 
    \end{pmatrix} 
    \label{eq:classical_sys_Newton_Euler_2nd_order}
\end{equation}

Both equations result in a system of ordinary differential equation (ODE) of second order
that requires two initial conditions, one for the initial linear position $\vec{x}(t=0) =
\vec{x}_0$ and the initial speed $\vec{v}(t=0) = \vec{v}_0$ (Newton equation) and one for
the initial angular position $\phi(t=0) = \phi_0$ and the initial angular speed
$\omega(t=0) = \omega_0$ (Euler equation) to be integrated. In the $2D$ case $\phi$ and
$\omega$ are scalar values, while in the $3D$ case they also become vectors each
representing three degrees of freedom.\\

To integrate the equations the system of $2^{nd}$ order must be transfered into a system
of $1^{st}$ order. We have a $2^{nd}$ order ODE for $y(t)$ with $\ddot{y} = \pdd{y}{t}$ of
the general form
\begin{equation}
    \ddot{\vec{y}} = f(\dot{\vec{y}}, \vec{y}, t)
\end{equation}
The vector $y$ has components that we index with $n$. We use an additional index to
distinguish between the position (index 0) and the speed (index 1) of the respective DoF.
We use the substitution
\begin{subeqnarray}
    y_{n,0} &=& y_n \qquad \rightarrow \textit{position} \\
    y_{n,1} &=& \dot{y}_n \qquad \rightarrow \textit{speed}
\end{subeqnarray}
resulting in the following system of $1^{st}$ order for each component $n$ of the vector
$\vec{y}$:
\begin{subeqnarray}
    \dot{y}_{n,0} &=& y_{n,1} \\
    \dot{y}_{n,1} &=& f(y_{n,1}, y_{n,0}, t)
\end{subeqnarray}
If we would have to integrate for more than one point, we could easily add an additional
index at the beginning to extent the system accordingly. However, for now we only focus on
integrating the system for one point (our center of mass) at which we wand to know
position, speed as well as angular orientation and angular speed as function of time. The
substitution above transforms the system in
eqn.~(\ref{eq:classical_sys_Newton_Euler_2nd_order}) into a system of $1^{st}$ order as
follows:
\begin{subeqnarray}
    \dot{x}_x &=& v_x \\
    \dot{v}_x &=& \frac{1}{m}f_x \\
    \dot{x}_y &=& v_y \\
    \dot{v}_y &=& \frac{1}{m}f_y \\
    \dot{\phi} &=& \omega \\
    \dot{\omega} &=& \frac{1}{J} M
\end{subeqnarray}
This $1^{st}$ order ODE system $\dot{\vec{u}} = rhs(\vec{u},t)$ can now be solved be
direct integration, using e.g. a Runge-Kutta-Scheme of $4^{th}$ order accuracy for precise
time integration, and using the initial conditions defined above. As a result we directly
get the positions and velocities as function of time.\\

The approach described above can be used to solve the resulting $PGA$
equation~(\ref{eq:ode_system_body_frame}) $\ddot{B}_b = I_b^{-1}\left[ F_{b} -
rcmt(\Omega_b,I_b[\Omega_b]) \right]$. The overall approach is identical, just with a
slightly different right-hand side adjusted to the $PGA$ formulation. However, this
equation does not directly contain the degrees of freedom we are interested in, like $X$,
$\dot{X}$, and $\ddot{X}$. Instead, we first have to solve the system for the
corresponding (bi)vectors\footnote{$2D$ case: vectors, $3D$ case: bivectors} $B(t)$,
$\dot{B} = \Omega(t)$, and $\ddot{B} = \dot{\Omega}(t)$. For the time integration we apply
suitable initial conditions, and then, as a follow-up step, we calculate the actual points
and their velocities from the defining equations (e.g. based on
equations~(\ref{eq:rexp_pga_motor}) and (\ref{eq:pga_sandwich_tranformation})) that link
the motion bivectors with the positions and speeds we are actually interested in to solve
the given physical task.\\

To make these remarks more concrete let's look into the $2D$ case, specifically into the
formulation of the initial conditions. The position and rate of change vectors in the
\emph{body frame} are defined as
\begin{equation}
    B_b = \begin{pmatrix}
        s_x \\ s_y \\ \phi
    \end{pmatrix}
    ,\quad
     \dot{B}_b = \Omega_b = \begin{pmatrix}
        v_x \\ v_y \\ \omega
    \end{pmatrix}
    ,\quad
     \dot{\Omega}_b = \begin{pmatrix}
        a_x \\ a_y \\ \alpha
    \end{pmatrix}    
\end{equation}
The vector containing all degrees of freedom in the system transformed to $1^{st}$ order
is then $u = (s_x, v_x, s_y, v_y, \phi, \omega)$. Suitable initial conditions for speed
and angular speed can then be directly derived from equations
(\ref{eq:vector_rot_rate_omega}) and (\ref{eq:vector_trans_rate_v_vec}) as starting values
for the integration. The intial value for the position $B_b(t=0)$ is typically set to
$B_{b,t=0} = (0, 0, 0)$ since it reflects the most suitable initial value for $B_b(t) =
\Omega_b(t) t + B_{b,t=0}$. This choice yields the identity transformation for $t=0$ as
$M(t=0) = exp_\veedot[B_b(t=0)] = \ps$. \\

Basic examples for translational and rotational motion based on the Newton-Euler equations
are implemented as tests in \path{ga_test/src/ga_appl_physics_test.hpp} in both
approaches, the classical approach and the approach directly formulated in $PGA$ for
simple reference cases. \\

\newpage