\subsection{Approaches with increasing expressiveness}
\label{subsection_comparison_of_approaches}

\subsubsection{Classic approach (vector algebra)}

\begin{itemize}
    \item Use of scalars and vectors, exclusively
    \item Use of the cross-product between vectors to create directions perpendicular to
    both vectors (a concept that can only be defined in three-dimensional space)
    \item Rotations are defined by a rotation axis (a concept that is only defined in
    three-dimensional space)\footnote{The last two concepts cannot be extended to
    higher-dimensional spaces.}
\end{itemize}

\subsubsection{General observations for Geometric Algebra (GA)}

\begin{itemize}
    \item The GA approach is coordinate-free!
    \item W.r.t. a vector, the dot product (or interior product) takes into account the
    parts parallel to the vector, exclusively. Thus, it is not invertible itself.
    \item W.r.t. a vector, the wedge product (or outer product) takes into account the
    parts perpendicular to the vector, exclusively. Thus, it is not invertible itself.
    \item Only the geometric product taking both parts into account is invertible.
\end{itemize}

\subsubsection{Euclidean geometric algebra}

\begin{itemize}
    \item Use of scalars, vectors, bivectors, ..., pseudoscalars (higher grade objects up
    to order $n$ depend on the dimension $n$ of the modelled space).
    \item Use of the outer product (wedge product) between vectors, which defines the
    space spanned by both vectors (can be defined in any dimension, used as join
    operation).
    \item Use of the regressive outer product (regressive wedge product) which defines the
    space shared by both objects (represents the intersection of the spaces used by the
    objects, used as meet operation).
    \item Modelled objects cover scalars, directions, planes (in $3D$) and the whole space
    directly.
    \item All objects modelled in the algebra contain the (non-distinguishable) origin
    $O=(0,0,...,0)$. Objects outside the origin must be modeled by explicitly adding
    translations (leading to objects that cannot directly be manipulated as easily as
    objects that directly represent subspaces within the algebra).
    \item Orthogonal transformations in the algebra can directly model reflections and
    rotations. Rotations are defined in a plane spanned by two vectors defining the
    rotation.
\end{itemize}

\subsubsection{Projective (Euclidean) geometric algebra}

\begin{itemize}
    \item Like $EGA$, but in addition having a distinguishable origin encoded in an
    additional dimension $O=(0,0,...,0,1)$ (in a projective space with an additional
    dimension projected onto a fixed value, typically 1).
    \item Use of the outer and regressive outer products as join and meet operations.
    \item Directly models points, directions and planes (in $3D$), as well as the whole
    space, including objects located in the horizon, i.e. at infinity (points at
    infinity are modeled as directions towards those points).
    \item Can directly model higher grade objects by products between objects in the
    algebra that do not contain the origin, e.g. the wedge product between two arbitrary
    points defines a line.
    \item Orthogonal transformations in the algebra can directly model reflections,
    rotations and translations.
\end{itemize}


\subsubsection{Comparison of approaches in different algebras}

% Three-column layout: Application (spanning) | Tag | Description
% Adjust \cwApp and \cwTag; description fills the rest.
\newlength{\cwApp}\setlength{\cwApp}{0.185\textwidth}
\newlength{\cwTag}\setlength{\cwTag}{0.06\textwidth}
\newlength{\cwDesc}\setlength{\cwDesc}{\dimexpr\textwidth - \cwApp - \cwTag
                                       - 4\arrayrulewidth - 6\tabcolsep\relax}

\begin{table}[H]
{\setlength{\extrarowheight}{2pt}\renewcommand{\arraystretch}{1.15}%
% \extrarowheight: absolute extra space added above each row (before top rule);
% \arraystretch:   scales the row strut height (1.15 ≈ +15%, adds ~1.5pt below content).
% Together: ~2pt top padding, ~1.5pt bottom padding per row.
\begin{tabular}{|p{\cwApp}|p{\cwTag}|p{\cwDesc}|}
    \hline
    \textbf{Application} & & \textbf{Description} \\
    \hline \hline
    \multirow{3}{\cwApp}{Create line $\bm{l}$ from points \newline
    $P$, $Q$} & VA  & $P=(p_x,p_y,p_z)$ \newline
    $\bm{l}\colon \vec{x}(t) = (Q-P) \cdot t + P$ \\
    \cline{2-3}
    & EGA & $P=(p_x,p_y,p_z)$, a vector represents a $1D$ subspace\newline
    $\bm{l}\colon x(t)=P \cdot t$ (explicit, line through origin and $P$)\newline
    $\bm{l}\colon x(t)=(Q-P) \cdot t + P$\newline
    (explicit, line in direction $Q-P$ through $P$)\newline
    $\bm{l}\colon(x-P) \wedge (Q-P) = 0$ \newline
    (implicit, line in direction $Q-P$ through $P$)\\
    \cline{2-3}
    & PGA & $P=(p_x,p_y,p_z,1)$ \newline
    $\bm{l} = P \wedge Q$ (line through points $P$ and $Q$, bivector)\\
    \hline \hline
    \multirow{3}{\cwApp}{Create plane $\bm{T}$ from points $P$, $Q$, $R$}
    & VA  & $\vec{u} = Q - P$, $\vec{v} = R - P$ \newline
            $\bm{T}\colon x(s,t) = \vec{u} \cdot s + \vec{v} \cdot t + P$ \\
    \cline{2-3}
    & EGA & $u = Q - P$, $v = R - P$, $B = u \wedge v$ \newline
    $\bm{T} = P \wedge Q$ (plane through $O$, $P$, $Q$; bivector) \newline
    $\bm{T}\colon (x-P) \wedge B = 0$ (plane through $P$, $Q$, $R$; bivector)\\
    \cline{2-3}
    & PGA & $\bm{T} = P \wedge Q \wedge R$ (plane through $P$, $Q$, $R$, trivector) \\
    \hline \hline
    \multirow{3}{\cwApp}{Intersect two lines $\bm{l_1}$, $\bm{l_2}$}
    & VA  & $\bm{l_1}\colon \vec{x}(s) = \vec{u} \cdot s + P$,
    $\bm{l_2}\colon \vec{y}(t) = \vec{v} \cdot t + Q$, \newline
    solve system in $2D$ for $s_i$ and $t_i$: $\vec{u} \cdot s + P = \vec{v} \cdot t + Q$
    \newline
    intersection point $I = \vec{x}(s_i) = \vec{y}(t_i)$\\
    \cline{2-3}
    & EGA & see VA\\
    \cline{2-3}
    & PGA & Intersect arbitrary lines $l_1$, $l_2$ in a common plane\newline
    intersection point $I$: $I = l_1 \vee l_2$ (projective point)\\
    \hline \hline
    \multirow{3}{\cwApp}{Intersect two planes $\bm{T_1}$, $\bm{T_2}$}
    & VA  & $x(s,t) = \vec{u} \cdot s + \vec{v} \cdot t + P$,
    $y(m,n) = \vec{m} \cdot p + \vec{n} \cdot q + Q$\newline
    solve system in $3D$: $\vec{u} \cdot s + \vec{v} \cdot t + P =
    \vec{m} \cdot p + \vec{n} \cdot q + Q$\\
    \cline{2-3}
    & EGA & see VA\\
    \cline{2-3}
    & PGA & Intersect arbitrary planes $T_1$, $T_2$\newline
    intersection line $I$: $I = T_1 \vee T_2$ (projective line, bivector)\\
    % \hline \hline
    % \multirow{3}{\cwApp}{...}
    % & VA  & ... \\
    % \cline{2-3}
    % & EGA & ... \\
    % \cline{2-3}
    % & PGA & ... \\
    \hline
\end{tabular}}% end \extrarowheight + \arraystretch group
\caption{Comparison of common geometric construction tasks across three approaches.
         \textbf{VA} = Vector Algebra (classical approach using vectors and parametric
         equations); \textbf{EGA} = Euclidean Geometric Algebra; \textbf{PGA} =
         Projective (Euclidean) Geometric Algebra.}
\end{table}

$PGA$ does not require solving systems of equations for simple geometrical tasks. They
can be solved rather easily by using the suitable product.

\newpage