\subsection{Glossary}
\label{subsection_glossary}

\textbf{Moment and torque} (source Wikipedia):
\begin{itemize}
    \item A moment is a mathematical expression involving the product of a distance and a
    physical quantity such as a force or electric charge. Moments are usually defined with
    respect to a fixed reference point and refer to physical quantities located some
    distance from the reference point. The moment of force, often called torque, is the
    product of a force on an object and the distance from the reference point to the
    object.

    \item In physics and mechanics, torque is the rotational correspondent of linear
    force. It is referred to as the moment of force, or simply the moment. Just as a
    linear force is a push or a pull applied to a body, a torque can be thought of as a
    twist applied to an object with respect to a chosen point.

    \item Torque is referred to using different vocabulary depending on geographical
    location and field of study. Torque and moment are often used interchangeably for the
    rotational effect of a force, but in mechanics, the main difference is application:
    torque refers to a force causing rotation in a dynamic system, while moment is often
    used for the turning effect in a static, or non-moving, system. Both are calculated as
    the force multiplied by the perpendicular distance from the pivot point.

    \item When modelling physics in $PGA$ forces and moments, describing the translational
    and rotational physical aspects, do not need to be considered separately, but can be
    modelled within one object containing all information. For example, by modelling the
    line of action $F$ of a force $f$ acting at a point $P$ as $F = P \wedge f$ as in
    equation~(\ref{eq:line_of_action}) both effects are covered, the force $f$ itself and
    the torque generated at any point are covered by the mathematical object $F$. The
    torque generated at the origin $M_{O} = bulk(F)$ can be as easily recovered from $F$
    as the resulting free moment at any other point.
\end{itemize}

\textbf{Momentum}
\begin{itemize}
    \item Momentum (German: Impuls, Bewegungsgröße) characterizes the movement of an
    object. It has a direction (the direction of the speed vector) and a magnitude (the
    product of the mass and the magnitude of the speed vector). A heavier object moving at
    the same speed has more momentum compared to an object of less mass. Momentum can
    change in direction and/or magnitude.
\end{itemize}

\newpage